### 2023-10-07 R/community-clustering.qmd
initial commit 🚀

---
title: "Supplement A: Community Clustering"
date: today
published-title: "COMPILED"
format:
  html: 
    embed-resources: true
    fig-align: center
    fig-dpi: 300
    fig-responsive: true
    number-sections: true
    number-depth: 3
    toc: true
    toc-depth: 3
    smooth-scroll: true
    theme: cosmo
    highlight-style: a11y
execute: 
  echo: true
  warning: false
  error: false
---

## Introudction

**Goal:** identify geographic communities

**Data:** VEPII and Cibola databases

**Method:** using a somewhat novel clustering algorithm outlined below

This document is setup so that it focuses on the actual steps of the analysis, using verbose names for functions that should make their purpose clear. You can also dig into what each function does by unfolding the relevant code chunk. The critical R package for doing path analysis is [`cppRouting`](https://github.com/vlarmet/cppRouting). It's optimized for road networks, but still very fast. Someone should write a path-finding package (maybe using Rust?) optimized for grid networks...

## R Preamble

```{r}

library(arrow)
library(cppRouting)
library(ggfx)
library(ggnewscale)
library(ggspatial)
library(gt)
library(here)
library(igraph)
library(magick)
library(patchwork)
library(sf)
library(terra)
library(tidyterra)
library(tidyverse)

```

Some ggplot defaults

```{r}
#| code-fold: true

region_colors <- c(
  "cib" = "#D57300", 
  "cmv" = "#009E8C", 
  "nrg" = "#9B70FF"
)

region_labels <- as_labeller(c(
  "cmv" = "Central Mesa Verde",
  "nrg" = "Northern Rio Grande",
  "cib" = "Cibola"
))

theme_set(theme_bw())

theme_update(
  axis.text = element_text(size = rel(0.75), color = "gray50"),
  axis.ticks = element_line(linewidth = 0.2, color = "gray85"),
  axis.title = element_text(size = rel(1)),
  legend.text = element_text(size = rel(0.9)),
  legend.title = element_text(size = rel(0.9)),
  panel.grid = element_blank(),
  plot.title = element_text(size = rel(1), margin = margin(b = 2)),
  strip.background = element_blank(),
  strip.text = element_text(size = rel(1))
)

# trim all white space around image, but then add a few white pixels back (dx, dy)
prepare_image <- function(x, dx = 2, dy = 30, color = "white") {

  img <- magick::image_read(path = x)

  img <- magick::image_trim(img)

  info <- magick::image_info(img)

  new_width <- info$width + dx
  new_height <- info$height + dy

  img <- magick::image_extent(
    img, 
    geometry_area(new_width, new_height), 
    color = color
  )

  magick::image_write(img, path = x)

}

```

## Data

```{r}
#| code-fold: true

build_region <- function(gpkg, region){

  region_query <- paste0("SELECT * FROM regions WHERE region = '", region, "'")
  r <- read_sf(gpkg, query = region_query)

  site_query <- paste0("SELECT * FROM sites WHERE region = '", region, "'")
  s <- read_sf(gpkg, query = site_query) |> st_filter(r)

  if ("communities" %in% st_layers(gpkg)$name){

    community_query <- paste0("SELECT * FROM communities WHERE region = '", region, "'")
    t <- read_sf(gpkg, query = community_query) |> st_filter(r)

  } else {

    t <- NULL

  }

  list(
    name = region,
    preferred_projection = ifelse(region == "nrg", 26913, 26912),
    region = r,
    sites = s,
    communities = t
  )

}

add_elevation <- function(x){

  dem_fn <- paste0("dem-", x$name, ".tif")
  x$dem <- here("data", dem_fn) |> rast()

  return(x)

}

# getters
get_farms <- function(x){

  x$sites |> 
    st_drop_geometry() |> 
    filter(center == 0)

}

get_centers <- function(x){

  x$sites |> 
    st_drop_geometry() |> 
    filter(center == 1)

}

```

```{r}

gpkg <- here("data", "community-centers.gpkg")

cmv <- build_region(gpkg, "cmv") |> add_elevation()
nrg <- build_region(gpkg, "nrg") |> add_elevation()
cib <- build_region(gpkg, "cib") |> add_elevation()

```

```{r}
#| code-fold: true

e <- function(x){

  tibble(
    region = x$name,
    min =  terra::global(x$dem, min, na.rm = TRUE)$min,
    max =  terra::global(x$dem, max, na.rm = TRUE)$max,
    mean = terra::global(x$dem, mean, na.rm = TRUE)$mean,
    sd =   terra::global(x$dem, sd, na.rm = TRUE)$sd
  )

}

bind_rows(
  e(cmv),
  e(nrg),
  e(cib)
) |> 
  gt() |> 
  tab_header("Elevation") |> 
  fmt_number(-region, decimals = 1) |> 
  opt_align_table_header("left")

```

```{r}
#| code-fold: true

plot_region <- function(x){

  hell <- function(x, n = 1000){

    slope <- terrain(x, "slope", unit = "radians")
    aspect <- terrain(x, "aspect", unit = "radians")

    hill <- shade(slope, aspect, 30, 45)
    hill <- setValues(hill, scales::rescale(values(hill), to = c(1, n)))
    hill <- round(hill)

    list(
      shade = hill,
      grays = hcl.colors(n, "Grays")[values(hill)]
    )

  }

  h <- hell(x$dem)

  ggplot() +
    geom_spatraster(
      data = h$shade, 
      fill = h$grays,
      maxcell = Inf
    ) +
    geom_spatraster(data = x$dem, alpha = 0.5) +
    scale_fill_hypso_c(
      name = "Elevation (m)",
      palette = "utah_1",
      limits = c(1350, 3850),
      alpha = 0.5,
      na.value = "transparent"
    ) +
    with_outer_glow(
      geom_sf(
        data = x$sites |> filter(center == 0),
        shape = 16,
        color = "gray25",
        alpha = 0.5,
        size = 0.6
      ),
      colour = "white",
      sigma = 0.5,
      expand = 1
    ) +
    with_outer_glow(
      geom_sf(
        data = x$sites |> filter(center == 1),
        shape = 17,
        color = "#CD3C00",
        alpha = 0.8,
        size = 1.1
      ),
      colour = "white",
      sigma = 0.5,
      expand = 1
    ) +
    annotation_scale(
      aes(location = "bl", line_col = "white", text_col = "white"),
      height = unit(0.18, "cm"),
      line_width = 0.5
    ) +
    coord_sf(expand = FALSE) +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      legend.position = "none",
      plot.margin = margin(l = 2)
    )

}

gg <- (plot_region(cmv) + ggtitle("Central Mesa Verde")) +
  (plot_region(nrg) + ggtitle("Northern Rio Grande")) + 
  (plot_region(cib) + ggtitle("Cibola"))

fn <- here("figures", "overview-maps.png")

ggsave(
  plot = gg,
  filename = fn,
  width = 7.5, 
  height = 4.5,
  dpi = 600,
  bg = "white"
)

prepare_image(fn)

remove(plot_region, gg, fn)

```

::: {#fig-overview}
![](../figures/overview-maps.png){width="100%" fig-alt="An overview map showing the distribution of farms and community centers across the three study areas."}

An overview map showing the distribution of farms and community centers across the three study areas.
:::

## Graph

It's important to note that throughout we track movement between grid cells, not site points.

### Convert grid to graph

```{r}
#| code-fold: true

add_graph <- function(x, max_slope = NULL, ...){

  g <- terra::adjacent(
    x$dem,
    cells = terra::cells(x$dem),
    directions = 8,
    pairs = TRUE
  )

  g <- as.data.frame(g)

  g <- g[!g$to %in% terra::cells(x$dem, NA_real_)[[1]], ]

  # distance between cells
  g$run <- terra::distance(
    terra::xyFromCell(x$dem, as.vector(g$from)),
    terra::xyFromCell(x$dem, as.vector(g$to)),
    lonlat = terra::is.lonlat(x$dem),
    pairwise = TRUE
  )

  # difference between cells
  g$from_elevation <- terra::extract(x$dem, as.vector(g$from))[,1]

  g$to_elevation <- terra::extract(x$dem, as.vector(g$to))[,1]

  g$rise <- g$from_elevation - g$to_elevation

  # slope
  g$slope <- g$rise/g$run

  # what is a realistic slope people would try to hike?
  if (!is.null(max_slope)) {

    g <- g[g$slope <= tan(max_slope * pi / 180), ]

  }

  g$speed <- campbell(g$slope, ...)

  g$cost <- g$run/g$speed

  g <- g[, c("from", "to", "cost")]

  coords <- cbind(
    ID = terra::cells(x$dem), 
    terra::xyFromCell(x$dem, terra::cells(x$dem))
  ) |> 
    as.data.frame() |> 
    rename_with(toupper)

  # force as.character(1000000) to be "1000000" instead of "1e-6"
  # this fixes problem with makegraph()
  options(scipen = 9999)

  x$graph <- makegraph(
    as.data.frame(g),
    directed = TRUE,
    coords = coords
  )

  return(x)

}

campbell <- function(x, decile = 30L) {

  if (length(decile) != 1) {

    stop("campbell accepts only one decile at a time.")

  }

  # campbell function wants degrees, not rise-over-run as in tobler
  # but, we're working with rr, so remember to convert radians to degrees with 180/pi!
  slope <- atan(x) * 180 / pi

  # values provided in Campbell 2019, Supplement, Table S1
  # for simplicity, I excluded all but the deciles
  deciles <-
    data.frame(
      decile = c(10L, 20L, 30L, 40L, 50L, 60L, 70L, 80L, 90L),
      a = c(-1.568, -1.71,  -1.858, -1.958, -2.171, -2.459, -2.823,  -3.371,  -3.06),
      b = c(13.328, 10.154,  8.412,  8.96,  10.064, 11.311, 12.784,  15.395,  16.653),
      c = c(38.892, 36.905, 39.994, 50.34,  63.66,  79.287, 98.697, 134.409, 138.875),
      d = c( 0.404,  0.557,  0.645,  0.649,  0.628,  0.599,  0.566,   0.443,   0.823),
      e = c(-0.003, -0.004, -0.004, -0.005, -0.005, -0.005, -0.005,  -0.005,  -0.0139)
    )

  ind <- which(deciles$decile == as.integer(decile))

  deciles <- deciles[ind, ]

  a <- deciles$a
  b <- deciles$b
  c <- deciles$c
  d <- deciles$d
  e <- deciles$e

  # lorentz distribution
  lorentz <- (1 / ((pi * b) * (1 + ((slope - a) / b)^2)))

  # modified lorentz
  (c * lorentz) + d + (e * slope)

}

```

```{r}

cmv <- cmv |> add_graph(max_slope = 45, decile = 10L)
nrg <- nrg |> add_graph(max_slope = 45, decile = 10L)
cib <- cib |> add_graph(max_slope = 45, decile = 10L)

```

### Add distance matrix

```{r}
#| code-fold: true

add_distance_matrix <- function(x){

  x$sites$cell <- terra::cells(x$dem, terra::vect(x$sites))[,2]

  # use unique cell IDs as some centers are in the same pixel, 
  # so calculations would be redundant
  dm <- get_distance_matrix(
    x$graph,
    from = x$sites$cell |> unique(),  
    to = x$sites$cell |> unique()
  )

  # make symmetric by averaging off-diagonals
  d <- cbind(dm[lower.tri(dm)], t(dm)[lower.tri(dm)])

  dm[lower.tri(dm)] <- rowMeans(d, na.rm = TRUE)
  dm[upper.tri(dm)] <- t(dm)[upper.tri(dm)]

  diag(dm) <- 0

  x$dm <- dm

  return(x)

}

```

```{r}

cmv <- cmv |> add_distance_matrix()
nrg <- nrg |> add_distance_matrix()
cib <- cib |> add_distance_matrix()

```

```{r}
#| code-fold: true

get_commutes <- function(x, n){

  centers <- get_centers(x) |> select(-center)
  farms   <- get_farms(x) |> select(-center)

  # subset dm (distance matrix) to get dm[farms, centers]
  i <- which(rownames(x$dm) %in% farms$cell)
  j <- which(colnames(x$dm) %in% centers$cell)

  dm <- x$dm[i, j]

  observed <- apply(dm, 1, min)
  observed <- density(observed, kernel = "gaussian", n = 512, from = 0, to = 7200)

  r <- terra::rast(
    nrows = nrow(x$dem), 
    ncols = ncol(x$dem), 
    extent = ext(x$dem), 
    crs = crs(x$dem)
  )

  dmax <- st_distance(
    x$sites |> filter(center == 0), 
    x$sites |> filter(center == 1)
  ) |> units::drop_units() |> apply(1, min) |> max()

  i <- cells(
    x$dem, 
    x$sites |> filter(center == 1) |> st_buffer(dmax) |> st_union() |> vect()
  )[, "cell"]

  i <- i[i %in% x$graph$dict$ref]

  r[i] <- 1

  # don't sample cells with centers in them
  j <- colnames(dm) |> as.numeric()

  r[j] <- NA

  random_points <- terra::spatSample(
    r, 
    size = n,
    method = "regular",
    cells = TRUE,
    na.rm = TRUE
  ) |> pull(cell) |> unique()

  dm <- get_distance_matrix(
    x$graph,
    from = random_points,  
    to = colnames(dm)
  )

  # make symmetric by averaging off-diagonals
  d <- cbind(dm[lower.tri(dm)], t(dm)[lower.tri(dm)])

  dm[lower.tri(dm)] <- rowMeans(d, na.rm = TRUE)
  dm[upper.tri(dm)] <- t(dm)[upper.tri(dm)]

  diag(dm) <- 0

  expected <- apply(dm, 1, min)
  expected <- density(expected, kernel = "gaussian", n = 512, from = 0, to = 7200)

  tibble(
    region = x$name,
    distance = observed$x,
    observed = observed$y,
    expected = expected$y
  )

}

set.seed(1701) # USS Enterprise

commutes <- bind_rows(
  cmv |> get_commutes(n = 1000),
  nrg |> get_commutes(n = 1000),
  cib |> get_commutes(n = 1000)
)

q <- function(x, p){

  centers <- get_centers(x) |> select(-center)
  farms   <- get_farms(x) |> select(-center)

  # subset dm (distance matrix) to get dm[farms, centers]
  i <- which(rownames(x$dm) %in% farms$cell)
  j <- which(colnames(x$dm) %in% centers$cell)

  dm <- x$dm[i, j]

  observed <- apply(dm, 1, min)

  tibble(
    region = x$name,
    distance = quantile(observed, p = p) |> unname()
  )

}

q95 <- bind_rows(
  q(cmv, 0.95),
  q(nrg, 0.95),
  q(cib, 0.95)
) |> 
  mutate(
    y = c(2.4, 1.8, 1.2),
    label = case_when(
      region == "cib" ~ "Cibola",
      region == "cmv" ~ "Central Mesa Verde",
      region == "nrg" ~ "Northern Rio Grande",
      TRUE ~ NA
    )
  )

gg <- ggplot() +
  geom_hline(
    yintercept = 0,
    color = "gray75",
    linewidth = 0.2
  ) +
  geom_line(
    data = commutes, 
    aes(distance/60, log(observed/expected), color = region), 
    linewidth = 0.9,
    lineend = "round"
  ) +
  geom_point(
    data = q95,
    aes(distance/60, y, color = region),
    size = 2
  ) +
  geom_text(
    data = q95,
    aes(distance/60, y, color = region, label = label),
    vjust = 0.5,
    hjust = 0,
    nudge_x = 2,
    size = 12/.pt
  ) +
  scale_color_manual(name = NULL, values = region_colors) +
  annotate(
    "point",
    x = min(q95$distance)/60,
    y = 3.0,
    color = "gray25",
    size = 2
  ) +
  annotate(
    "text",
    label = "95% Quantile of Observed",
    x = min(q95$distance)/60 + 2,
    y = 3.0,
    color = "gray25",
    vjust = 0.5,
    hjust = 0,
    size = 12/.pt
  ) +
  labs(
    x = "Commute to Nearest Center [mins]",
    y = "Density\nlog(Observed) - log(Random)"
  ) +
  scale_x_continuous(breaks = seq(0, 120, by = 30), expand = expansion(0.03)) +
  coord_cartesian(xlim = c(0, 120)) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_line(linewidth = 0.2, color = "gray85")
  )

fn <- here("figures", "commute-kl.png")

ggsave(
  plot = gg,
  filename = fn,
  width = 5,
  height = 3.3,
  dpi = 600
)

prepare_image(fn)

remove(get_commutes, q, q95, gg)

```

::: {#fig-commutes}
![](../figures/commute-kl.png){width="60%" fig-alt="Difference between log probability density of obversed commute times from farms to nearest centers and log probability density of commute times from random locations to nearest centers."}

Difference between log probability density of obversed commute times from farms to nearest centers and log probability density of commute times from random locations to nearest centers.
:::

## Communities

Our community detection algorithm proceeds as follows:

1. Identify community centers   
2. Identify farms within commute distance 
3. Join communities to their overlapping neighbors 
4. Exclude farms closer to another community center (tie-breaking step) 
5. Draw smallest concave hull encompassing all farms, centers, and paths

Step 1 is already done and stored in the data.

### Filter by commute distance

```{r}
#| code-fold: true

subset_distance_matrix <- function(x){

  centers <- get_centers(x) |> select(-center)
  farms   <- get_farms(x) |> select(-center)

  # subset dm (distance matrix) to get dm[farms, centers]
  i <- which(rownames(x$dm) %in% farms$cell)
  j <- which(colnames(x$dm) %in% centers$cell)

  x$dm[i, j]

}

filter_distance_matrix <- function(x, d){

  nearest_farms <- apply(x, 1, \(x){ which(x <= d) })
  nearest_farms <- lapply(nearest_farms, \(x){ as.numeric(names(x)) })

  arrow_table(
    center = unlist(nearest_farms),
    farm = rep(names(nearest_farms), lengths(nearest_farms)) |> as.numeric()
  ) |> arrange(center, farm)

}

find_nearby_farms <- function(x, dmax){

  x$dmax <- dmax

  x$nearest_farms <- x |> 
    subset_distance_matrix() |> 
    filter_distance_matrix(dmax)

  sites <- x$sites |> 
    st_drop_geometry() |> 
    select(cell, n_room) |> 
    group_by(cell) |> 
    summarize(n_room = sum(n_room))

  x$nearest_farms <- x$nearest_farms |> left_join(sites, by = join_by(farm == cell))

  return(x)

}

```

```{r}

# how many seconds in ... ?
one_hour <- 3600

cmv <- cmv |> find_nearby_farms(dmax = one_hour)
nrg <- nrg |> find_nearby_farms(dmax = one_hour)
cib <- cib |> find_nearby_farms(dmax = one_hour)

```

### Find neighboring centers

> For any two centers $C_1$ and $C_2$ with nearby farm room counts $N_1 < N_2$, if $pN_1$ of $C_1$’s rooms are within a distance $d$ of $C_1$ and $C_2$, then $C_1$ is part of the same community as $C_2$.

```{r}
#| code-fold: true

join_neighboring_centers <- function(x, p, djoin){

  x$djoin <- djoin

  nearest_farms <- x |> 
    subset_distance_matrix() |> 
    filter_distance_matrix(djoin)

  sites <- x$sites |> 
    st_drop_geometry() |> 
    select(cell, n_room) |> 
    group_by(cell) |> 
    summarize(n_room = sum(n_room))

  nearest_farms <- nearest_farms |> left_join(sites, by = join_by(farm == cell))

  remove(sites)

  room_counts <- nearest_farms |> 
    group_by(center) |> 
    summarize(n_room = sum(n_room)) |> 
    collect()

  R <- outer(room_counts$n_room, room_counts$n_room, FUN = '<=')
  diag(R) <- FALSE

  colnames(R) <- room_counts$center
  rownames(R) <- room_counts$center

  R <- apply(R, 1, which)
  R <- lapply(R, \(x){ as.numeric(names(x)) })

  neighbors_matrix <- imap(R, \(x, idx){

    N <- room_counts |> filter(center == as.numeric(idx)) |> pull(n_room)

    focal_farms <- nearest_farms |> 
      filter(center == as.numeric(idx)) |> 
      collect() |> 
      pull(farm)

    cntrs <- nearest_farms |> 
      filter( # this filter gives the intersection of farms/cells 
        center %in% x,
        farm %in% focal_farms
      ) |> 
      group_by(center) |> 
      summarize(proportion = sum(n_room)/N) |> 
      filter(proportion >= p) |> 
      collect() |> 
      pull(center)

    names(R) %in% cntrs

  })

  neighbors_matrix <- do.call("rbind", neighbors_matrix)
  diag(neighbors_matrix) <- TRUE

  colnames(neighbors_matrix) <- names(R)
  rownames(neighbors_matrix) <- names(R)

  # igraph...
  g <- igraph::graph.adjacency(neighbors_matrix)
  membership <- igraph::components(g)$membership

  x$lookup <- tibble(
    center = names(membership) |> as.numeric(),
    community = unname(membership)
  )

  return(x)

}

```

```{r}

# how many seconds in ... ?
half_hour <- 1800

cmv <- cmv |> join_neighboring_centers(p = 0.8, djoin = half_hour)
nrg <- nrg |> join_neighboring_centers(p = 0.8, djoin = half_hour)
cib <- cib |> join_neighboring_centers(p = 0.8, djoin = half_hour)

```

### Find nearest center

```{r}
#| code-fold: true

find_nearest_center <- function(x){

  farms_to_centers <- subset_distance_matrix(x)

  # find the nearest center for each farm
  nearest_center <- apply(farms_to_centers, 1, \(x) which.min(x))
  nearest_center <- colnames(farms_to_centers)[nearest_center]

  distance <- apply(farms_to_centers, 1, min)

  x$nearest_center <- tibble(
    farm = farms_to_centers |> rownames() |> as.numeric(),
    center = nearest_center |> as.numeric(),
    distance = unname(distance)
  ) |> filter(distance <= x$dmax)

  return(x)

}

```

```{r}

cmv <- cmv |> find_nearest_center() 
nrg <- nrg |> find_nearest_center() 
cib <- cib |> find_nearest_center()

```

### Define boundaries

Now that we have each farm associated with its nearest community center, and each center joined with its neighbors, assuming it has any, we can define a boundary for the community using a concave hull. We choose a concave rather than convex hull because the latter tends to artificially inflate the size of the community, often to a significant degree. On the other hand, the concave hull can generate extremely unrealistic shapes for communities. To avoid these issues, we first compute the least-cost path between each farm in the community and every other farm in the community. We then add the vertices that define these paths as virtual points to the community. We then generate the concave hull for the expanded set of community points. The consequence is that an individual setting off from one farm in the community to another should never have to leave the community. Their walk will always be within the community boundary.

The result is still not perfect, however, so we further simplify polygons (including filling holes) by adding then removing a small buffer. Two things to note here: (1) we project the boundaries first as Google's `S2` system for geocomputations on the sphere (the default now in `{sf}`) is janky with how it does buffering and (2) we add a small buffer to smooth some of the noise in the hull.

```{r}
#| code-fold: true

collect_members <- function(x){

  lookup <- x$lookup

  centers <- get_centers(x) |> 
    left_join(lookup, by = join_by(cell == center)) |> 
    select(site_id, n_room, initial_phase, final_phase, community) |> 
    filter(!is.na(community)) |> 
    mutate(type = "center")

  farms <- get_farms(x) |> 
    select(-center) |> 
    left_join(x$nearest_center, by = join_by(cell == farm)) |> 
    left_join(lookup, by = "center") |> 
    select(site_id, n_room, initial_phase, final_phase, community) |> 
    filter(!is.na(community)) |> 
    mutate(type = "farm")

  x$members <- bind_rows(farms, centers) |> 
    arrange(community, site_id) |> 
    left_join(x$sites[, "site_id"], by = "site_id") |> 
    mutate(region = x$name, .before = everything()) |> 
    st_sf()

  return(x)

}

define_boundaries <- function(x, concavity_ratio, dilate){

  community_ids <- x$members |> pull(community) |> unique()

  communities <- vector("list", length(community_ids))

  # cli::cli_progress_bar(
  #   "Define boundaries", 
  #   total = length(community_ids)
  # )

  for (i in community_ids){

    m <- x$members |> filter(community == i)

    cells <- terra::cells(x$dem, vect(m))[, 2] |> unique()

    paths <- get_multi_paths(
      x$graph, 
      from = cells, 
      to = cells
    )

    paths <- paths |> unlist(recursive = TRUE) |> as.numeric() |> unique()

    xy <- rbind(
      terra::xyFromCell(x$dem, paths),
      st_coordinates(m)[, 1:2]
    )

    xy <- xy |> st_multipoint() |> st_sfc(crs = st_crs(m)) |> st_cast("POINT")

    m <- st_geometry(m)

    border <- st_sf(geometry = c(xy, m)) |> 
      summarize() |> 
      st_concave_hull(ratio = concavity_ratio) |> 
      mutate(community = i)

    communities[[i]] <- border

    remove(i, m, cells, paths, xy, border)

    # cli::cli_progress_update()

  }

  x$communities <- communities |> 
    bind_rows() |> 
    st_transform(x$preferred_projection) |> 
    st_buffer(dilate[1]) |> 
    st_buffer(dilate[2]) |> 
    st_transform(4326) |> 
    mutate(region = x$name, .before = "community")

  return(x)

}

```

```{r}

cmv <- cmv |> 
  collect_members() |> 
  define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))

nrg <- nrg |> 
  collect_members() |> 
  define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))

cib <- cib |> 
  collect_members() |> 
  define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))

```

### Drop small communities

```{r}
#| code-fold: true

drop_small_communities <- function(x, threshold){

  final_list <- x$members |> 
    st_drop_geometry() |> 
    filter(type == "farm") |> 
    group_by(community) |> 
    summarize(n = n()) |> 
    filter(n > threshold) |> 
    pull(community)

  x$communities <- x$communities |> filter(community %in% final_list)

  x$members <- x$members |> st_filter(x$communities)

  return(x)

}

```

```{r}

cmv <- cmv |> drop_small_communities(threshold = 3)
nrg <- nrg |> drop_small_communities(threshold = 3)
cib <- cib |> drop_small_communities(threshold = 3)

```

## Results

Here we show the community polygons themselves as well as the log-density of rooms in each.

```{r}
#| code-fold: true

visualize_communities <- function(x, lbl = region_labels){

  hell <- function(x, n = 1000){

    slope <- terrain(x, "slope", unit = "radians")
    aspect <- terrain(x, "aspect", unit = "radians")

    hill <- shade(slope, aspect, 30, 45)
    hill <- setValues(hill, scales::rescale(values(hill), to = c(1, n)))
    hill <- round(hill)

    list(
      shade = hill,
      grays = hcl.colors(n, "Grays")[values(hill)]
    )

  }

  h <- hell(x$dem)

  members <- x$members |> 
    st_drop_geometry() |> 
    group_by(community) |> 
    summarize(n_room = sum(n_room))

  communities <- x$communities |> 
    left_join(members, by = "community") |> 
    mutate(
      area = st_area(geometry) |> units::set_units(km2) |> units::drop_units(),
      density = n_room/area,
      density = log(density+1),
      density = cut(density, 9, labels = FALSE)
    )

  ggplot() +
    ggtitle(lbl(x$name)[[1]]) +
    geom_spatraster(
      data = h$shade, 
      fill = h$grays,
      maxcell = Inf
    ) +
    geom_spatraster(data = x$dem, alpha = 0.5,) +
    scale_fill_distiller(palette = "Greys", guide = "none") +
    ggnewscale::new_scale_fill() +
    geom_sf(
      data = communities,
      fill = "transparent",
      color = "white",
      alpha = 0.1,
      linewidth = 0.5
    ) +
    geom_sf(
      data = communities,
      aes(fill = density, color = density),
      linewidth = 0.1
    ) + 
    scale_color_gradientn(
      name = "Log Density", 
      colors = RColorBrewer::brewer.pal(9, "YlOrBr"),
      breaks = c(1, 5, 9),
      labels = c("Low", "Medium", "High"),
      guide = guide_legend()
    ) +
    scale_fill_gradientn(
      name = "Log Density", 
      colors = RColorBrewer::brewer.pal(9, "YlOrBr") |> alpha(0.6),
      breaks = c(1, 5, 9),
      labels = c("Low", "Medium", "High"),
      guide = guide_legend()
    ) +
    annotation_scale(
      aes(location = "bl", line_col = "white", text_col = "white"),
      height = unit(0.18, "cm"),
      line_width = 0.5
    ) +
    coord_sf(expand = FALSE) +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.title = element_blank(),
      legend.key = element_rect(color = "gray30", fill = "transparent"),
      plot.margin = margin(l = 2)
    )

}

zg <- list(cmv, nrg, cib) |> 
  map(visualize_communities) |> 
  patchwork::wrap_plots(ncol = 3)

# patchwork is great, but not perfect...
bob <- theme(
  legend.box.margin = margin(t = -10),
  legend.key.size = unit(0.5, "cm"),
  legend.margin = margin(),
  legend.justification = "left",
  legend.position = "bottom"
)

# had to fiddle with this to get the sizes consistent with the overview maps
fn <- here("figures", "communities-all.png")

ggsave(
  plot = zg + plot_layout(guides = "collect") & bob,
  filename = fn,
  width = 7.5, 
  height = 4.5,
  dpi = 600,
  bg = "white"
)

prepare_image(fn)

remove(visualize_communities, zg, bob, fn)

```

::: {#fig-communities}
![](../figures/communities-all.png){width="100%" fig-alt="Map shows defined community boundaries with fill colors representing the log density of rooms in each."}

Map shows defined community boundaries with fill colors representing the log density of rooms in each.
:::

### Community Distributions

```{r}
#| code-fold: true

get_community_data <- function(x){

  # AREA
  A <- x$communities |> 
    st_area() |> 
    units::set_units(km2) |> 
    units::drop_units()

  tbl_area <- tibble(
    region = x$name, 
    community = x$communities$community,
    area = A
  )

  # ROOMS
  tbl_rooms <- x$sites |> 
    st_join(x$communities[, "community"]) |> 
    select(region, community, n_room) |> 
    st_drop_geometry() |> 
    group_by(community) |> 
    summarize(
      region = unique(region),
      n_room = sum(n_room)
    ) |> 
    relocate(region)

  # FARMS
  tbl_sites <- x$members |> 
    st_drop_geometry() |> 
    filter(type == "farm") |> 
    group_by(community) |> 
    summarize(n_farm = n(), region = unique(region))

  # CENTERS
  tbl_centers <- x$members |> 
    st_drop_geometry() |> 
    filter(type == "center") |> 
    group_by(community) |> 
    summarize(n_center = n(), region = unique(region))

  # COMMUTES
  cell_lookup <- x$sites |> 
    st_drop_geometry() |> 
    select(site_id, cell)

  tbl_commutes <- x$members |> 
    st_drop_geometry() |> 
    filter(type == "center") |> 
    left_join(cell_lookup, by = "site_id") |> 
    left_join(x$nearest_center, by = join_by(cell == center), relationship = "many-to-many") |> 
    group_by(community) |> 
    summarize(commute = mean(distance, na.rm = TRUE), region = unique(region))

  # GATHER EVERYTHING INTO ONE TABLE
  tbl_area |> 
    left_join(tbl_rooms, by = c("region", "community")) |> 
    left_join(tbl_sites, by = c("region", "community")) |>
    left_join(tbl_centers, by = c("region", "community")) |> 
    left_join(tbl_commutes, by = c("region", "community")) |> 
    mutate(room_density = n_room/area)

}

```

```{r}

results <- list(cmv, nrg, cib) |> map(get_community_data) |> bind_rows()

```

```{r}
#| code-fold: true

results |> 
  mutate(commute = commute/60) |> 
  group_by(region) |> 
  summarize(across(area:room_density, list("η" = median, "µ" = mean, "σ" = sd))) |> 
  rename_with(str_remove, pattern = "n_|room_") |> 
  pivot_longer(
    !region,
    names_to = c("variable", "statistic"),
    names_sep = "_"
  ) |> 
  pivot_wider(
    names_from = "variable",
    values_from = "value"
  ) |> 
  mutate(across(area:density, \(x) paste0(statistic, ": ", round(x,2)))) |> 
  group_by(region) |> 
  summarize(across(area:density, \(x) paste0(x, collapse = "<br>"))) |> 
  rename(
    " " = region,
    "Area (km2)" = area,
    "Rooms (N)" = room,
    "Farms (N)" = farm,
    "Centers (N)" = center,
    "Commute (mins)" = commute,
    "Room Density (N/km2)" = density
  ) |> 
  slice(c(2, 3, 1)) |> 
  gt() |> 
  tab_header(
    title = "Community summaries",
    subtitle = md("η = median, µ = mean, σ = standard deviation")
  ) |> 
  fmt_markdown(columns = everything()) |> 
  cols_width(
    " " ~ pct(8),
    "Area (km2)" ~ pct(13),
    "Rooms (N)" ~ pct(13),
    "Farms (N)" ~ pct(13),
    "Centers (N)" ~ pct(13),
    "Commute (mins)" ~ pct(18),
    "Room Density (N/km2)" ~ pct(22)
  ) |> 
  opt_align_table_header("left")

```

```{r}
#| eval: false
#| echo: false

# this code is just for exploring communities
# and associating important (re: named) centers with them
plot_labels <- function(x){

  ggplot(x$communities) +
    geom_sf() +
    geom_sf_label(aes(label = community)) +
    theme_void()

}

get_members <- function(x){

  x$members |> 
    st_drop_geometry() |> 
    mutate(region = x$name, .before = everything())

}

get_sites <- function(x){ x$sites |> st_drop_geometry() }

results |> 
  group_by(region) |> 
  slice_max(room_density, n = 5) |> 
  left_join(
    map(list(cmv, nrg, cib), get_members) |> bind_rows(), 
    by = c("region", "community")
  ) |> 
  left_join(
    map(list(cmv, nrg, cib), get_sites) |> bind_rows() |> filter(center == 1), 
    by = c("region", "site_id")
  ) |> 
  select(region, community, site_id, site_name, area) |> 
  filter(region == "cib") |> View()

```

```{r}
#| code-fold: true

lbls <- results |> 
  group_by(region) |> 
  summarize(x = quantile(area/n_center, p = 0.75)) |> 
  mutate(
    variable = "center_area",
    y = c(1.25, 3.25, 2.25),
    label = case_when(
      region == "cib" ~ "Cibola",
      region == "cmv" ~ "Central Mesa Verde",
      region == "nrg" ~ "Northern Rio Grande",
      TRUE ~ region
    )
  )

lblr <- as_labeller(c(
  "commute" = "Mean Commute to Nearest Center [mins]",
  "room_density" = "Room Density [N/km2]",
  "farm_area" = "Mean Area for Farms [km2/N]",
  "center_area" = "Mean Area for Centers [km2/N]",
  "farm_ratio" = "Ratio of Farms [N] to Centers [N]",
  "area" = "Total Area [km2]"
))

gg <- results |> 
  mutate(
    farm_area = area/n_farm,
    center_area = area/n_center,
    commute = commute/60,
    farm_ratio = n_farm/n_center
  ) |> 
  select(-n_room, -n_farm, -n_center) |> 
  pivot_longer(c(-region, -community), names_to = "variable") |> 
  ggplot() + 
  geom_boxplot(aes(
    x = value, 
    y = fct_relevel(region, "cib", "nrg"), 
    fill = region
  )) +
  geom_text(
    data = lbls,
    aes(x, y, label = label, color = region),
    size = 11.5/.pt,
    nudge_x = 1,
    hjust = 0
  ) +
  scale_color_manual(values = region_colors) +
  scale_fill_manual(values = alpha(region_colors, 0.6)) +
  labs(
    x = NULL,
    y = NULL
  ) +
  facet_wrap(
    vars(variable), 
    ncol = 2, 
    nrow = 3,
    scale = "free_x",
    strip.position = "bottom",
    labeller = lblr
  ) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "none",
    panel.grid.major.x = element_line(linewidth = 0.2, color = "gray85"),
    strip.placement = "outside",
    strip.text = element_text(size = rel(1), margin = margin(t = -1, b = 10))
  )

fn <- here("figures", "results-boxplot.png")

ggsave(
  plot = gg,
  filename = fn,
  width = 6.5,
  height = 7.0,
  dpi = 600
)

prepare_image(fn)

remove(lbls, lblr, gg, fn)

```

::: {#fig-distributions}
![](../figures/results-boxplot.png){width="75%" fig-alt="Distributions of important variables across communities and regions."}

Distributions of important variables across communities and regions.
:::

### Save results

```{r}

bind_rows(
  cmv$members |> mutate(region = "cmv", .before = site_id),
  nrg$members |> mutate(region = "nrg", .before = site_id),
  cib$members |> mutate(region = "cib", .before = site_id)
) |> 
  st_drop_geometry() |> 
  write_sf(gpkg, layer = "members")

bind_rows(
  cmv$communities, 
  nrg$communities, 
  cib$communities
) |> 
  left_join(results, by = c("region", "community")) |> 
  write_sf(gpkg, layer = "communities")

```

## Session Info

```{r}
#| code-fold: true

# save the session info as an object
pkg_sesh <- sessioninfo::session_info(pkgs = "attached")

# inject the quarto info
pkg_sesh$platform$quarto <- paste(
  quarto::quarto_version(), 
  "@", 
  quarto::quarto_path()
)

# print it out
pkg_sesh

```

### 2023-10-07 manuscript/manuscript.qmd
initial commit 🚀

---
title: A method for defining dispersed community territories
date: today
author:
  - name: Kenneth B. Vernon
    email: Kenneth.Vernon@colorado.edu
    affiliations: 
      - id: colorado
        address: | 
          Center for Collaborative Synthesis in Archaeology,  
          Institute of Behavioral Science,  
          University of Colorado, Boulder,  
          1440 15th Street,  
          Boulder, CO 80302
    attributes:
        corresponding: true 
  - name: Scott G. Ortman
    affiliations:
      - ref: colorado
abstract: |
  The transition from dispersed to aggregated forms of settlement reflects a critical shift in the relative value of social and primary (food) modes of production. Investigating trade-offs between them, however, requires estimates of the extent of community territories, including their nearby arable land. Here we demonstrate a simple algorithm to do that. Our algorithm is analogous to that used to define core-based statistical areas for the US census, though instead of modern metropolitan areas, we rely on community centers (or areas of known and persistent interaction between unrelated individuals). We provide examples of our algorithm by applying it to archaeological sites in the central Mesa Verde, northern Rio Grande regions, and Cibola regions. Together, these data provide a comprehensive, high resolution record of prehistoric maize farming in the North American Southwest, representing varying levels of aggregation and dispersal through time.
keywords:
  - Aggregation
  - Core-Based Statistical Area
  - Community Detection
  - Intensification
target-journal: Journal of Archaeological Science
bibliography: bibliography.bib
linenumbers: false
watermark: false
format: 
  simple-pdf: default
execute: 
  echo: false
  error: false
  # warning: false
engine: knitr
---

# Introduction

Urban geographers recognize density as a crucial property of human communities which influences many other properties, including productivity, resource needs, and infectious disease rates [@angel_etal2021; @bettencourt2021; @duranton_puga2020]. The archaeological record of subsistence farming societies contains physical traces of community formations with a wide range of densities, often glossed as "dispersed" or "aggregated" [@birch2013; @drennan_etal2015; @gyucha2019]. The primary difference between the two has to do with the spatial distribution of residents relative to primary (food) production. In dispersed formations, households are scattered across an area, interspersed with the land that they farmed, thus requiring them to commute to central places for various forms of social production. In aggregated formations, in contrast, households are clustered in a village or town, such that they reside in locations of social production, but must commute to fields for primary production.

From a complex systems perspective, the main way these two settlement morphologies vary is with respect to which costs are being minimized. In the dispersed pattern, costs associated with primary (food) production by the household are being minimized, whereas in the aggregated pattern, it is costs associated with various forms of social production, from government and ritual to economic exchange and warfare. Presumably, choices regarding which sorts of costs to minimize are driven by the relative productivities of primary and social production, which is to say, their relative importance, broadly construed, for human well-being in a given context.

There are many episodes in history where human settlements transitioned from dispersed to aggregated, and presumably this is a signal of a change in the relative value of social vs. primary production for the residents. This can occur for a variety of reasons, including an increase in the social cost of not being aggregated (e.g., due to warfare); a decrease in transport costs for staples (due to animal traction, wheeled vehicles, roads, etc.); an increase in land productivity which changes the balance of costs of transport for primary vs. social production; or an increase in the contribution of exchange to household incomes, which can derive from increases in community size and density. In addition, while one can conceptualize transitions between dispersed and aggregated forms of settlement in general systems terms, in real human communities a variety of political or ideological factors can keep people from adopting an energetically balanced form of settlement given prevailing conditions. So, there are opportunities for archaeologists to study deviations from general equilibrium conditions in addition to factors that shift the equilibrium in a given context.

For all this work, however, there is a fundamental issue: the identification of community territories. It is relatively straightforward to define the boundaries of aggregated settlements by finding the extent of built space and/or artifactual remains. However, to define the community boundary one needs to know the extent of agricultural land used by the residents, and this is often difficult to determine from the village remains themselves [@varien1999]. In contrast, it is often quite difficult to determine the boundaries of dispersed communities due to gaps in survey coverage or relatively consistent distributions of farmsteads, but once one has determined the community territory it is straightforward to define the associated agricultural land.

In this paper, we present a method we developed to solve such problems, which enables us to compare the properties and resilience of dispersed vs. aggregated communities on the same landscapes. We draw on three extensive compilations of archaeological data from different portions of the ancestral Pueblo region of the US Southwest, using the locations of what we call community centers as tethers for defining territories that we infer were used by social communities over extended periods of time [@glowacki_ortman2012; @varien_etal2007]. In some regions these are sites containing more households than could plausibly be related by reckoned kinship or some form of civic architecture. The key feature of these centers is that they represent locations against which travel time distances to other sites can be compared.

While not strictly required by our algorithm, we make the simplifying assumption that once established, community territories were relatively fixed in space for the duration of their associated sites. This allows us to use the distribution of all recorded residences, regardless of their periods of occupation, in defining community territories. There are several reasons why we believe this is a reasonable simplification. First, there are strong cross-cultural regularities in the time individuals spend in daily travel from and to their residences, so dispersed farming communities tend to be dispersed across the distance associated with this typical travel time, regardless of the number of households involved [@marchetti1994]. Second, the archaeological record of our study areas supports the idea that community territories did not shrink when their populations aggregated into villages. In some cases, farmsteads were converted to field houses, with some building materials being reused elsewhere and evidence for continued limited activity use [@varien1999; @varien1999a; @varien2002]. In other cases, grid gardens were constructed directly within older rubble mounds by residents of newer, aggregated villages nearby [@gauthier_herhahn2005]. Third, an analysis that allowed community territories to vary by time step would reduce variation in the density of the resulting units, thus washing out variation in one of the most important social properties one would ideally want to examine. Fourth, in areas where there is strong survey coverage, dispersed residences form clusters, often but not always centered on a few larger settlements, and this clustering is apparent even when the dispersed residences are plotted by time step [@schachner2012; @varien2002]. Fifth, the effects of population growth for dispersed farming communities are not parallel to the effects of population expansion in aggregated settlements. In the latter, the area of the settlement must grow somewhat because the residential density is already relatively high. But in dispersed settlements, there is often unused land between existing farmsteads that can be settled. This, combined with the strong cross-cultural regularity in typical commute times, leads to much greater consistency in the spatial extent of residents in a dispersed community than in the case of an aggregated community. Finally, given the inherent incompleteness of the archaeological record, it seems most appropriate to use all remains to define what is effectively the maximum extent of community territories. While it was undoubtedly true that individual community territories expanded or contracted over time, the assumption of constant area seems more reasonable than the assumption of constant density.

# Data

The datasets we apply our algorithm to come from three regions in the US Southwest: Central Mesa Verde (CMV), the Northern Rio Grande (NRG), and the Cibola or Zuni region (CIB) (see @fig-overview). The CMV and NRG datasets derive from Phase II of the Village Ecodynamics Project [@ortman2016; @schwindt_etal2016]. The CIB data derive from data syntheses produced by Schachner and Peeples [@peeples2018; @peeples_schachner2012; @schachner2012]. All of these data are now included in cyberSW [@mills_etal2020]. For summaries of the archaeological and environmental context in each region, see @tbl-regions.

![
Overview map showing the locations of farmsteads (black dots) and community centers (red triangles) in Central Mesa Verde, Northern Rio Grande, and Cibola.
](../figures/overview-maps.png){#fig-overview layout-align="center" width="100%"}

The CMV area encompasses the southwestern corner of Colorado, from the Mesa Verde cuesta in the south and east to the Dolores River in the north, and the Utah state line in the west. The elevation ranges from roughly 1,399.8 m to 3,032.8 m, with an average elevation of 1,999.0 m ($\sigma$ = 276.5 m). SKOPE [@bocinsky_etal2022] estimates suggest annual precipitation in the CMV study area has averaged around 39.5 cm ($\sigma$ = 9.1 cm) over the last 2,000 years, and it puts summer growing degree days for maize around 2,390.1 °F ($\sigma$ = 97.9 °F). Across most areas in the region, it is believed that direct precipitation was sufficiently high to support dry farming in most years [@bocinsky_kohler2014; @bocinsky_etal2016].

The NRG study area in north central New Mexico includes basically all of the Rio Grande valley between the Jemez Mountains to the west and the southernmost mountains of the Sangre De Cristo range to the east, with the Rio Grande itself running along the western edge of the valley at the foot of the Pajarito Plateau. The elevation ranges from roughly 1,587.9 m to 3,836.2 m, with an average elevation of 2,220.0 m ($\sigma$ = 371.3 m). SKOPE estimates the average annual precipitation in the NRG over the last 2,000 years to be around 40.6 cm ($\sigma$ = 8.2 cm), and it puts summer growing degree days for maize around 2,154.4 °F ($\sigma$ = 102.1 °F). It is argued that direct rainfall on the Pajarito Plateau made dry farming possible there, but in the valley farming required floodwater irrigation along watercourses [@duwe_anschuetz2013; @bocinsky_kohler2014; @bocinsky_etal2016; @ortman2016a].

The CIB region straddles the southern slope of the Colorado Plateau along the border between Arizona and New Mexico. Although this is a rugged terrain by any reasonable standard, it is notably less so than the other study areas. The elevation ranges from roughly 1,771.3 m to 2,777.1 m, with an average elevation of 2,145.9 m ($\sigma$ = 160.0 m). SKOPE estimates suggest annual precipitation in the CIB area over the last 2,000 years has averaged around 34.3 cm ($\sigma$ = 7.6 cm), and it puts summer growing degree days for maize around 2,317.3 °F ($\sigma$ = 93.8 °F). However, no area in the region receives sufficient direct rainfall to support dry farming, so some form of water management was required [@kintigh1985; @muenchrath_etal2002].

::: {#tbl-regions}

```{=latex}

\begin{singlespace}
\small
\begin{tabular}{lllll}
    & \textbf{Area [km2]} & \textbf{Dates [CE]} & \textbf{Archaeology} & \textbf{Environment} \\ 
    \hline  
    \noalign{\vskip2mm}  
  CMV & 4,556.4             & 725-1280            & \begin{tabular}[c]{@{}r@{}}Rooms: \hfill 67,746\\ Farms: \hfill 5,415\\ Centers: \hfill 512\end{tabular} & \begin{tabular}[r]{@{}r@{}}Precipitation: \hfill 39.5 cm ($\sigma$: 9.1)\\ Maize GDD: \hfill 2,390.1 $^{\circ}$F ($\sigma$: 97.9)\\ Elevation: \hfill 1,999.0 m ($\sigma$: 276.5)\end{tabular} \\[5mm]  
  \hline  
  \noalign{\vskip2mm}  
  NRG & 6,958.0             & 900-1550            & \begin{tabular}[c]{@{}r@{}}Rooms: \hfill 46,226\\ Farms: \hfill 2,316\\ Centers: \hfill 181\end{tabular} & \begin{tabular}[r]{@{}r@{}}Precipitation: \hfill 40.6 cm ($\sigma$: 8.2)\\ Maize GDD: \hfill 2,154.4 $^{\circ}$F ($\sigma$: 102.1)\\ Elevation: \hfill 2,222.0 m ($\sigma$: 371.3)\end{tabular} \\[5mm]  
  \hline  
  \noalign{\vskip2mm}  
  CIB & 7,424.1             & 700-1540            & \begin{tabular}[c]{@{}r@{}}Rooms: \hfill 27,234\\ Farms: \hfill 713\\ Centers: \hfill 78\end{tabular}    & \begin{tabular}[r]{@{}r@{}}Precipitation: \hfill 34.3 cm ($\sigma$: 7.6)\\ Maize GDD: \hfill 2,317.3 $^{\circ}$F ($\sigma$: 93.8)\\ Elevation: \hfill 2,145.9 m ($\sigma$: 160.0)\end{tabular} \\[5mm] 
  \hline
  \noalign{\vskip2mm}
\end{tabular}
\end{singlespace}

```

Regional Summaries

:::

# Methods

Our clustering algorithm draws inspiration from the procedure used by the United States Office of Management and Budget to define Core Based Statistical Areas (CBSA) [@usomb2021]. We start by defining a core area, in this case a community center location. We then associate neighboring residential sites or farmsteads with this center based on their commute times, and we merge core areas into larger agglomerations based on the proportion of their populations that they share. Our method may be described as a guided density-based clustering algorithm in that it does not rely on a random set of points when initialized. Instead, the selection of core areas is guided by archaeological data and regional expertise. We think this is a key argument in its favor.

From an archaeological perspective, one can think of our algorithm as combining insights from @varien1999 and @reese_etal2019, specifically Varien’s idea of a community catchment, which is an isochrone defined by a uniform commute time in all directions from a community center, and Reese’s suggestion that a spatial or geographic community can be identified by grouping individuals in terms of the commute distances between them. Our way of combining these ideas is to associate farm sites with specific community centers in terms of their commute distances and then draw the community catchment around those associated farms. We also share with them and others [@lipe_hegmon1989; @mahoney_etal2000; @peterson_drennan2005] a focus on "geographic" as opposed to "purely social" communities; that is, communities constrained in space and comprised of individuals that can be expected to interact on a regular, even daily, basis.

In broad outline, our proposed community detection algorithm has the following steps:

1. Identify community centers
2. Identify farms within a commute time threshold ($D{\text -}max$)
3. Join communities to their overlapping neighbors ($P$, $D{\text -}join$)
4. Exclude farms closer to another community center (tie-breaking step)
5. Draw smallest concave hull encompassing all farms, centers, and paths

\noindent After step 4, we also apply a filter to remove communities that have less than a minimal number of dispersed farmsteads, specifically four. Although this is a somewhat arbitrary threshold, we note that three vertices are the bare minimum required to define a polygon and measure its area. By removing small communities, we also minimize potential underestimates of total arable land area. In addition, our larger ambition is to understand processes occurring within dispersed farming communities, not necessarily to define all the dispersed communities that once existed in an area. For community centers with few or no farms, there is no meaningful dispersed farming community to examine, regardless of whether this is due to limited survey coverage or actual past behavior, hence our desire to exclude them.

The primary tuning parameters in our algorithm are $P$, the proportion of a community population used to join communities; $D{\text -}join$, the commute time required to calculate P; and $D{\text -}max$, the commute time used to exclude distant farms. These are described in more detail below.

## Identify community centers

**Rule:** Any site, whether residential or otherwise, with known and persistent interaction between unrelated individuals should count as a community center.

Why persistent interactions? Intuitively, a community center is a place where individuals from different households cross paths with the intention to interact. A simple way to demonstrate that intention is to show that individuals would still be willing to pay some non-negligible cost to visit those centers under various counterfactual changes to local conditions. This serves to distinguish them from merely accidental exchanges or chance encounters. Of course, there can be some path-dependence to this process, with locations of ephemeral or one-off encounters coming to be locations of persistent interaction. The point is that the location will come to be a community center so long as individuals continue to visit that location despite the underlying variance in their circumstances.

Why unrelated individuals? The simple answer is that a community is not a family. As a unit of social organization, it does share something in common with a family, namely, a set of shared interests; but unlike in a family, those shared interests form only a small part of a much larger set of potentially conflicting goals and desires. As a consequence, communities are more susceptible to problems with coordination or collective action. On the other hand, the mere existence of a community center implies some level of coordination - even if that coordination is just about the location of the community center itself! So, community members share less in common than families, but more in common than a purely random sample of the larger population.

We can summarize these ideas by noting that persistent interactions at a location make that location a *center*, and persistent interactions between unrelated individuals at that location make it a *community* center. To make these considerations more useful for defining community boundaries, however, we need to operationalize them, to make them applicable to specific archaeological contexts, and that requires a deeper understanding of the archaeology in those contexts. For the two regions represented by the VEP II dataset, we rely on two different but related sets of criteria to identify community centers. In the Central Mesa Verde area, a site is considered a community center if it includes civic architecture, specifically a great kiva or great house, or eight or more pit structures (small kivas) representing eight or more households. In the Northern Rio Grande, a site counts as a community center if it is classified as a village (a site with 50 or more rooms) or town (a site with 500 or more rooms). In the Cibola region, a community center includes at least 50 rooms or civic architecture. The large number of households in these regions are thought to be too large to be plausibly related by reckoned kinship.

## Join farms within a commute time

**Rule:** For each center, join all farms within commute time, $D{\text -}max$.

Following @reese_etal2019, we define nearness or proximity in terms of commute time rather than geographic distance. This is done for the obvious reason that the primary opportunity cost associated with daily pedestrian movement is the time spent walking, during which other productive activities cannot take place. There is also a much stronger cross-cultural regularity in the length of time individuals devote to commuting than the distance covered, as the latter is a function of the speed of movement, which is influenced by transport technology [@marchetti1994]. As a result, both the frequency and intensity of interaction at community centers should decay with commute time. The longer it takes to get to a community center, the less frequent and less intense the interactions at that center should be [@varien1999; @peterson_drennan2005].

While different constraints will naturally arise for different modes of transportation, leading to differences in both the routes taken and the speed and distance covered, for the populations considered here the only available mode of transportation was pedestrian. The biggest obstacle to interaction was thus the landscape itself, in particular its topography - intuitively, the steeper the uphill slope, the slower the hiking speed, the longer the commute time. So, all else being equal, individuals at a farm that is equidistant (in terms of geographic distance) between two community centers should prefer traveling to the one separated by less rugged terrain.

The concept of commute time is a critical variable in many places in our algorithm, but as a first approximation, we incorporate the idea by setting a commute time threshold, $D{\text -}max$. This variable, in effect, defines an isochrone around each community center, what @varien1999 calls a "catchment." All farm sites that fall within that isochrone get associated with its community center. This threshold also serves as a filter, as we drop all farm sites that are farther than that commute time to any center. Doing so helps to minimize potential overestimates of total arable land area for each community. Here, we have defined $D{\text -}max$ as one hour. This is about twice the median daily commute time observed in cross-cultural studies and it includes roughly 95% of the farmsteads in each region (see @fig-commutes).

For details of how we estimate commute times over costly terrains, see “Least-cost path analysis,” below.

## Join centers with overlapping neighbors

**Rule:** For centers $C_1$ and $C_2$ with nearby farm populations $N_1 ≤ N_2$, if $P \cdot N_1$ are within distance $D{\text -}join$ of $C_1$ and $C_2$, then join $C_1$ to $C_2$.

The larger and more dense farming populations become, the harder it gets to tell them apart and, more importantly, the harder it gets to justify keeping them apart, so we need to articulate a rule that specifies when two communities should be merged into one. For CBSAs, that decision hinges on a distinction between central and outlying counties, or counties where some fraction of the population live in an urban core and counties where some fraction commute to and from a central county on a regular basis. Basically, if the central counties of one CBSA qualify as the outlying counties of another CBSA, then the two CBSAs are merged. The intuition here is that when rates of interaction and exchange between CBSAs are roughly equal to rates within a CBSA, the CBSAs should be merged (USOMB 2021).

We carry that intuition over to our merger rule for dispersed maize farming communities, albeit with some modification to account for the fact that we are working with point locations rather than administrative boundaries. We assume that if some proportion, $P$, of the population related to one community center lives within a certain commute time, $D{\text -}join$, of another community center, that those two community centers are effectively parts of the same dispersed community. We do this in the direction of letting larger communities absorb smaller ones.

More precisely, the rule requires that $P \cdot N_1$ be within $D{\text -}join$ of *both* $C_1$ and $C_2$. This promotes joins when $C_1$ and $C_2$ are themselves in close proximity to each other, as the farther apart they are, the smaller the number of sites that overlap within $D{\text -}join$. It also tends to encourage joining two communities when a substantial proportion of the population lives between the two centers. In either case, the result is what @varien1999 sometimes refers to as a "macro-community" or "multi-community cluster," meaning a community organized around multiple community centers.

The rule was partly inspired by the density-based clustering algorithm known as DBSCAN [@ester_etal1996], as we rely on a moving window (an isochrone, in this case) set around each center, and look for some measure of density to guide join-decisions. As noted above, however, our method does not rely on a random initialization of points, but rather targets community centers. It is also hierarchical in the sense that it exploits additional structure in our data, namely the distinction between community centers and farm sites.

As we cannot measure population directly, however, we use information we have about farm sites, specifically estimates of their room counts, as a proxy. For many sites in the CMV, pit structures or small kivas are more reliably identified than surface rooms. Fortunately, each pit structure is typically the central roofed space of a unit pueblo residence that includes an average of six surface rooms [@adler1990], so we simply multiply the number of pit structures by six to obtain room estimates for each site. CIB and NRG sites all have room estimates, so no additional assumptions are made about them. Finally, we follow the well-established rule of thumb in US Southwest archaeology that each surface room in a settlement represents a single person, leading to a one to one conversion from rooms to residents [@duwe_etal2016; @kintigh1985; @lekson1989; @lipe1989; @ortman2016].

For this analysis, we specify $D{\text -}join$ to be one half hour and $P$ to be 0.8. We acknowledge that these values are somewhat arbitrary. In their defense, they set an extremely high bar, so joining centers that satisfy the rule seems reasonable.

## Exclude farms closer to another community center

**Rule:** For each farm that falls within $D{\text -}max$ of multiple community centers, associate it with the community center to which it has a shorter commute time.

On occasion a farm may fall within $D{\text -}max$ of multiple community centers that are not joined into a single community, which makes them members of multiple communities. For the dispersed farming communities we aim to identify, this would imply that multiple communities could simultaneously farm the same plot of land, which is unrealistic, so we need a rule for breaking ties. Because we are already relying so heavily on commute time, the simplest solution is to associate a farm that falls within $D{\text -}max$ of multiple community centers with the center and, thus, the community that it is closer to, and this is what we have elected to do. We can justify this step on similar grounds as well, with the safe assumption that individuals should be more likely to interact the nearer they are in time and space. We also note that this tie-breaking rule ensures that community polygons are continuous, that they do not contain any holes, which would be quite odd considering these community boundaries are supposed to represent the spatial extent of their arable land. It also encourages some separation between communities, preventing them from overlapping to any significant degree.

## Draw community boundary

**Rule:** For each community, draw a polygon that encompasses all farms, community centers, and commute paths using a concave hull.

So far, we have only grouped site points into clusters or communities. The goal, however, is to define their spatial extent. That means, for each set of site points, that we need some way of drawing a polygon that encompasses all of them. This can be done in a number of ways. The simplest strategy would be to find the centroid of the set of points and draw a circle with the smallest radius that still includes them all. This would be utterly arbitrary, however, and frankly unrealistic, not representing any sort of meaningful boundary given the landscape. An alternative would be to use the convex hull of the set of points, but this tends to exaggerate the total area of the community, especially when the distribution of points is suggestive of a concave shape, like the letter ‘C’. A concave hull would handle peculiar shapes, but it would also introduce the opposite problem, restricting the area of the community to an unreasonable degree. Our somewhat brute force way of threading this needle is to identify paths with shortest commute times between every pair of farms in a community, extract the vertices from those paths, and then incorporate those into the set of points used to define the concave hull. The result is a polygon that is concave as well in the dimension of time, meaning the paths with shortest commute times between any two farms never leaves the community.

In some cases, this may lead to overlapping community boundaries, though the amount of overlap should be minimal given the way that we cluster points, particularly when there are many densely distributed points. To some extent, of course, this sort of overlap is both unavoidable and expected as even in the modern world there are disagreements about the exact locations of legal and political boundaries.

It should be noted, too, that we also dilate each concave hull, briefly expanding it with a positive buffer, then shrinking it with a negative buffer having a slightly smaller size. This should have only a minimal effect on the total area of each polygon and also serves to smooth out noise along the edges.

## Least-cost path analysis

The key to our path analysis is to take information about the terrain in our study regions and convert it first into meaningful estimates of hiking speed and then into meaningful estimates of travel time. Fortunately, attempts to model hiking speed across a range of slopes have advanced considerably in recent years, most notably in @campbell_etal2019. Drawing on a large, crowd-sourced fitness tracker data set consisting of approximately 420,000 individual hikes, jogs, and trail runs from nearly 30,000 individuals in Salt Lake City, UT, @campbell_etal2019 offer what is easily the most empirically robust estimate of hiking speed to date. An interesting feature of their analysis - though one less important for the current study - is its sensitivity to variance in the physical and athletic prowess of individuals, which allows for estimates of different speeds at a given slope for different deciles of the sample. For simplicity, our application of Campbell’s hiking function relies on coefficients from the first decile of individuals in their sample (the slowest ten percent of hikers) and assumes that everyone in the areas of our analysis were largely equal in their walking speed and endurance. This makes their walking speeds comparable to, but not the same as, what would be estimated by Tobler’s hiking function [@tobler1993].

We take the following steps to calculate least-cost travel times in our study regions. We first download a digital elevation model for each study area from the USGS National Elevation Dataset using the FedData package in R [@bocinsky2020]. The original resolution of these gridded data is 30 m, but we aggregate grid cells by a factor of 3, making them closer to 100 m in resolution. Aggregation is a familiar speed-up technique for least-cost analysis on large grids containing tens of millions of cells. We believe the resulting loss in precision owing to such aggregation is reasonable given that our goal is not so much to find specific paths as it is to determine relations between the locations that these paths connect.

The next step is to convert the aggregated elevation data into a graph in which each node represents a grid cell and each edge represents a straight line between adjacent grid cells in the Moore neighborhood (the eight adjacent grid cells). Edge weights are first defined by the slope between adjacent grid cells. Those slope estimates are then fed to Campbell’s hiking function to derive estimates of hiking speed along each edge. The inverse of the hiking speed (the pace) is then multiplied by the distance between adjacent grid cell centers to estimate travel time.

Next, we associate each farm and community center with the grid cell that contains it, and hence to a specific node in the graph. In effect, we treat all the sites falling into a grid cell as a single site. We then apply a bi-directional form of Dijsktra’s algorithm to the entire set of nodes with associated sites, as this algorithm significantly reduces processing time when the desired outcome involves every combination of start and end points [@sturtevant_felner2018]. This gives us two datasets: (i) a dense distance matrix with least cost travel-times from all farms and community centers to all farms and community centers and (ii) a set of linestring geometries whose vertices are the grid cells traversed on shortest paths.

For simplicity, we average travel times to and from each origin and destination point, thus making the distance matrix symmetric, which effectively treats the round-trip commute as the relevant cost-distance. All of the clustering steps in our algorithm are implemented as operations on this matrix. 
All analyses, including the cost-distance modeling and our implementation of the algorithm, are conducted in the R programming language and environment [@rcoreteam2023]. For more details, please see Supplementary Materials, which include all code necessary to reproduce these analyses and the figures in the text.

# Results

The goals of this analysis are (i) to identify dispersed farming communities based on the distribution of farmsteads and their proximity to community centers and then (ii) to regionalize those communities by giving them meaningful boundaries. This was done by developing a simple algorithm inspired by the concept of a CBSA combined with previous work by @varien1999 and @reese_etal2019. Using parameter values D-join = 30 minutes, D-max = 1 hour, and P = 0.8, our algorithm identified 85 communities in CMV, 55 communities in NRG, and 21 communities in CIB (see @fig-communities). The mean areas of these communities are 11.56 km$^2$ ($\sigma$ = 7.54, $\eta$ = 10.5) in CMV, 7.20 km$^2$ ($\sigma$ = 5.56, $\eta$ = 5.86) in NRG, and 8.80 km$^2$ ($\sigma$ = 7.66, $\eta$ = 5.93) in CIB; the mean room counts (for all community centers and farms) are 781.62 ($\sigma$ = 991.89, $\eta$ = 420.0) in CMV, 627.73 ($\sigma$ = 594.33, $\eta$ = 443.0) in NRG, and 747.71 ($\sigma$ = 573.92, $\eta$ = 555.0) in CIB; and the mean room densities are 74.92 ($\sigma$ = 67.1, $\eta$ = 52.66) in CMV, 138.58 ($\sigma$ = 150.07, $\eta$ = 92.5) in NRG, and 158.04 ($\sigma$ = 179.00, $\eta$ = 92.32) in CIB. We emphasize that the means in several cases deviate to a considerable degree from median values (denoted above with the Greek letter $\eta$), indicating skew in the data and the potential presence of large outliers. This is illustrated in the boxplots in @fig-densities, which show the distributions of these and other variables of interest that can be derived from our analysis and also used to evaluate its merits (see also @tbl-results).

![
Map showing the size, shape, and spatial distribution of community polygons. Color represents the log density of rooms (the number of rooms per square kilometer), with lighter colors indicating lower values and darker colors indicating higher values. Note that densities in each region were re-scaled to have the same range, so this figure cannot be used to evaluate absolute differences between regions!
](../figures/communities-all.png){#fig-communities layout-align="center" width="100%"}

In addition to showing the community polygons, @fig-communities also offers a simple demographic profile for each region based on the spatial distribution of all rooms (their count per square kilometer) across communities irrespective of time period. This provides a simple demonstration of the utility of this method. For visualization purposes, the log of the density is used, as it helps to normalize the distribution and, in particular, to rein in large outliers that swamp the variance. It is important to note, too, that values in each region are re-scaled to be in the same range, so the figure cannot be used to estimate absolute differences between regions. Still, the relative differences within each region can be enlightening, but even they must be interpreted with caution.

Not surprisingly, virtually all of the Mesa Verde cuesta communities show high densities. The community that includes Goodman Point and Shield Pueblo in the central area of the CMV region just north of Ute Mountain also has a relatively high density, as do the communities in the far northwest that include Turkey House, Lowry, and Cottonwood Ruin. The Dolores area community around Grass Mesa Village in the northeast is also high density. In the NRG, the highest density community is a narrow band that lies in a shallow canyon running south from Cundiyo on the east side of the valley. The thin east-west running community near San Ildefonso Pueblo is also high density. Note that it crosses the Rio Grande, following the contours of the Pojoaque River drainage. Just west of there on the Pajarito Plateau, there is a high density community that includes the Sandia center. South of Sandia there are two additional high density communities, one near the center of the Pajarito Plateau and another further south along the Chiquito Rio northwest of Cochiti Lake. In the CIB area, the three densest communities can be found in the far eastern edge of the study area near El Morro and Tinaja. One dense community just to the west of there includes the Ramah School center. Another lies in the far northwest near Manuelito.

::: {#tbl-results}

```{=latex}

\begin{singlespace}
\small
\begin{tabular}{lrllllll}
    &                                                                   & \textbf{\begin{tabular}[c]{@{}l@{}}Area\\ (km2)\end{tabular}}     & \textbf{\begin{tabular}[c]{@{}l@{}}Rooms\\ (N)\end{tabular}}        & \textbf{\begin{tabular}[c]{@{}l@{}}Farms\\ (N)\end{tabular}}     & \textbf{\begin{tabular}[c]{@{}l@{}}Centers\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Commute\\ (mins)\end{tabular}}  & \textbf{\begin{tabular}[c]{@{}l@{}}Room Density\\ (N/km2)\end{tabular}} \\
    \hline
    \noalign{\vskip2mm}
CMV & \begin{tabular}[c]{@{}l@{}}$\eta$\\ $\mu$\\ $\sigma$\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 10.5\\ : 11.56\\ : 7.54\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 420\\ : 781.62\\ : 991.89\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 28\\ : 59.29\\ : 85.78\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 1\\ : 6.01\\ : 13.33\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 21.61\\ : 22.35\\ : 8.46\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 52.66\\ : 74.92\\ : 67.1\end{tabular}      \\[5mm]
  \hline
  \noalign{\vskip2mm}
NRG & \begin{tabular}[c]{@{}l@{}}$\eta$\\ $\mu$\\ $\sigma$\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 5.86\\ : 7.2\\ : 5.56\end{tabular}   & \begin{tabular}[c]{@{}l@{}}: 443\\ : 627.73\\ : 594.33\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 20\\ : 39.53\\ : 37.63\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 2\\ : 2.62\\ : 2.75\end{tabular}  & \begin{tabular}[c]{@{}l@{}}: 16.59\\ : 17.41\\ : 7.21\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 92.5\\ : 138.58\\ : 150.07\end{tabular}    \\[5mm]
  \hline
  \noalign{\vskip2mm}
CIB & \begin{tabular}[c]{@{}l@{}}$\eta$\\ $\mu$\\ $\sigma$\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 5.93\\ : 8.8\\ : 7.66\end{tabular}   & \begin{tabular}[c]{@{}l@{}}: 555\\ : 747.71\\ : 573.92\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 19\\ : 28.24\\ : 21.96\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 2\\ : 2.29\\ : 1.55\end{tabular}  & \begin{tabular}[c]{@{}l@{}}: 18.52\\ : 19.64\\ : 9.11\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 92.32\\ : 158.04\\ : 179\end{tabular} \\[5mm]
  \hline
  \noalign{\vskip2mm}
\end{tabular}
\end{singlespace}

```

Community Summaries

:::

# Discussion

Our approach offers several advantages over previous efforts to define dispersed community territories. First, it handles variation in survey coverage by seeking to define communities we have evidence for, rather than all communities that may once have existed. Second, it handles variation in the degree to which surveyors have lumped or split architectural remains in defining sites [@kintigh2003]. This is a special problem in the CMV. The tradition of site survey recording within the Mesa Verde cuesta has generally involved treating each room block as a separate site, even when there is a continuous artifact scatter between them. The tradition in the Montezuma Valley, in contrast, has been to include all room blocks within a single artifact scatter as part of a single site. Our method is robust to these differences in site definition, enabling properties of the resulting communities to be compared more directly across these areas. Finally, our method focuses on travel time, a strong regularity in human affairs, and can be applied to landscapes with dramatically different topographies. The resulting community territories take obvious natural topographic barriers to interaction into account, but also allow territories to cross such boundaries when this is suggested by the data themselves.

![
Figure shows the difference between the log density of commute times (in  minutes) from observed farms and random locations to the nearest community center for each region, along with the location of their 95% quantile (this is just the 95% quantile for observed farms, not random points).
](../figures/commute-kl.png){#fig-commutes layout-align="center" width="60%"}

Because our algorithm requires that we calculate the commute time between each farm and community center, we can use it to evaluate potential spatial interactions between them. @fig-commutes tries to get at this idea by showing the difference between (i) the log probability density of commute times from observed farms to their nearest centers and (ii) the log probability density of commute times from random locations to their nearest centers, along with the 95% quantiles of the observed distributions. This is analogous to what @reese_etal2019 refer to as the "null difference cost distance." One can interpret it as showing where commute times differ from what one might expect were farm sites distributed at random across the landscape, making it similar to the Kullback-Leibler Divergence - basically, what a point process model like MaxEnt tries to estimate [@elith_etal2011; @merow_etal2013]. Values much larger than zero indicate that there are more farms located at those commute times from centers than you would expect by chance. Conversely, values less than zero indicate that there are less farms located at those locations than one might expect by chance. In either case, these differences tell us that there is positive spatial covariance between farmsteads and nearby centers at commute times of less than approximately one hour (the value we asisgned to $D-max$).

![
Figure shows for each region the distribution of relevant variables across communities.  
](../figures/results-boxplot.png){#fig-densities layout-align="center" width="75%"}

The shape of the responses are not, however, identical. We attribute this to three factors: sampling intensity, local topography, and regional settlement patterning. For instance, the large spike in the CIB region at or around the 30 minute point along with the longer tail (and the larger 95% quantile) is at least partially owing to the fact that block survey in that area is limited to a few disconnected locations. The area is also much flatter than in the CMV and NRG regions, making the background commute distribution much more uniform, at least within the surveyed areas. We probably could have accounted for the limited sampling by sampling random background points closer to the observed site locations, but that is not really the point of this exercise.

Perhaps more interesting though is the fact that these regions represent different settlement patterns over different periods of time. In the CMV area, the pattern of settlement was largely one of growth and dispersal, with local populations reaching an asymptote after which they spread out into neighboring areas [@schwindt_etal2016]. Nearly all centers in the CMV area arose within farming clusters, but they were relatively small compared to those in the other study areas, and occurred only to a limited degree in the period leading up to the Great Drought in 1280 CE (and to a lesser degree the period leading up to and including much of the 10th century), with no large aggregations occurring afterwards [@varien1999; @glowacki_ortman2012; @ortman2016a]. In the NRG region, communities had a few dispersed farms early on, more in the period immediately following 1280 (mostly concentrated in the northern Pajarito Plateau), and then very large aggregations after that [@ortman2016; @ortman2016a]. Communities in the CIB region had almost no dispersed farming for most of the sequence, mostly skipping over that pattern and moving straight into large aggregations as populations surged in the period after 1280 [@kintigh1985; @schachner2012; @peeples2018], as suggested by the relatively small ratio of farms to centers (see @fig-densities). As in the CMV area, aggregations in the NRG and CIB regions sometimes developed within dispersed farming communities, but unlike in the CMV, many also seem to have been established in areas away from where dispersed farms had been previously located [@schachner2015; @ortman2016a]. This would explain why our algorithm dropped 29 of the 77 centers in the CIB region and 37 of the 181 centers in the NRG but only 1 of the 512 centers in the CMV.

It should be emphasized that this is perhaps the biggest limitation of our algorithm, that it simply cannot define community boundaries when the relationship between community centers and dispersed farms is not obvious, either because the center is geographically isolated from the farms or because there is no center at all. We note, however, that our concept of a community center does not technically require the presence of an archaeological site. That is just our way of identifying the locations of centers in the archaeological record.

In [@reese_etal2019], the area of the community is allowed to expand and contract, with the average area per farm remaining relatively constant on the Mesa Verde cuesta, in the range 0.15 to 0.27 km$^2$ per farm. In our algorithm, the total area of the community remains fixed, but the density is allowed to change. With the tuning parameter values we chose, the algorithm put about 50% of CMV communities in the range 0.2 to 0.55 km$^2$ per farm, with similar spreads for CIB and NRG. That, however, represents a minimum possible level in our analysis and would likely be greater as only a proportion of the farms were actually occupied at a given time. Still, this means that the communities we have defined are more likely to meet the 0.14 km$^2$ acre threshold required for a household to subsist on maize [@benson2011; @bocinsky_varien2017]. But, when we look at the area per room (the inverse of the room density), the picture is a little different. For the Mesa Verde region, the median area per room is roughly 0.04 km$^2$, which puts it at the upper end of the range reported in [@reese_etal2019]. In the NRG and CIB regions, however, the median area per room is closer to 0.01 km$^2$, with a first quartile closer to 0.005 km$^2$. Again, these are minimum thresholds assuming every room is occupied at the exact same time, so there was likely more elbow room at any given time. Stll, it does suggest that the area for farming in those regions may have been more circumscribed than in the CMV, which is also indicated by the boxplots showing total area in @fig-densities. That would imply that levels of intensification [*sensu* @boserup1965] in those regions were likely greater, too, with aggregation and social integration being one way that intensification probably played out.

# Conclusion

The primary goal of this analysis was to derive the boundaries of dispersed maize farming communities from the point locations of archaeological sites. To that end, we developed a method that leverages the difference between farmsteads and community centers to cluster sites into communities and define the boundaries of those communities. This was inspired by the concept of a CBSA, along with important contributions from @varien1999 and @reese_etal2019. An important difference between our approach and that of [@reese_etal2019] is that our community boundaries are static, though crucially this allows us to look at changes in density through time. So, the next step in our analysis will be to determine the number of rooms inhabited during each time step. A variety of methods are possible for this, but in future studies we intend to utilize uniform probability density analysis of pottery assemblages from each site [@ortman2016] to apportion rooms to time slices. This will allow us to capture the changing properties of both dispersed and aggregated communities using consistent measurements. The properties that can be compared include:

- Community area (agricultural land in the case of aggregated settlements);
- Settlement area (village area for aggregated communities, community area for dispersed communities);
- Community population by time step;
- Average distance between individuals (square root of community area / 2 for dispersed communities, and square root of village area / 2 for aggregated communities);
- Environmental conditions (captured by importing community territories into the paleoCAR niche modeling tool in <https://www.openskope.org/>);
- Topography (the intersection of community territories with a DEM);
- Properties of civic architecture found within the community, by time step; and
- Nearest neighbor distances to adjacent community centers.

\noindent The ability to compare these properties for communities of varying scales and densities is crucial for evaluating the relative costs and benefits of these properties for human welfare.

# Acknowledgements

Portions of this research were supported by a grant from the National Science Foundation (#2213921, to Ortman).

\pagebreak

# References

::: {#refs}
:::

### 2023-10-07 R/get-elevation.R

@@ -1,0 +1,41 @@
+
+# get elevation data
+
+library(FedData)
+library(here)
+library(purrr)
+library(sf)
+library(terra)
+
+download_dem <- function(region, geometry){
+  
+  r <- geometry |> 
+    st_sfc(crs = 4326) |> 
+    get_ned(res = "1", label = region) |> 
+    rast() |> 
+    aggregate(fact = 3)
+  
+  names(r) <- region
+  
+  return(r)
+  
+}
+
+write_dem <- function(x){
+  
+  fn <- here("data", paste0("dem-", names(x), ".tif"))
+  
+  writeRaster(x, filename = fn, overwrite = TRUE)
+  
+}
+
+# for some reason, values aren't read in after downloading the first time,
+# so have to call this twice
+here("data", "community-centers.gpkg") |> 
+  read_sf(layer = "regions") |> 
+  pmap(download_dem)
+
+here("data", "community-centers.gpkg") |> 
+  read_sf(layer = "regions") |> 
+  pmap(download_dem) |> 
+  walk(write_dem)

### 2024-02-14 R/community-clustering.qmd
add sensitivity analysis

@@
 
+```{r}
+#| echo: false
+
+# allow for timing code chunks
+# see https://bookdown.org/yihui/rmarkdown-cookbook/time-chunk.html
+knitr::knit_hooks$set(time_it = local({
+
+  function(before, options) {
+
+    if(before) {
+
+      assign('.start_time', Sys.time(), envir = .GlobalEnv)
+
+    } else {
+
+      proc_time <- difftime(Sys.time(), .start_time, units = "secs")
+      proc_time <- round(as.numeric(proc_time), 2)
+      proc_time <- lubridate::duration(proc_time, "seconds")
+
+      style <- "'text-align: right; font-size: 70%; color: rgba(55,58,60,.75)'"
+
+      txt <- paste0("Code execution time: ", proc_time)
+
+      paste0("<p style=", style, ">", txt, "</p>")
+
+    }
+
+  }
+
+}))
+
+```
+
 ## Introudction
@@
 
-This document is setup so that it focuses on the actual steps of the analysis, using verbose names for functions that should make their purpose clear.
-You can also dig into what each function does by unfolding the relevant code chunk.
-The critical R package for doing path analysis is [`cppRouting`](https://github.com/vlarmet/cppRouting).
-It's optimized for road networks, but still very fast.
-Someone should write a path-finding package (maybe using Rust?) optimized for grid networks...
+This document is setup so that it focuses on the actual steps of the analysis,
+using verbose names for functions that should make their purpose clear. You can
+also dig into what each function does by unfolding the relevant code chunk. The
+critical R package for doing path analysis is
+[`cppRouting`](https://github.com/vlarmet/cppRouting). It's optimized for road
+networks, but still very fast. Someone should write a path-finding package
+(maybe using Rust?) optimized for grid networks...
 
@@
 
-library(arrow)
 library(cppRouting)
+library(furrr)
 library(ggfx)
+library(ggh4x)
 library(ggnewscale)
@@
 
+```{r}
+
+# cppRouting uses as.character() on node IDs, which is fine,
+# but as.character(1000000) leads to "1e-6" instead of "1000000", and
+# this can lead to issues when matching IDs
+# to fix this, just remove the use of scientific notation
+options(scipen = 9999)
+
+# suppress progress bar and other terra output
+terra::terraOptions(progress = 0, print = FALSE)
+
+# for the sensitivity analysis, we run the entire workflow for each region in 
+# its own R session, hopefully to speed things up
+plan(multisession, workers = 3)
+
+```
+
 Some ggplot defaults
@@
 
-region_labels <- as_labeller(c(
+region_labels <- c(
   "cmv" = "Central Mesa Verde",
@@
   "cib" = "Cibola"
+)
+
+pretty_labels <- as_labeller(c(
+  "n_communities" = "No. of Communities",
+  "commute" = "Mean Commute to Nearest Center [mins]",
+  "room_density" = "Room Density [N/sq.km]",
+  "farm_area" = "Mean Area for Farms [sq.km/N]",
+  "center_area" = "Mean Area for Centers [sq.km/N]",
+  "farm_ratio" = "Ratio of Farms [N] to Centers [N]",
+  "area" = "Total Area [sq.km]",
+  "djoin" = "D-join [mins]",
+  "dmax" = "D-max [mins]",
+  "p" = "p [proportion in D-join]",
+  "agg_factor" = "Grid Resolution [m]"
+))
+
+squished_labels <- as_labeller(c(
+  "n_communities" = "No. of\nCommunities",
+  "commute" = "Mean Commute\nto Nearest Center\n[mins]",
+  "room_density" = "Room Density\n[N/sq.km]",
+  "farm_area" = "Mean Area\nfor Farms [sq.km/N]",
+  "center_area" = "Mean Area\nfor Centers [sq.km/N]",
+  "farm_ratio" = "Ratio of\nFarms [N] to Centers [N]",
+  "area" = "Total Area\n[sq.km]",
+  "djoin" = "D-join\n[mins]",
+  "dmax" = "D-max\n[mins]",
+  "p" = "p\n[prop. in D-join]",
+  "agg_factor" = "Grid Resolution\n[m]"
 ))
@@
 
-# trim all white space around image, but then add a few white pixels back (dx, dy)
+# trim all white space around image, 
+# but then add a few white pixels back (dx, dy)
+# all done with the magic of {magick}
 prepare_image <- function(x, dx = 2, dy = 30, color = "white") {
 
-  img <- magick::image_read(path = x)
-
-  img <- magick::image_trim(img)
-
-  info <- magick::image_info(img)
-
-  new_width <- info$width + dx
-  new_height <- info$height + dy
-
-  img <- magick::image_extent(
+  img <- image_read(path = x)
+
+  img <- image_trim(img)
+
+  info <- image_info(img)
+
+  new_width <- info[["width"]] + dx
+  new_height <- info[["height"]] + dy
+
+  img <- image_extent(
     img, 
@@
 
-  magick::image_write(img, path = x)
+  image_write(img, path = x)
 
@@
 
+The necessary data for the analysis include a polygon defining the region of
+interest, the point locations of farms and centers, and a digital elevation
+model (DEM).
+
 ```{r}
@@
 
-build_region <- function(gpkg, region){
+build_region <- function(region){
+
+  gpkg <- here("data", "community-centers.gpkg")
 
@@
 
-  if ("communities" %in% st_layers(gpkg)$name){
-
-    community_query <- paste0("SELECT * FROM communities WHERE region = '", region, "'")
-    t <- read_sf(gpkg, query = community_query) |> st_filter(r)
-
-  } else {
-
-    t <- NULL
-
-  }
-
-  list(
-    name = region,
-    preferred_projection = ifelse(region == "nrg", 26913, 26912),
+  dem <- here("data", paste0("dem-", region, ".tif")) |> rast()
+
+  region_list <- list(
     region = r,
     sites = s,
-    communities = t
-  )
-
-}
-
-add_elevation <- function(x){
-
-  dem_fn <- paste0("dem-", x$name, ".tif")
-  x$dem <- here("data", dem_fn) |> rast()
-
-  return(x)
-
-}
-
-# getters
-get_farms <- function(x){
-
-  x$sites |> 
-    st_drop_geometry() |> 
-    filter(center == 0)
-
-}
-
-get_centers <- function(x){
-
-  x$sites |> 
-    st_drop_geometry() |> 
-    filter(center == 1)
-
-}
-
-```
-
-```{r}
-
-gpkg <- here("data", "community-centers.gpkg")
-
-cmv <- build_region(gpkg, "cmv") |> add_elevation()
-nrg <- build_region(gpkg, "nrg") |> add_elevation()
-cib <- build_region(gpkg, "cib") |> add_elevation()
+    dem = dem
+  )
+
+  # save metadata
+  prj <- ifelse(region == "nrg", 26913, 26912)
+
+  class(region_list) <- "region"
+
+  attr(region_list, "name") <- region
+  attr(region_list, "preferred_projection") <- prj
+  attr(region_list, "grid_crs") <- st_crs(crs(dem))$epsg
+
+  region_list
+
+}
+
+```
+
+```{r}
+#| eval: false
+#| include: false
+
+# this is just for my own use while running this code interactively
+print.region <- function(x, ...){
+
+  attrs <- attributes(x)
+
+  n_farms <- x[["sites"]] |> filter(center == 0) |> nrow()
+  n_centers <- x[["sites"]] |> filter(center == 1) |> nrow()
+
+  list_stuff <- map(attrs[['names']], \(.y){
+
+    has_class <-  class(x[[.y]])[1]
+
+    paste0(.y, " (", has_class, ")")
+
+  }) |> unlist()
+
+  cli::cli_h2("Study Area Metadata")
+  cli::cli_text("Region: {attrs[['name']]}")
+  cli::cli_text("No. of farms: {n_farms}")
+  cli::cli_text("No. of centers: {n_centers}")
+  cli::cli_h3("Spatial Data")
+  cli::cli_text("Grid CRS: {attrs[['grid_crs']]}")
+  cli::cli_text("Preferred Projection: {attrs[['preferred_projection']]}")
+  cli::cli_text("Aggregation Factor: {attrs[['agg_fact']]}")
+  cli::cli_text("Max Slope: {attrs[['max_slope']]}")
+  cli::cli_h3("Clustering Parameters")
+  cli::cli_text("D-max: {attrs[['dmax']]} s")
+  cli::cli_text("No. of outliers: {attrs[['n_outliers']]}")
+  cli::cli_text("D-join: {attrs[['djoin']]} s")
+  cli::cli_text("P: {attrs[['proportion']]}")
+  cli::cli_text("Min. no. of farms: {attrs[['min_size']]}")
+  cli::cli_h3("Community Data")
+  cli::cli_text("No. of communities: {attrs[['n_communities']]}")
+  cli::cli_text("No. dropped: {attrs[['n_dropped']]}")
+  cli::cli_text("Concavity: {attrs[['concavity']]}")
+  cli::cli_text("Dilation: {attrs[['dilate']]} m (out, in)")
+  cli::cli_text("")
+  cli::cli_text("List includes: {list_stuff}")
+
+}
+
+```
+
+```{r}
+#| time_it: true
+
+cmv <- build_region("cmv")
+nrg <- build_region("nrg")
+cib <- build_region("cib")
 
@@
 
-e <- function(x){
+list(cmv, nrg, cib) |> 
+  map(\(x){
 
     tibble(
-    region = x$name,
-    min =  terra::global(x$dem, min, na.rm = TRUE)$min,
-    max =  terra::global(x$dem, max, na.rm = TRUE)$max,
-    mean = terra::global(x$dem, mean, na.rm = TRUE)$mean,
-    sd =   terra::global(x$dem, sd, na.rm = TRUE)$sd
-  )
-
-}
-
-bind_rows(
-  e(cmv),
-  e(nrg),
-  e(cib)
-) |> 
+      region = attr(x, "name"),
+      min =  global(x[["dem"]], min, na.rm = TRUE)$min,
+      max =  global(x[["dem"]], max, na.rm = TRUE)$max,
+      mean = global(x[["dem"]], mean, na.rm = TRUE)$mean,
+      sd =   global(x[["dem"]], sd, na.rm = TRUE)$sd
+    )
+
+  }) |> 
+  bind_rows() |> 
   gt() |> 
@@
 
-plot_region <- function(x){
-
-  hell <- function(x, n = 1000){
-
-    slope <- terrain(x, "slope", unit = "radians")
-    aspect <- terrain(x, "aspect", unit = "radians")
+plot.region <- function(x, ...){
+
+  slope <- terrain(x[["dem"]], "slope", unit = "radians")
+  aspect <- terrain(x[["dem"]], "aspect", unit = "radians")
 
   hill <- shade(slope, aspect, 30, 45)
-    hill <- setValues(hill, scales::rescale(values(hill), to = c(1, n)))
+  hill <- setValues(hill, scales::rescale(values(hill), to = c(1, 1000)))
   hill <- round(hill)
-
-    list(
-      shade = hill,
-      grays = hcl.colors(n, "Grays")[values(hill)]
-    )
-
-  }
-
-  h <- hell(x$dem)
 
@@
     geom_spatraster(
-      data = h$shade, 
-      fill = h$grays,
+      data = hill, 
+      fill = hcl.colors(1000, "Grays")[values(hill)],
       maxcell = Inf
     ) +
-    geom_spatraster(data = x$dem, alpha = 0.5) +
+    geom_spatraster(data = x[["dem"]], alpha = 0.5) +
     scale_fill_hypso_c(
@@
       geom_sf(
-        data = x$sites |> filter(center == 0),
+        data = x[["sites"]] |> filter(center == 0),
         shape = 16,
@@
       geom_sf(
-        data = x$sites |> filter(center == 1),
+        data = x[["sites"]] |> filter(center == 1),
         shape = 17,
@@
 
-gg <- (plot_region(cmv) + ggtitle("Central Mesa Verde")) +
-  (plot_region(nrg) + ggtitle("Northern Rio Grande")) + 
-  (plot_region(cib) + ggtitle("Cibola"))
+gg <- (plot(cmv) + ggtitle("Central Mesa Verde")) +
+  (plot(nrg) + ggtitle("Northern Rio Grande")) + 
+  (plot(cib) + ggtitle("Cibola"))
 
@@
 
-remove(plot_region, gg, fn)
+remove(plot.region, gg, fn)
 
@@
 ::: {#fig-overview}
-![](../figures/overview-maps.png){width="100%" fig-alt="An overview map showing the distribution of farms and community centers across the three study areas."}
-
-An overview map showing the distribution of farms and community centers across the three study areas.
+![](../figures/overview-maps.png){width="100%"
+fig-alt="An overview map showing the distribution of farms and community centers across the three study areas."}
+
+An overview map showing the distribution of farms and community centers across
+the three study areas.
 :::
@@
 
-It's important to note that throughout we track movement between grid cells, not site points.
+It's important to note that throughout we track movement between grid cells, not
+site points. So, all site points that fall into the same grid cell are assumed
+to be the same location, and if any of those sites are centers, then the whole
+place is considered a center.
 
@@
 
+Before converting the grid to a graph, we aggregate grid cells. Aggregating
+reduces the resolution of the grid and by extension the number of grid cells the
+raster contains. Since the grid cells are nodes in the grid graph, aggregating
+simplifies the graph as well. This speeds up the path analysis considerably, but
+it does come at the cost of precision in terms of path analysis across the
+landscape, but that is not as bad as it sounds. The idea that we could build a
+model that is perfectly isomorphic with the actual or true path taken by people
+a thousand years ago is wildly implausible. The idea of the 'true path' is also
+problematic as we're not evaluating a single commute, but rather the path people
+should take more often than not. So, the real question is what level of
+precision is sufficient for evaluating that. And the answer is, of course, it
+just depends.
+
+Are you walking blindfolded along the edge of a deep, precipitous cliff? Then
+centimeters are probably important to you - really, really important to you. But
+if you're just trying to get a reasonable estimate of the agricultural catchment
+of an entirely pedestrian population based on the distribution of farms around
+urban centers, probably the difference between \~30 m and \~80 m grid cells
+isn't that big of a deal. If that doesn't convince you, though, we do some
+limited sensitivity analysis down below.
+
 ```{r}
@@
 
-add_graph <- function(x, max_slope = NULL, ...){
-
-  g <- terra::adjacent(
-    x$dem,
-    cells = terra::cells(x$dem),
+aggregate_grid <- function(x, .factor){
+
+  if (.factor > 1){
+
+    x[["dem"]] <- terra::aggregate(
+      x[["dem"]], 
+      fact = .factor,
+      fun = "mean",
+      na.rm = TRUE
+    )
+
+  }
+
+  attr(x, "agg_fact") <- .factor
+
+  x
+
+}
+
+group_sites_by_cell <- function(x){
+
+  # get raster cell IDs for site points
+  x[["sites"]] <- x[["sites"]] |> 
+    mutate(cell = cells(x[["dem"]], vect(x[["sites"]]))[,2])
+
+  # group relevant site attribute data by grid cell
+  x[["cell_table"]] <- x[["sites"]] |> 
+    st_drop_geometry() |> 
+    select(cell, center, n_room) |> 
+    group_by(cell) |> 
+    summarize(
+      center = ifelse(sum(center) > 0, 1, 0),
+      n_room = sum(n_room)
+    )
+
+  x
+
+}
+
+add_graph <- function(x, max_slope = NULL){
+
+  # keep track of metadata
+  attr(x, "max_slope") <- max_slope
+
+  # cells(<SpatRaster>) returns cell IDs 
+  # for all cells with non-NA values
+  non_NA_cells <- cells(x[["dem"]])
+
+  # cells(<SpatRaster>, <numeric>) returns cell IDs 
+  # for all cells with values matching <numeric>
+  NA_cells <- cells(x[["dem"]], NA_real_)[[1]]
+
+  # get adjacency matrix 
+  graph <- adjacent(
+    x[["dem"]],
+    cells = non_NA_cells,
     directions = 8,
     pairs = TRUE
-  )
-
-  g <- as.data.frame(g)
-
-  g <- g[!g$to %in% terra::cells(x$dem, NA_real_)[[1]], ]
-
-  # distance between cells
-  g$run <- terra::distance(
-    terra::xyFromCell(x$dem, as.vector(g$from)),
-    terra::xyFromCell(x$dem, as.vector(g$to)),
-    lonlat = terra::is.lonlat(x$dem),
+  ) |> as.data.frame()
+
+  # remove rows with to-cells that have NA values
+  graph <- graph |> filter(!to %in% NA_cells)
+
+  # rise = difference between adjacent cells
+  # run = distance between adjacent cells
+  # slope = rise/run
+  # convert slope 
+  #   to radians with rads = atan(rise/run)
+  #   to degrees with degs = rads * 180/pi 
+  graph <- graph |> 
+    mutate(
+      from_elevation = terra::extract(x[["dem"]], from)[,1],
+      to_elevation = terra::extract(x[["dem"]], to)[,1],
+      rise = from_elevation - to_elevation,
+      run = distance(
+        xyFromCell(x[["dem"]], from),
+        xyFromCell(x[["dem"]], to),
+        lonlat = is.lonlat(x[["dem"]]),
         pairwise = TRUE
-  )
-
-  # difference between cells
-  g$from_elevation <- terra::extract(x$dem, as.vector(g$from))[,1]
-
-  g$to_elevation <- terra::extract(x$dem, as.vector(g$to))[,1]
-
-  g$rise <- g$from_elevation - g$to_elevation
-
-  # slope
-  g$slope <- g$rise/g$run
+      ),
+      slope = atan(rise/run) * (180/pi)
+    )
 
   # what is a realistic slope people would try to hike?
-  if (!is.null(max_slope)) {
-
-    g <- g[g$slope <= tan(max_slope * pi / 180), ]
-
-  }
-
-  g$speed <- campbell(g$slope, ...)
-
-  g$cost <- g$run/g$speed
-
-  g <- g[, c("from", "to", "cost")]
-
-  coords <- cbind(
-    ID = terra::cells(x$dem), 
-    terra::xyFromCell(x$dem, terra::cells(x$dem))
-  ) |> 
-    as.data.frame() |> 
+  # remove all edges with slopes greater than max_slope
+  if (!is.null(max_slope)) { graph <- graph |> filter(slope <= max_slope) }
+
+  # now that we have slope, the strategy is this:
+  # 1) calculate the hiking speed on each slope estimate
+  # 2) calculate the length of each slope (the length of the hypotenuse)
+  # 3) divide length by speed to get travel time (this is the edge weight)
+
+  # Campbell's hiking function, see Campbell et al 2019
+  # note: campbell function wants degrees!
+  campbell <- function(x) {
+
+    # values provided in Supplement Table S3
+    # using the 5th percentile of the Lorentz curve as recommended by Campbell
+    a <- -1.527
+    b <- 14.041
+    c <- 36.813
+    d <- 0.320
+    e <- -0.00273
+
+    # lorentz distribution
+    lorentz <- (1 / ((pi * b) * (1 + ((x - a) / b)^2)))
+
+    # modified lorentz
+    (c * lorentz) + d + (e * x)
+
+  }
+
+  # calculate 
+  #   speed using Campbell's hiking function
+  #   length of hypotenuse
+  #   travel time between adjacent cells
+  graph <- graph |> 
+    mutate(
+      speed = campbell(slope),
+      cost = sqrt(run^2L + rise^2L)/speed
+    )
+
+  # get coordinates table with cell IDs for safe joins
+  coords <- data.frame(ID = non_NA_cells) |> 
+    bind_cols(xyFromCell(x[["dem"]], non_NA_cells)) |> 
     rename_with(toupper)
 
-  # force as.character(1000000) to be "1000000" instead of "1e-6"
-  # this fixes problem with makegraph()
-  options(scipen = 9999)
-
-  x$graph <- makegraph(
-    as.data.frame(g),
+  x[["graph"]] <- makegraph(
+    graph |> select(from, to, cost),
     directed = TRUE,
@@
 
-  return(x)
-
-}
-
-campbell <- function(x, decile = 30L) {
-
-  if (length(decile) != 1) {
-
-    stop("campbell accepts only one decile at a time.")
-
-  }
-
-  # campbell function wants degrees, not rise-over-run as in tobler
-  # but, we're working with rr, so remember to convert radians to degrees with 180/pi!
-  slope <- atan(x) * 180 / pi
-
-  # values provided in Campbell 2019, Supplement, Table S1
-  # for simplicity, I excluded all but the deciles
-  deciles <-
-    data.frame(
-      decile = c(10L, 20L, 30L, 40L, 50L, 60L, 70L, 80L, 90L),
-      a = c(-1.568, -1.71,  -1.858, -1.958, -2.171, -2.459, -2.823,  -3.371,  -3.06),
-      b = c(13.328, 10.154,  8.412,  8.96,  10.064, 11.311, 12.784,  15.395,  16.653),
-      c = c(38.892, 36.905, 39.994, 50.34,  63.66,  79.287, 98.697, 134.409, 138.875),
-      d = c( 0.404,  0.557,  0.645,  0.649,  0.628,  0.599,  0.566,   0.443,   0.823),
-      e = c(-0.003, -0.004, -0.004, -0.005, -0.005, -0.005, -0.005,  -0.005,  -0.0139)
-    )
-
-  ind <- which(deciles$decile == as.integer(decile))
-
-  deciles <- deciles[ind, ]
-
-  a <- deciles$a
-  b <- deciles$b
-  c <- deciles$c
-  d <- deciles$d
-  e <- deciles$e
-
-  # lorentz distribution
-  lorentz <- (1 / ((pi * b) * (1 + ((slope - a) / b)^2)))
-
-  # modified lorentz
-  (c * lorentz) + d + (e * slope)
-
-}
-
-```
-
-```{r}
-
-cmv <- cmv |> add_graph(max_slope = 45, decile = 10L)
-nrg <- nrg |> add_graph(max_slope = 45, decile = 10L)
-cib <- cib |> add_graph(max_slope = 45, decile = 10L)
+  x
+
+}
+
+```
+
+```{r}
+#| time_it: true
+
+cmv <- cmv |> 
+  aggregate_grid(.factor = 3) |> 
+  group_sites_by_cell() |> 
+  add_graph(max_slope = 45)
+
+nrg <- nrg |> 
+  aggregate_grid(.factor = 3) |> 
+  group_sites_by_cell() |> 
+  add_graph(max_slope = 45)
+
+cib <- cib |> 
+  aggregate_grid(.factor = 3) |> 
+  group_sites_by_cell() |> 
+  add_graph(max_slope = 45)
 
@@
 
+To calculate travel time between every pair of site points, we use a
+bi-directional version of Dijkstra's algorithm as implemented in the R package
+`{cppRouting}`. Note that we also average the from- and to- travel times, so the
+results are isotropic (meaning direction is irrelevant).
+
 ```{r}
@@
 
-  x$sites$cell <- terra::cells(x$dem, terra::vect(x$sites))[,2]
-
-  # use unique cell IDs as some centers are in the same pixel, 
-  # so calculations would be redundant
+  cell_ids <- x[["cell_table"]][["cell"]]
+
+  cells_in_graph <- x[["graph"]][["dict"]][["ref"]]
+
+  missing_cells <- setdiff(cell_ids, cells_in_graph)
+
+  if (length(missing_cells) > 0){
+
+    cli::cli_abort(c(
+      "Some grid cells have sites but are not in the graph!",
+      "i" = "Cells: {missing_cells}"
+    ))
+
+  }
+
   dm <- get_distance_matrix(
-    x$graph,
-    from = x$sites$cell |> unique(),  
-    to = x$sites$cell |> unique()
+    x[["graph"]],
+    from = cell_ids,  
+    to = cell_ids
   )
@@
   # make symmetric by averaging off-diagonals
-  d <- cbind(dm[lower.tri(dm)], t(dm)[lower.tri(dm)])
-
-  dm[lower.tri(dm)] <- rowMeans(d, na.rm = TRUE)
-  dm[upper.tri(dm)] <- t(dm)[upper.tri(dm)]
+  lt <- lower.tri(dm)
+  ut <- upper.tri(dm)
+
+  dm[lt] <- rowMeans(cbind(dm[lt], t(dm)[lt]), na.rm = TRUE)
+  dm[ut] <- t(dm)[ut]
 
@@
 
-  x$dm <- dm
-
-  return(x)
-
-}
-
-```
-
-```{r}
+  # add to list and return
+  x[["dm"]] <- dm
+
+  x
+
+}
+
+```
+
+```{r}
+#| time_it: true
 
@@
 
-get_commutes <- function(x, n){
-
-  centers <- get_centers(x) |> select(-center)
-  farms   <- get_farms(x) |> select(-center)
+randomize_commutes <- function(x, n = 1000){
 
   # subset dm (distance matrix) to get dm[farms, centers]
-  i <- which(rownames(x$dm) %in% farms$cell)
-  j <- which(colnames(x$dm) %in% centers$cell)
-
-  dm <- x$dm[i, j]
-
+  farms <- x[["cell_table"]] |> filter(center == 0)
+  centers <- x[["cell_table"]] |> filter(center == 1)
+
+  i <- which(rownames(x[["dm"]]) %in% farms[["cell"]])
+  j <- which(colnames(x[["dm"]]) %in% centers[["cell"]])
+
+  dm <- x[["dm"]][i, j]
+
+  # get observed density
   observed <- apply(dm, 1, min)
-  observed <- density(observed, kernel = "gaussian", n = 512, from = 0, to = 7200)
-
-  r <- terra::rast(
-    nrows = nrow(x$dem), 
-    ncols = ncol(x$dem), 
-    extent = ext(x$dem), 
-    crs = crs(x$dem)
-  )
-
+  observed <- density(
+    observed,
+    kernel = "gaussian", 
+    n = 512, 
+    from = 0, 
+    to = 7200
+  )
+
+  # build raster to randomly sample cells
+  r <- setValues(x[["dem"]], NA)
+
+  # don't sample past furthest observed distance from nearest center
   dmax <- st_distance(
-    x$sites |> filter(center == 0), 
-    x$sites |> filter(center == 1)
-  ) |> units::drop_units() |> apply(1, min) |> max()
-
-  i <- cells(
-    x$dem, 
-    x$sites |> filter(center == 1) |> st_buffer(dmax) |> st_union() |> vect()
-  )[, "cell"]
-
-  i <- i[i %in% x$graph$dict$ref]
+    x[["sites"]] |> filter(center == 0), 
+    x[["sites"]] |> filter(center == 1)
+  )
+
+  dmax <- units::drop_units(dmax)
+  dmax <- apply(dmax, 1, min)
+  dmax <- max(dmax)
+
+  sample_area <- x[["sites"]] |> 
+    filter(center == 1) |> 
+    st_buffer(dist = dmax) |> 
+    st_union() |> 
+    vect()
+
+  i <- cells(x[["dem"]], sample_area)[, "cell"]
+
+  # only use those that have nodes in the graph (excludes big slopes)
+  i <- i[i %in% x[["graph"]][["dict"]][["ref"]]]
 
@@
   # don't sample cells with centers in them
-  j <- colnames(dm) |> as.numeric()
+  j <- unique(centers[["cell"]])
 
@@
 
-  random_points <- terra::spatSample(
+  random_points <- spatSample(
     r, 
@@
     na.rm = TRUE
-  ) |> pull(cell) |> unique()
+  )
+
+  random_points <- unique(random_points[["cell"]])
 
   dm <- get_distance_matrix(
-    x$graph,
+    x[["graph"]],
     from = random_points,  
@@
   expected <- apply(dm, 1, min)
-  expected <- density(expected, kernel = "gaussian", n = 512, from = 0, to = 7200)
-
-  tibble(
-    region = x$name,
-    distance = observed$x,
-    observed = observed$y,
-    expected = expected$y
+  expected <- density(
+    expected,
+    kernel = "gaussian", 
+    n = 512, 
+    from = 0, 
+    to = 7200
+  )
+
+  tibble::tibble(
+    region = attr(x, "name"),
+    distance = observed[["x"]],
+    observed = observed[["y"]],
+    expected = expected[["y"]]
   )
@@
 
-commutes <- bind_rows(
-  cmv |> get_commutes(n = 1000),
-  nrg |> get_commutes(n = 1000),
-  cib |> get_commutes(n = 1000)
-)
+commutes <- list(cmv, nrg, cib) |> 
+  map(randomize_commutes) |> 
+  bind_rows()
 
@@
 
-  centers <- get_centers(x) |> select(-center)
-  farms   <- get_farms(x) |> select(-center)
-
   # subset dm (distance matrix) to get dm[farms, centers]
-  i <- which(rownames(x$dm) %in% farms$cell)
-  j <- which(colnames(x$dm) %in% centers$cell)
-
-  dm <- x$dm[i, j]
+  farms <- x[["cell_table"]] |> filter(center == 0)
+  centers <- x[["cell_table"]] |> filter(center == 1)
+
+  i <- which(rownames(x[["dm"]]) %in% farms[["cell"]])
+  j <- which(colnames(x[["dm"]]) %in% centers[["cell"]])
+
+  dm <- x[["dm"]][i, j]
 
@@
 
-  tibble(
-    region = x$name,
-    distance = quantile(observed, p = p) |> unname()
-  )
-
-}
-
-q95 <- bind_rows(
-  q(cmv, 0.95),
-  q(nrg, 0.95),
-  q(cib, 0.95)
-) |> 
+  quant <- quantile(observed, p = p)
+  quant <- unname(quant)
+
+  tibble::tibble(
+    region = attr(x, "name"),
+    distance = quant
+  )
+
+}
+
+q95 <- list(cmv, nrg, cib) |> 
+  map(\(x){ q(x, 0.95) }) |> 
+  bind_rows() |> 
   mutate(
@@
     nudge_x = 2,
-    size = 12/.pt
+    size = 11.5/.pt
   ) +
@@
     "point",
-    x = min(q95$distance)/60,
+    x = min(q95[["distance"]])/60,
     y = 3.0,
@@
     "text",
-    label = "95% Quantile of Observed",
-    x = min(q95$distance)/60 + 2,
+    label = "Observed 95% Quantile",
+    x = min(q95[["distance"]])/60 + 2,
     y = 3.0,
@@
     hjust = 0,
-    size = 12/.pt
+    size = 11.5/.pt
   ) +
@@
   ) +
-  scale_x_continuous(breaks = seq(0, 120, by = 30), expand = expansion(0.03)) +
+  scale_x_continuous(
+    breaks = seq(0, 120, by = 30), 
+    expand = expansion(0.03)
+  ) +
   coord_cartesian(xlim = c(0, 120)) +
@@
   filename = fn,
-  width = 5,
+  width = 5.1,
   height = 3.3,
@@
 
-remove(get_commutes, q, q95, gg)
+remove(randomize_commutes, commutes, q, q95, gg)
 
@@
 ::: {#fig-commutes}
-![](../figures/commute-kl.png){width="60%" fig-alt="Difference between log probability density of obversed commute times from farms to nearest centers and log probability density of commute times from random locations to nearest centers."}
-
-Difference between log probability density of obversed commute times from farms to nearest centers and log probability density of commute times from random locations to nearest centers.
+![](../figures/commute-kl.png){width="60%"
+fig-alt="Difference between log probability density of obversed commute times from farms to nearest centers and log probability density of commute times from random locations to nearest centers."}
+
+Difference between log probability density of observed commute times from farms
+to nearest centers and log probability density of commute times from random
+locations to nearest centers.
 :::
@@
 1.  Identify community centers
-2. Identify farms within commute distance 
-3. Join communities to their overlapping neighbors 
-4. Exclude farms closer to another community center (tie-breaking step) 
+2.  Remove distant farms
+3.  Join centers to their overlapping neighbors
+4.  Join farms to their nearest center
 5.  Draw smallest concave hull encompassing all farms, centers, and paths
 
-Step 1 is already done and stored in the data.
-
-### Filter by commute distance
+Step 1 is already done and stored in the data. Step 2 can technically be done
+after 2 and 3, but it makes sense to do it first, since it means those sites can
+be ignored afterwards.
+
+### Filter out distant farms
 
@@
 
-subset_distance_matrix <- function(x){
-
-  centers <- get_centers(x) |> select(-center)
-  farms   <- get_farms(x) |> select(-center)
+remove_distant_farms <- function(x, dmax){
 
   # subset dm (distance matrix) to get dm[farms, centers]
-  i <- which(rownames(x$dm) %in% farms$cell)
-  j <- which(colnames(x$dm) %in% centers$cell)
-
-  x$dm[i, j]
-
-}
-
-filter_distance_matrix <- function(x, d){
-
-  nearest_farms <- apply(x, 1, \(x){ which(x <= d) })
-  nearest_farms <- lapply(nearest_farms, \(x){ as.numeric(names(x)) })
-
-  arrow_table(
-    center = unlist(nearest_farms),
-    farm = rep(names(nearest_farms), lengths(nearest_farms)) |> as.numeric()
-  ) |> arrange(center, farm)
-
-}
-
-find_nearby_farms <- function(x, dmax){
-
-  x$dmax <- dmax
-
-  x$nearest_farms <- x |> 
-    subset_distance_matrix() |> 
-    filter_distance_matrix(dmax)
-
-  sites <- x$sites |> 
-    st_drop_geometry() |> 
-    select(cell, n_room) |> 
-    group_by(cell) |> 
-    summarize(n_room = sum(n_room))
-
-  x$nearest_farms <- x$nearest_farms |> left_join(sites, by = join_by(farm == cell))
-
-  return(x)
-
-}
-
-```
-
-```{r}
+  farms <- x[["cell_table"]] |> filter(center == 0)
+  centers <- x[["cell_table"]] |> filter(center == 1)
+
+  i <- which(rownames(x[["dm"]]) %in% farms[["cell"]])
+  j <- which(colnames(x[["dm"]]) %in% centers[["cell"]])
+
+  dm <- x[["dm"]][i, j]
+
+  # if the distance from a farm to its nearest center
+  # is greater than dmax, we remove it from the cell table
+  min_distance <- apply(dm, 1, min)
+
+  dm <- dm[min_distance <= dmax, ]
+
+  x[["cell_table"]] <- x[["cell_table"]] |> 
+    filter(cell %in% c(rownames(dm), colnames(dm)))
+
+  # collect metadata
+  attr(x, "dmax") <- dmax
+  attr(x, "n_outliers") <- nrow(farms) - nrow(dm)
+
+  x
+
+}
+
+```
+
+```{r}
+#| time_it: true
 
@@
 
-cmv <- cmv |> find_nearby_farms(dmax = one_hour)
-nrg <- nrg |> find_nearby_farms(dmax = one_hour)
-cib <- cib |> find_nearby_farms(dmax = one_hour)
-
-```
-
-### Find neighboring centers
-
-> For any two centers $C_1$ and $C_2$ with nearby farm room counts $N_1 < N_2$, if $pN_1$ of $C_1$’s rooms are within a distance $d$ of $C_1$ and $C_2$, then $C_1$ is part of the same community as $C_2$.
+cmv <- cmv |> remove_distant_farms(dmax = one_hour)
+nrg <- nrg |> remove_distant_farms(dmax = one_hour)
+cib <- cib |> remove_distant_farms(dmax = one_hour)
+
+```
+
+### Join neighboring centers
+
+> For any two centers $C_1$ and $C_2$ with catchment room counts $N_1 < N_2$, if
+> $pN_1$ of $C_1$'s rooms are within a distance $D$ of $C_2$, then $C_1$ is part
+> of the same community as $C_2$.
+
+Here the population, $N$, of focal center $c$ is defined as
+
+$$N_{c} = R_{c} + \sum_{k=1}^{S} R_{k} \cdot I(t_{kc} \leq D_{c})$$
+
+for all sites $S$ (this includes both farms and centers), with $R_{c}$ being the
+room count at $c$ and $R_{k}$ the room count at site $k$, $t_{kc}$ the travel
+time from $k$ to $c$, $D_{c}$ the threshold travel time to $c$ (referred to as
+*D-join* in the paper), and $I$ the indicator function that is 1 if $t \leq D$
+and 0 otherwise.
 
@@
 
-  x$djoin <- djoin
-
-  nearest_farms <- x |> 
-    subset_distance_matrix() |> 
-    filter_distance_matrix(djoin)
-
-  sites <- x$sites |> 
-    st_drop_geometry() |> 
-    select(cell, n_room) |> 
-    group_by(cell) |> 
-    summarize(n_room = sum(n_room))
-
-  nearest_farms <- nearest_farms |> left_join(sites, by = join_by(farm == cell))
-
-  remove(sites)
-
-  room_counts <- nearest_farms |> 
+  # subset dm (distance matrix) to get dm[farms and centers, centers]
+  center_cells <- x[["cell_table"]] |> filter(center == 1) |> pull(cell)
+
+  j <- which(colnames(x[["dm"]]) %in% center_cells)
+
+  dm <- x[["dm"]][, j]
+
+  # find farms and centers that are within distance djoin of other centers
+  djoins <- apply(dm, 1, \(.x){ which(.x <= djoin) })
+
+  # remove farms that are isolated (ie, greater than djoin from a center)
+  # note: this should just be farms because all centers are less than
+  #   djoin from themselves
+  djoins <- djoins[lengths(djoins) > 0]
+
+  # collect list into a long-form table
+  djoins <- tibble(
+    center = colnames(dm)[unlist(djoins)],
+    farm_or_center = rep(names(djoins), lengths(djoins))
+  ) |> mutate(across(everything(), as.numeric))
+
+  # now we combine the room counts with the djoins table
+  djoins <- djoins |> 
+    left_join(
+      x[["cell_table"]] |> select(-center), 
+      by = join_by(farm_or_center == cell)
+    ) |> 
+    rename("population" = n_room)
+
+  # calculate the total population of the catchment around each center
+  # note that this involves a lot of double-counting, farms and centers
+  # counting towards the population size of multiple centers, but this is OK
+  # as we are interested in using the shared population as a measure of 
+  # interaction between centers, not as an estimate of a community's population
+  gen_pop <- djoins |> 
     group_by(center) |> 
-    summarize(n_room = sum(n_room)) |> 
-    collect()
-
-  R <- outer(room_counts$n_room, room_counts$n_room, FUN = '<=')
+    summarize(population = sum(population))
+
+  # for each center, what other centers have larger catchment populations?
+  # using outer() is extremely fast
+  # in this case, it creates an asymmetric logical matrix
+  R <- outer(gen_pop[["population"]], gen_pop[["population"]], FUN = '<=')
   diag(R) <- FALSE
 
-  colnames(R) <- room_counts$center
-  rownames(R) <- room_counts$center
-
+  colnames(R) <- gen_pop[["center"]]
+  rownames(R) <- gen_pop[["center"]]
+
+  # now, we turn that matrix into a big list 
+  # each element is a vector with the cell IDs of the larger centers
   R <- apply(R, 1, which)
-  R <- lapply(R, \(x){ as.numeric(names(x)) })
-
-  neighbors_matrix <- imap(R, \(x, idx){
-
-    N <- room_counts |> filter(center == as.numeric(idx)) |> pull(n_room)
-
-    focal_farms <- nearest_farms |> 
+  R <- map(R, \(x){ as.numeric(names(x)) })
+
+  # for each center, we look at all the centers with bigger catchment
+  # populations than it, and ask: is the shared population size some fraction
+  # of the total population of the catchment?
+  neighbors_matrix <- purrr::imap(R, \(x, idx){
+
+    N <- gen_pop |> 
       filter(center == as.numeric(idx)) |> 
-      collect() |> 
-      pull(farm)
-
-    cntrs <- nearest_farms |> 
+      pull(population)
+
+    focal_neighbors <- djoins |> 
+      filter(center == as.numeric(idx)) |> 
+      pull(farm_or_center)
+
+    cntrs <- djoins |> 
       filter( # this filter gives the intersection of farms/cells 
         center %in% x,
-        farm %in% focal_farms
+        farm_or_center %in% c(idx, focal_neighbors)
       ) |> 
       group_by(center) |> 
-      summarize(proportion = sum(n_room)/N) |> 
+      summarize(proportion = sum(population)/N) |> 
       filter(proportion >= p) |> 
-      collect() |> 
       pull(center)
 
+    # return a logical vector the length of R
     names(R) %in% cntrs
@@
 
-  # igraph...
+  # there's a recursive dimension to this:
+  # if a center is joined with a center that's joined with a center, etc.
+  # i don't know how to do that in R
+  # so, igraph it is...
   g <- igraph::graph.adjacency(neighbors_matrix)
-  membership <- igraph::components(g)$membership
-
-  x$lookup <- tibble(
+  membership <- igraph::components(g)[["membership"]]
+
+  # this gives us the community that every center is part of
+  x[["center_communities"]] <- tibble(
     center = names(membership) |> as.numeric(),
@@
 
-  return(x)
-
-}
-
-```
-
-```{r}
+  # collect metadata
+  attr(x, "djoin") <- djoin
+  attr(x, "n_communities") <- length(unique(unname(membership)))
+  attr(x, "proportion") <- p
+
+  x
+
+}
+
+```
+
+```{r}
+#| time_it: true
 
@@
 
-### Find nearest center
+### Collect members
+
+Here we just associate each farm with its nearest community center. So, we have
+full community membership defined. Then we drop communities that have three or
+less sites associated with them. This is only done because our main goal with
+this analysis is to estimate the extent of farm land associated with these
+dispersed farming communities.
 
@@
 
-find_nearest_center <- function(x){
-
-  farms_to_centers <- subset_distance_matrix(x)
-
-  # find the nearest center for each farm
-  nearest_center <- apply(farms_to_centers, 1, \(x) which.min(x))
-  nearest_center <- colnames(farms_to_centers)[nearest_center]
-
-  distance <- apply(farms_to_centers, 1, min)
-
-  x$nearest_center <- tibble(
-    farm = farms_to_centers |> rownames() |> as.numeric(),
-    center = nearest_center |> as.numeric(),
-    distance = unname(distance)
-  ) |> filter(distance <= x$dmax)
-
-  return(x)
-
-}
-
-```
-
-```{r}
-
-cmv <- cmv |> find_nearest_center() 
-nrg <- nrg |> find_nearest_center() 
-cib <- cib |> find_nearest_center()
-
-```
-
-### Define boundaries
-
-Now that we have each farm associated with its nearest community center, and each center joined with its neighbors, assuming it has any, we can define a boundary for the community using a concave hull.
-We choose a concave rather than convex hull because the latter tends to artificially inflate the size of the community, often to a significant degree.
-On the other hand, the concave hull can generate extremely unrealistic shapes for communities.
-To avoid these issues, we first compute the least-cost path between each farm in the community and every other farm in the community.
-We then add the vertices that define these paths as virtual points to the community.
-We then generate the concave hull for the expanded set of community points.
-The consequence is that an individual setting off from one farm in the community to another should never have to leave the community.
-Their walk will always be within the community boundary.
-
-The result is still not perfect, however, so we further simplify polygons (including filling holes) by adding then removing a small buffer.
-Two things to note here: (1) we project the boundaries first as Google's `S2` system for geocomputations on the sphere (the default now in `{sf}`) is janky with how it does buffering and (2) we add a small buffer to smooth some of the noise in the hull.
-
-```{r}
-#| code-fold: true
-
 collect_members <- function(x){
 
-  lookup <- x$lookup
-
-  centers <- get_centers(x) |> 
-    left_join(lookup, by = join_by(cell == center)) |> 
-    select(site_id, n_room, initial_phase, final_phase, community) |> 
-    filter(!is.na(community)) |> 
-    mutate(type = "center")
-
-  farms <- get_farms(x) |> 
-    select(-center) |> 
-    left_join(x$nearest_center, by = join_by(cell == farm)) |> 
-    left_join(lookup, by = "center") |> 
-    select(site_id, n_room, initial_phase, final_phase, community) |> 
-    filter(!is.na(community)) |> 
-    mutate(type = "farm")
-
-  x$members <- bind_rows(farms, centers) |> 
-    arrange(community, site_id) |> 
-    left_join(x$sites[, "site_id"], by = "site_id") |> 
-    mutate(region = x$name, .before = everything()) |> 
-    st_sf()
-
-  return(x)
-
-}
-
-define_boundaries <- function(x, concavity_ratio, dilate){
-
-  community_ids <- x$members |> pull(community) |> unique()
-
-  communities <- vector("list", length(community_ids))
-
-  # cli::cli_progress_bar(
-  #   "Define boundaries", 
-  #   total = length(community_ids)
-  # )
-
-  for (i in community_ids){
-
-    m <- x$members |> filter(community == i)
-
-    cells <- terra::cells(x$dem, vect(m))[, 2] |> unique()
-
-    paths <- get_multi_paths(
-      x$graph, 
-      from = cells, 
-      to = cells
-    )
-
-    paths <- paths |> unlist(recursive = TRUE) |> as.numeric() |> unique()
-
-    xy <- rbind(
-      terra::xyFromCell(x$dem, paths),
-      st_coordinates(m)[, 1:2]
-    )
-
-    xy <- xy |> st_multipoint() |> st_sfc(crs = st_crs(m)) |> st_cast("POINT")
-
-    m <- st_geometry(m)
-
-    border <- st_sf(geometry = c(xy, m)) |> 
-      summarize() |> 
-      st_concave_hull(ratio = concavity_ratio) |> 
-      mutate(community = i)
-
-    communities[[i]] <- border
-
-    remove(i, m, cells, paths, xy, border)
-
-    # cli::cli_progress_update()
-
-  }
-
-  x$communities <- communities |> 
-    bind_rows() |> 
-    st_transform(x$preferred_projection) |> 
-    st_buffer(dilate[1]) |> 
-    st_buffer(dilate[2]) |> 
-    st_transform(4326) |> 
-    mutate(region = x$name, .before = "community")
-
-  return(x)
-
-}
-
-```
-
-```{r}
-
-cmv <- cmv |> 
-  collect_members() |> 
-  define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
-
-nrg <- nrg |> 
-  collect_members() |> 
-  define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
-
-cib <- cib |> 
-  collect_members() |> 
-  define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
-
-```
-
-### Drop small communities
-
-```{r}
-#| code-fold: true
+  # subset dm (distance matrix) to get dm[farms, centers]
+  farms <- x[["cell_table"]] |> filter(center == 0)
+  centers <- x[["cell_table"]] |> filter(center == 1)
+
+  i <- which(rownames(x[["dm"]]) %in% farms[["cell"]])
+  j <- which(colnames(x[["dm"]]) %in% centers[["cell"]])
+
+  dm <- x[["dm"]][i, j]
+
+  # get nearest center to each farm
+  k <- apply(dm, 1, which.min)
+  nearest_centers <- tibble(
+    farm = as.numeric(names(k)),
+    nearest_center = as.numeric(colnames(dm)[k])
+  )
+
+  # collect the farms and the centers they are closest to, 
+  # and by extension the communities they are part of
+  farms <- x[["cell_table"]] |> 
+    filter(center == 0) |> 
+    left_join(
+      nearest_centers, 
+      by = join_by(cell == farm)
+    ) |> 
+    left_join(
+      x[["center_communities"]], 
+      by = join_by(nearest_center == center)
+    ) |> 
+    select(-nearest_center)
+
+  # collect the centers and their community affiliations
+  centers <- x[["cell_table"]] |> 
+    filter(center == 1) |> 
+    left_join(
+      x[["center_communities"]], 
+      by = join_by(cell == center)
+    )
+
+  # combine and save
+  x[["cell_table"]] <- bind_rows(farms, centers)
+
+  x
+
+}
 
@@
 
-  final_list <- x$members |> 
-    st_drop_geometry() |> 
-    filter(type == "farm") |> 
+  final_list <- x[["cell_table"]] |> 
+    filter(center == 0) |> 
     group_by(community) |> 
@@
     filter(n > threshold) |> 
-    pull(community)
-
-  x$communities <- x$communities |> filter(community %in% final_list)
-
-  x$members <- x$members |> st_filter(x$communities)
-
-  return(x)
-
-}
-
-```
-
-```{r}
-
-cmv <- cmv |> drop_small_communities(threshold = 3)
-nrg <- nrg |> drop_small_communities(threshold = 3)
-cib <- cib |> drop_small_communities(threshold = 3)
+    pull(community) |> 
+    unique()
+
+  x[["cell_table"]] <- x[["cell_table"]] |> filter(community %in% final_list)
+
+  # update metadata
+  attr(x, "min_size") <- threshold
+  attr(x, "n_dropped") <- attr(x, "n_communities") - length(final_list)
+
+  x
+
+}
+
+```
+
+```{r}
+#| time_it: true
+
+cmv <- cmv |> 
+  collect_members() |> 
+  drop_small_communities(threshold = 3)
+
+nrg <- nrg |> 
+  collect_members() |> 
+  drop_small_communities(threshold = 3)
+
+cib <- cib |> 
+  collect_members() |> 
+  drop_small_communities(threshold = 3)
+
+```
+
+### Define boundaries
+
+Now that we have community centers joined to their overlapping neighbors, and
+farms joined to their nearest centers, we can define boundaries for the
+resulting communities using a concave hull. We choose a concave rather than
+convex hull because the latter tends to artificially inflate the size of the
+community, often to a significant degree. On the other hand, the concave hull
+can generate extremely unrealistic shapes for communities, even when moving the
+concavity ratio closer to the convex hull.
+
+Our somewhat brute force strategy for navigating between the Scylla and
+Charybdis of being too big or too strange is to first compute the least-cost
+path between every site on the concave hull of the community's site points. In
+this case the concave hull gives us all the points on the convex hull along with
+some on the interior of the convex hull. The points that are excluded should
+have shortest paths to the outer points that at some point intersect with the
+shortest paths that are actually calculated, so we get a reasonable
+approximation and save some time doing it. We then add the vertices that define
+the computed paths as virtual points to the community and generate the concave
+hull for the expanded set of community points. The consequence is that an
+individual setting off from one farm in the community to any other site in the
+community (farm or center) should never have to leave the community. This, in
+effect, makes the community polygons concave with respect to time.
+
+The result can still be a little noisy, so we simplify polygons - including, in
+some case, filling holes - by dilating them, that is, adding then removing a
+small buffer.
+
+Note that we also do some graph pruning in each iteration, basically cropping
+the graph to an expanded area around the points assigned to a specific
+community. This prevents the search algorithm from searching out away from the
+other sites in the community.
+
+```{r}
+#| code-fold: true
+
+define_boundaries <- function(x, concavity_ratio, dilate){
+
+  members <- x[["cell_table"]]
+
+  sites <- x[["sites"]] |> 
+    select(cell) |> 
+    st_transform(attr(x, "grid_crs"))
+
+  community_ids <- unique(members[["community"]])
+
+  communities <- vector("list", length(community_ids))
+
+  for (i in community_ids){
+
+    m <- members |> filter(community == i)
+
+    # get cells along the edge of point clusters using concave hull
+    # so we get points that are on the convex hull and in
+    # the interior of the cluster, too
+    pts <- xyFromCell(x[["dem"]], m[["cell"]]) |> 
+      st_multipoint() |> 
+      st_sfc(crs = attr(x, "grid_crs")) |> 
+      st_sf(geometry = _)
+
+    cells <- pts |> 
+      st_cast("POINT") |> 
+      mutate(cell = m[["cell"]]) |> 
+      st_filter(st_concave_hull(pts, 0.2), .predicate = st_touches) |> 
+      pull(cell)
+
+    # prune the graph so search doesn't go too far away from the convex hull,
+    # meaning away from all the other points in the community
+    # we use a somewhat arbitrary 1.5 km distance for this, so we can reasonably 
+    # assume we are not chopping off the shortest path, while at the same time
+    # cutting down the total number of paths to search by a considerable amount
+    bb8 <- st_convex_hull(pts) |> 
+      st_buffer(1500) |> 
+      st_bbox() |> 
+      st_as_sfc(crs = st_crs(pts))
+
+    cells_in_bb8 <- cells(x[["dem"]], vect(bb8))[,2]
+
+    graph <- x[["graph"]]
+
+    node_ids <- graph[["dict"]] |> 
+      filter(ref %in% cells_in_bb8) |> 
+      pull(id)
+
+    graph[["data"]] <- graph[["data"]] |> 
+      filter(from %in% node_ids, to %in% node_ids)
+
+    # get all short paths 
+    # this returns a list with length equal to the length of cells
+    # each element of the list is a vector of the grid cells that the short
+    # path passes through
+    paths <- get_multi_paths(graph, from = cells, to = cells)
+
+    # unraveling this list gives us a vector of all the cells on shortest paths
+    paths <- paths |> unlist(recursive = TRUE) |> as.numeric() |> unique()
+
+    site_xy <- sites |> 
+      filter(cell %in% m[["cell"]]) |> 
+      st_coordinates()
+
+    names(site_xy) <- tolower(names(site_xy))
+
+    xy <- rbind(
+      xyFromCell(x[["dem"]], paths),
+      site_xy
+    )
+
+    border <- xy |> 
+      st_multipoint() |> 
+      st_sfc(crs = attr(x, "grid_crs")) |> 
+      st_sf(geometry = _) |> 
+      st_concave_hull(ratio = concavity_ratio) |> 
+      mutate(community = i)
+
+    communities[[i]] <- border
+
+  }
+
+  x[["communities"]] <- communities |> 
+    bind_rows() |> 
+    st_transform(attr(x, "preferred_projection")) |> 
+    st_buffer(dilate[1]) |> 
+    st_buffer(dilate[2]) |> 
+    st_transform(4326)
+
+  # update metadata
+  attr(x, "concavity") <- concavity_ratio
+  attr(x, "dilate") <- dilate
+
+  x
+
+}
+
+```
+
+```{r}
+#| time_it: true
+
+cmv <- cmv |> define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
+nrg <- nrg |> define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
+cib <- cib |> define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
 
@@
 
-Here we show the community polygons themselves as well as the log-density of rooms in each.
+Here we show the community polygons themselves as well as the log-density of
+rooms in each.
 
@@
 
-visualize_communities <- function(x, lbl = region_labels){
+visualize_communities <- function(x){
 
@@
 
-  h <- hell(x$dem)
-
-  members <- x$members |> 
-    st_drop_geometry() |> 
+  h <- hell(x[["dem"]])
+
+  members <- x[["cell_table"]] |> 
     group_by(community) |> 
@@
 
-  communities <- x$communities |> 
+  communities <- x[["communities"]] |> 
     left_join(members, by = "community") |> 
@@
   ggplot() +
-    ggtitle(lbl(x$name)[[1]]) +
+    ggtitle(region_labels[attr(x, "name")]) +
     geom_spatraster(
-      data = h$shade, 
-      fill = h$grays,
+      data = h[["shade"]], 
+      fill = h[["grays"]],
       maxcell = Inf
     ) +
-    geom_spatraster(data = x$dem, alpha = 0.5,) +
+    geom_spatraster(data = x[["dem"]], alpha = 0.5) +
     scale_fill_distiller(palette = "Greys", guide = "none") +
@@
 ::: {#fig-communities}
-![](../figures/communities-all.png){width="100%" fig-alt="Map shows defined community boundaries with fill colors representing the log density of rooms in each."}
-
-Map shows defined community boundaries with fill colors representing the log density of rooms in each.
+![](../figures/communities-all.png){width="100%"
+fig-alt="Map shows defined community boundaries with fill colors representing the log density of rooms in each."}
+
+Map shows community boundaries with fill colors representing the log density of
+rooms in each.
 :::
@@
   # AREA
-  A <- x$communities |> 
+  A <- x[["communities"]] |> 
     st_area() |> 
@@
   tbl_area <- tibble(
-    region = x$name, 
-    community = x$communities$community,
+    community = x[["communities"]][["community"]],
     area = A
@@
   # ROOMS
-  tbl_rooms <- x$sites |> 
-    st_join(x$communities[, "community"]) |> 
-    select(region, community, n_room) |> 
-    st_drop_geometry() |> 
+  tbl_rooms <- x[["cell_table"]] |> 
     group_by(community) |> 
-    summarize(
-      region = unique(region),
-      n_room = sum(n_room)
+    summarize(n_room = sum(n_room))
+
+  # FARMS
+  tbl_sites <- x[["cell_table"]] |> 
+    filter(center == 0) |> 
+    group_by(community) |> 
+    summarize(n_farm = n())
+
+  # CENTERS
+  tbl_centers <- x[["cell_table"]] |> 
+    filter(center == 1) |> 
+    group_by(community) |> 
+    summarize(n_center = n())
+
+  # COMMUTES
+  # subset dm (distance matrix) to get dm[farms, centers]
+  farms <- x[["cell_table"]] |> filter(center == 0)
+  centers <- x[["cell_table"]] |> filter(center == 1)
+
+  i <- which(rownames(x[["dm"]]) %in% farms[["cell"]])
+  j <- which(colnames(x[["dm"]]) %in% centers[["cell"]])
+
+  dm <- x[["dm"]][i, j]
+
+  min_dist <- tibble(
+    farm = as.numeric(rownames(dm)),
+    distance = apply(dm, 1, min)
+  )
+
+  tbl_commutes <- x[["cell_table"]] |> 
+    filter(center == 0) |> 
+    left_join(min_dist, by = join_by(cell == farm)) |> 
+    group_by(community) |> 
+    summarize(commute = mean(distance, na.rm = TRUE))
+
+  # GATHER EVERYTHING INTO ONE TABLE
+  tbl_area |> 
+    left_join(tbl_rooms, by = "community") |> 
+    left_join(tbl_sites, by = "community") |>
+    left_join(tbl_centers, by = "community") |> 
+    left_join(tbl_commutes, by = "community") |> 
+    mutate(
+      region = attr(x, "name"),
+      room_density = n_room/area
     ) |> 
@@
 
-  # FARMS
-  tbl_sites <- x$members |> 
-    st_drop_geometry() |> 
-    filter(type == "farm") |> 
-    group_by(community) |> 
-    summarize(n_farm = n(), region = unique(region))
-
-  # CENTERS
-  tbl_centers <- x$members |> 
-    st_drop_geometry() |> 
-    filter(type == "center") |> 
-    group_by(community) |> 
-    summarize(n_center = n(), region = unique(region))
-
-  # COMMUTES
-  cell_lookup <- x$sites |> 
-    st_drop_geometry() |> 
-    select(site_id, cell)
-
-  tbl_commutes <- x$members |> 
-    st_drop_geometry() |> 
-    filter(type == "center") |> 
-    left_join(cell_lookup, by = "site_id") |> 
-    left_join(x$nearest_center, by = join_by(cell == center), relationship = "many-to-many") |> 
-    group_by(community) |> 
-    summarize(commute = mean(distance, na.rm = TRUE), region = unique(region))
-
-  # GATHER EVERYTHING INTO ONE TABLE
-  tbl_area |> 
-    left_join(tbl_rooms, by = c("region", "community")) |> 
-    left_join(tbl_sites, by = c("region", "community")) |>
-    left_join(tbl_centers, by = c("region", "community")) |> 
-    left_join(tbl_commutes, by = c("region", "community")) |> 
-    mutate(room_density = n_room/area)
-
-}
-
-```
-
-```{r}
-
-results <- list(cmv, nrg, cib) |> map(get_community_data) |> bind_rows()
+}
+
+```
+
+```{r}
+
+results <- list(cmv, nrg, cib) |> 
+  map(get_community_data) |> 
+  bind_rows()
 
@@
 ```{r}
-#| eval: false
-#| echo: false
-
-# this code is just for exploring communities
-# and associating important (re: named) centers with them
-plot_labels <- function(x){
-
-  ggplot(x$communities) +
-    geom_sf() +
-    geom_sf_label(aes(label = community)) +
-    theme_void()
-
-}
-
-get_members <- function(x){
-
-  x$members |> 
-    st_drop_geometry() |> 
-    mutate(region = x$name, .before = everything())
-
-}
-
-get_sites <- function(x){ x$sites |> st_drop_geometry() }
-
-results |> 
-  group_by(region) |> 
-  slice_max(room_density, n = 5) |> 
-  left_join(
-    map(list(cmv, nrg, cib), get_members) |> bind_rows(), 
-    by = c("region", "community")
-  ) |> 
-  left_join(
-    map(list(cmv, nrg, cib), get_sites) |> bind_rows() |> filter(center == 1), 
-    by = c("region", "site_id")
-  ) |> 
-  select(region, community, site_id, site_name, area) |> 
-  filter(region == "cib") |> View()
-
-```
-
-```{r}
 #| code-fold: true
@@
 
-lblr <- as_labeller(c(
-  "commute" = "Mean Commute to Nearest Center [mins]",
-  "room_density" = "Room Density [N/km2]",
-  "farm_area" = "Mean Area for Farms [km2/N]",
-  "center_area" = "Mean Area for Centers [km2/N]",
-  "farm_ratio" = "Ratio of Farms [N] to Centers [N]",
-  "area" = "Total Area [km2]"
-))
+q95 <- quantile(results[["room_density"]], 0.95)
 
@@
     commute = commute/60,
-    farm_ratio = n_farm/n_center
+    farm_ratio = n_farm/n_center,
+    room_density = ifelse(room_density > q95, NA, room_density)
   ) |> 
@@
     strip.position = "bottom",
-    labeller = lblr
+    labeller = pretty_labels
   ) +
@@
 ::: {#fig-distributions}
-![](../figures/results-boxplot.png){width="75%" fig-alt="Distributions of important variables across communities and regions."}
-
-Distributions of important variables across communities and regions.
+![](../figures/results-boxplot.png){width="75%"
+fig-alt="Distributions of important variables across communities and regions."}
+
+Distributions of important variables across communities and regions. Some of the
+extreme room density estimates were dropped to make the main distribution more
+visible.
 :::
@@
 
+gpkg <- here("data", "community-centers.gpkg")
+
 bind_rows(
-  cmv$members |> mutate(region = "cmv", .before = site_id),
-  nrg$members |> mutate(region = "nrg", .before = site_id),
-  cib$members |> mutate(region = "cib", .before = site_id)
+  cmv[["sites"]] |> st_join(cmv[["communities"]]),
+  nrg[["sites"]] |> st_join(nrg[["communities"]]),
+  cib[["sites"]] |> st_join(cib[["communities"]])
 ) |> 
@@
 bind_rows(
-  cmv$communities, 
-  nrg$communities, 
-  cib$communities
+  cmv[["communities"]] |> mutate(region = "cmv"), 
+  nrg[["communities"]] |> mutate(region = "nrg"), 
+  cib[["communities"]] |> mutate(region = "cib")
 ) |> 
@@
   write_sf(gpkg, layer = "communities")
+
+```
+
+## Sensitivity Analysis
+
+This is expensive computing, but it's worth checking. Note that we don't do
+every combination of parameters, but we do get decent variety relative to the
+default values chosen above. Most of the simulations take about 4 minutes to run
+on my machine. The exceptions are those involving high resolution graphs
+(constructed from DEMs with a low aggregation factor). Those take considerably
+longer, as much as an hour, even with all the shortcuts I added. There are 19
+total simulations. All together, they take about 2.2 hours to run.
+
+For reference, here are the specs on my machine:
+
+-   CPU: AMD Ryzen 7 5800U with Radeon Graphics
+    -   Base speed: 1.90 GHz
+    -   Cores: 8
+    -   Logical processors: 16
+-   RAM: 16.0 GB (15.4 GB usable)
+-   System type: 64-bit operating system, x64-based processor
+-   OS: Windows 11 Home
+
+```{r}
+#| code-fold: true
+
+run_simulation <- function(.x, .args){
+
+  .x |> 
+    build_region() |> 
+    aggregate_grid(.factor = .args[["agg_factor"]]) |> 
+    group_sites_by_cell() |> 
+    add_graph(max_slope = 45) |> 
+    add_distance_matrix() |> 
+    remove_distant_farms(dmax = .args[["dmax"]]) |> 
+    join_neighboring_centers(p = .args[["p"]], djoin = .args[["djoin"]]) |> 
+    collect_members() |> 
+    drop_small_communities(threshold = 3) |> 
+    define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100)) |> 
+    get_community_data()
+
+}
+
+sensitivity <- function(x, args){
+
+  # it seems my computer isn't up to the task of doing this one in parallel
+  if (args[["agg_factor"]] == 1){
+
+    dfs <- map(x, \(.x, .args = args){
+
+      run_simulation(.x, .args)
+
+    })
+
+  } else {
+
+    # there's no randomization in this code, but furrr belches out warnings
+    # anyway, so i add the furrr option to set seed.
+    dfs <- future_map(x, \(.x, .args = args){
+
+      options(scipen = 9999)
+      terra::terraOptions(progress = 0, print = FALSE)
+
+      run_simulation(.x, .args)
+
+    }, .options = furrr_options(seed = 1701))
+
+  }
+
+  bind_rows(dfs)
+
+}
+
+```
+
+```{r}
+#| time_it: true
+
+sensitivity_parameters <- bind_rows(
+  tibble(
+    focal_variable = "agg_factor",
+    agg_factor = 1:4,
+    dmax = one_hour,
+    p = 0.8,
+    djoin = half_hour
+  ),
+  tibble(
+    focal_variable = "dmax",
+    agg_factor = 3,
+    dmax = 60 * seq(40, 80, by = 10),
+    p = 0.8,
+    djoin = half_hour
+  ),
+  tibble(
+    focal_variable = "p",
+    agg_factor = 3,
+    dmax = one_hour,
+    p = seq(0.5, 0.9, by = 0.1),
+    djoin = half_hour
+  ),
+  tibble(
+    focal_variable = "djoin",
+    agg_factor = 3,
+    dmax = one_hour,
+    p = 0.8,
+    djoin = 60 * seq(20, 40, by = 5)
+  )
+)
+
+sensitivity_results <- sensitivity_parameters |> 
+  rowwise() |> 
+  mutate(
+    analysis = list(sensitivity(
+      c("cmv", "nrg", "cib"), 
+      args = list(
+        "agg_factor" = agg_factor, 
+        "dmax" = dmax, 
+        "p" = p, 
+        "djoin" = djoin
+      )
+    ))
+  ) |> 
+  ungroup()
+
+```
+
+```{r}
+#| code-fold: true
+
+graph_data <- sensitivity_results |> 
+  mutate(analysis = map(analysis, \(.x){
+
+    .x |>
+      group_by(region) |>
+      summarize(
+        n_communities = n(),
+        area = median(area),
+        commute = median(commute/60),
+        room_density = median(room_density)
+      )
+
+  })) |> 
+  unnest(analysis) |> 
+  pivot_longer(
+    c(n_communities, area, room_density, commute),
+    names_to = "outcome_variable",
+    values_to = "outcome_value"
+  ) |> 
+  mutate(
+    focal_value = case_when(
+      focal_variable == "dmax" ~ dmax,
+      focal_variable == "p" ~ p,
+      focal_variable == "djoin" ~ djoin,
+      focal_variable == "agg_factor" ~ agg_factor,
+      TRUE ~ NA
+    ),
+    focal_value = case_when(
+      focal_variable == "agg_factor" & focal_value == 1 ~ 27,
+      focal_variable == "agg_factor" & focal_value == 2 ~ 55,
+      focal_variable == "agg_factor" & focal_value == 3 ~ 82,
+      focal_variable == "agg_factor" & focal_value == 4 ~ 109,
+      TRUE ~ focal_value
+    ),
+    region = fct_relevel(region, "cmv", "nrg")
+  )
+
+agg_factor <- graph_data |> 
+  filter(focal_variable == "agg_factor") |> 
+  pull(focal_value) |> 
+  unique() |> 
+  sort()
+dmax <- 60 * seq(40, 80, by = 10)
+djoin <- 60 * seq(20, 40, by = 5)
+
+x_scale <- function(x, .labels = x){
+  scale_x_continuous(breaks = x, labels = .labels)
+}
+
+x_scales <- list(
+  focal_variable == "agg_factor" ~ x_scale(agg_factor),
+  focal_variable == "djoin" ~ x_scale(djoin, .labels = djoin/60),
+  focal_variable == "dmax" ~ x_scale(dmax, .labels = dmax/60),
+  focal_variable == "p" ~ x_scale(seq(0.5, 0.9, by = 0.1))
+)
+
+gg <- ggplot(graph_data, aes(focal_value, outcome_value, color = region)) + 
+  geom_line() +
+  geom_point(size = 1.2) +
+  scale_color_manual(
+    name = NULL, 
+    values = region_colors,
+    labels = region_labels
+  ) +
+  facet_grid(
+    rows = vars(outcome_variable),
+    cols = vars(focal_variable),
+    switch = "both",
+    labeller = squished_labels,
+    scales = "free"
+  ) +
+  facetted_pos_scales(x = x_scales) +
+  theme(
+    axis.title = element_blank(),
+    legend.box.margin = margin(b = -7),
+    legend.direction = "vertical",
+    legend.justification = "left",
+    legend.key.height = unit(10, "pt"),
+    legend.margin = margin(),
+    legend.position = "top",
+    legend.spacing.y = unit(5, "pt"),
+    strip.placement = "outside",
+    strip.text = element_text(size = rel(1), margin = margin(t = -1, b = 10))
+  ) +
+  guides(color = guide_legend(byrow = TRUE))
+
+fn <- here("figures", "sensitivity-analysis.png")
+
+ggsave(
+  plot = gg,
+  filename = fn,
+  width = 6.6,
+  height = 7.33,
+  dpi = 600
+)
+
+prepare_image(fn)
+
+remove(graph_data, agg_factor, dmax, djoin, x_scale, x_scales, gg, fn)
+
+```
+
+<div>
+
+![](../figures/sensitivity-analysis.png){width="100%"
+fig-alt="Figure shows results of sensitivity analysis."}
+
+Figure shows results of sensitivity analysis, focusing on three main paremeters
+(djoin, dmax, and p) and the level of aggregation of the grid, which corresponds
+to its resolution. Measures of some key outcomes are also shown, including the
+number of communities, their total area, room density, and average commute time
+from farms to the nearest community center.
+
+</div>
+
+### Save Results
+
+```{r}
+
+sensitivity_results |> 
+  unnest(analysis) |> 
+  write_csv(file = here("data", "sensitivity-analysis.csv"))
 

### 2024-02-14 R/get-elevation.R
previous code stopped working, so refactored to use vrt directly, also dropped aggregation and saved at native resolution

@@ -1,41 +1,44 @@
 
 # get elevation data
-
-library(FedData)
 library(here)
 library(purrr)
 library(sf)
 library(terra)
 
-download_dem <- function(region, geometry){
+download_dem <- function(x){
   
-  r <- geometry |> 
-    st_sfc(crs = 4326) |> 
-    get_ned(res = "1", label = region) |> 
-    rast() |> 
-    aggregate(fact = 3)
-  
-  names(r) <- region
-  
-  return(r)
+  # location of the virtual raster pointing to the
+  # USGS one arc second elevation data (~30 m resolution)
+  elevation_source <- file.path(
+    "https://prd-tnm.s3.amazonaws.com/StagedProducts",
+    "Elevation/1/TIFF",
+    "USGS_Seamless_DEM_1.vrt"
+  )
+
+  # connect to vrt
+  rr <- terra::rast(elevation_source, vsi = TRUE)
+
+  feature <- terra::vect(x)
+  feature <- terra::project(feature, terra::crs(rr))
+  feature <- terra::ext(feature)
+  feature <- terra::vect(feature, crs = terra::crs(rr))
+
+  fn <- paste0("dem-", x[["region"]], ".tif")
+
+  rr <- terra::crop(
+    rr,
+    feature,
+    snap = "out",
+    mask = TRUE,
+    filename = here("data", fn),
+    datatype = "FLT4S",
+    gdal = c("COMPRESS=DEFLATE", "ZLEVEL=9"),
+    overwrite = TRUE
+  )
   
 }
 
-write_dem <- function(x){
-  
-  fn <- here("data", paste0("dem-", names(x), ".tif"))
-  
-  writeRaster(x, filename = fn, overwrite = TRUE)
-  
-}
-
-# for some reason, values aren't read in after downloading the first time,
-# so have to call this twice
 here("data", "community-centers.gpkg") |> 
   read_sf(layer = "regions") |> 
-  pmap(download_dem)
-
-here("data", "community-centers.gpkg") |> 
-  read_sf(layer = "regions") |> 
-  pmap(download_dem) |> 
-  walk(write_dem)
+  split(1:3) |> 
+  walk(download_dem)

### 2024-02-16 R/community-clustering.qmd
update description of algorithm and replace depracated igraph function

@@
 
-Here the population, $N$, of focal center $c$ is defined as
-
-$$N_{c} = R_{c} + \sum_{k=1}^{S} R_{k} \cdot I(t_{kc} \leq D_{c})$$
-
-for all sites $S$ (this includes both farms and centers), with $R_{c}$ being the
-room count at $c$ and $R_{k}$ the room count at site $k$, $t_{kc}$ the travel
-time from $k$ to $c$, $D_{c}$ the threshold travel time to $c$ (referred to as
-*D-join* in the paper), and $I$ the indicator function that is 1 if $t \leq D$
-and 0 otherwise.
+Here the catchment room count, $N$, of focal center $c$ is defined as
+
+$$N_{c} = \sum_{k=1}^{S} R_{k} \cdot I(t_{kc} \leq D)$$
+
+for all sites $S$ (including both farms and centers), with $R_{k}$ being the
+room count at site $k$, $t_{kc}$ the travel time from $k$ to $c$, $D$ the
+threshold travel time to a center (referred to as *D-join* in the paper), and
+$I$ the indicator function that is 1 if $t \leq D$ and 0 otherwise. Note that
+$S$ includes $c$ and that $t_{cc} = 0$ (the travel time from a center to the
+same center is precisely 0), so $R_{c}$ is always included in $N_{c}$.
+
+The catchment room count within $D$ of two centers $c$ and $d$ is then
+
+$$N_{cd} = \sum_{k=1}^{S} R_{k} \cdot I(t_{kc} \leq D) \cdot I(t_{kd} \leq D)$$
+
+and
+
+$$p_{cd} = N_{cd}/N_{c}$$
+
+so we could also state our rule as: if $p_{cd} \geq \alpha$ for some critical
+threshold $\alpha$, then $c$ is part of the same community as $d$. In this case,
+the default threshold is 0.8 relative to a commute time distance of one half
+hour.
 
@@
   # so, igraph it is...
-  g <- igraph::graph.adjacency(neighbors_matrix)
+  g <- igraph::graph_from_adjacency_matrix(neighbors_matrix)
   membership <- igraph::components(g)[["membership"]]
@@
   filename = fn,
-  width = 6.6,
-  height = 7.33,
+  width = 6.7,
+  height = 7.,
   dpi = 600

### 2024-03-02 R/community-clustering.qmd
considerable refactoring to ensure everything is doing what it is supposed to be doing

@@
 
-## Introudction
+## Introduction
 
@@
 
-**Data:** VEPII and Cibola databases
+**Data:** Village Ecodynamics Project and Southwest Social Networks databases
 
@@
   "dmax" = "D-max [mins]",
-  "p" = "p [proportion in D-join]",
+  "p" = "P [proportion in D-join]",
   "agg_factor" = "Grid Resolution [m]"
@@
   "dmax" = "D-max\n[mins]",
-  "p" = "p\n[prop. in D-join]",
+  "p" = "P\n[prop. in D-join]",
   "agg_factor" = "Grid Resolution\n[m]"
@@
 it does come at the cost of precision in terms of path analysis across the
-landscape, but that is not as bad as it sounds. The idea that we could build a
-model that is perfectly isomorphic with the actual or true path taken by people
-a thousand years ago is wildly implausible. The idea of the 'true path' is also
-problematic as we're not evaluating a single commute, but rather the path people
-should take more often than not. So, the real question is what level of
-precision is sufficient for evaluating that. And the answer is, of course, it
+landscape, though that is not as bad as it sounds. The idea that we could build
+a model that is perfectly isomorphic with the actual or true path taken by
+people a thousand years ago is wildly implausible. The idea of the 'true path'
+is also problematic as we're not evaluating a single commute, but rather the
+path people should take more often than not. So, the real question is what level
+of precision is sufficient for evaluating that. And the answer is, of course, it
 just depends.
@@
 of an entirely pedestrian population based on the distribution of farms around
-urban centers, probably the difference between \~30 m and \~80 m grid cells
+urban centers, probably the difference between \~30 m and \~90 m grid cells
 isn't that big of a deal. If that doesn't convince you, though, we do some
@@
 
-To calculate travel time between every pair of site points, we use a
-bi-directional version of Dijkstra's algorithm as implemented in the R package
-`{cppRouting}`. Note that we also average the from- and to- travel times, so the
-results are isotropic (meaning direction is irrelevant).
+To calculate travel time between every pair of site points, we use Dijkstra's
+algorithm as implemented in the R package `{cppRouting}`. Note that we also
+average the from- and to- travel times, so the results are isotropic (meaning
+direction is irrelevant).
 
@@
 1.  Identify community centers
-2.  Remove distant farms
-3.  Join centers to their overlapping neighbors
+2.  Join centers to their overlapping neighbors
+3.  Remove distant farms
 4.  Join farms to their nearest center
@@
 
-Step 1 is already done and stored in the data. Step 2 can technically be done
-after 2 and 3, but it makes sense to do it first, since it means those sites can
-be ignored afterwards.
+Step 1 is already done and stored in the data. Step 3 is technically done before
+2 since it means those sites can be ignored afterwards.
 
@@
 
+```{r}
+#| code-fold: true
+
+list(cmv, nrg, cib) |> 
+  map(\(x){
+
+    tibble(
+      region = attr(x, "name"),
+      n_outliers = attr(x, "n_outliers")
+    )
+
+  }) |> 
+  bind_rows() |> 
+  gt() |> 
+  tab_header("No. of Dropped Farms") |> 
+  opt_align_table_header("left") |> 
+  tab_options(column_labels.hidden = TRUE)
+
+```
+
 ### Join neighboring centers
@@
 
-Here the catchment room count, $N$, of focal center $c$ is defined as
-
-$$N_{c} = \sum_{k=1}^{S} R_{k} \cdot I(t_{kc} \leq D)$$
-
-for all sites $S$ (including both farms and centers), with $R_{k}$ being the
-room count at site $k$, $t_{kc}$ the travel time from $k$ to $c$, $D$ the
+Here the catchment room count, $N$, of focal center $i$ is defined as
+
+$$N_{c} = \sum_{i=1}^{S} R_{i} \cdot I(t_{ic} \leq D)$$
+
+for all sites $S$ (including both farms and centers), with $R_{i}$ being the
+room count at site $i$, $t_{ic}$ the travel time from $i$ to $c$, $D$ the
 threshold travel time to a center (referred to as *D-join* in the paper), and
@@
 
-$$N_{cd} = \sum_{k=1}^{S} R_{k} \cdot I(t_{kc} \leq D) \cdot I(t_{kd} \leq D)$$
+$$N_{cd} = \sum_{i=1}^{S} R_{i} \cdot I(t_{ic} \leq D) \cdot I(t_{id} \leq D)$$
 
@@
   # subset dm (distance matrix) to get dm[farms and centers, centers]
-  center_cells <- x[["cell_table"]] |> filter(center == 1) |> pull(cell)
-
-  j <- which(colnames(x[["dm"]]) %in% center_cells)
-
-  dm <- x[["dm"]][, j]
+  i <- x[["cell_table"]] |> pull(cell)
+  j <- x[["cell_table"]] |> filter(center == 1) |> pull(cell)
+
+  i <- which(rownames(x[["dm"]]) %in% i)
+  j <- which(colnames(x[["dm"]]) %in% j)
+
+  dm <- x[["dm"]][i, j]
+
+  # room count in each farm/center
+  Rk <- x[["cell_table"]] |> pull(n_room)
+  names(Rk) <- x[["cell_table"]] |> pull(cell)
+
+  # want to make sure these are arranged in the same order as rows in dm
+  Rk <- Rk[rownames(dm)]
 
   # find farms and centers that are within distance djoin of other centers
-  djoins <- apply(dm, 1, \(.x){ which(.x <= djoin) })
-
-  # remove farms that are isolated (ie, greater than djoin from a center)
-  # note: this should just be farms because all centers are less than
-  #   djoin from themselves
-  djoins <- djoins[lengths(djoins) > 0]
-
-  # collect list into a long-form table
-  djoins <- tibble(
-    center = colnames(dm)[unlist(djoins)],
-    farm_or_center = rep(names(djoins), lengths(djoins))
-  ) |> mutate(across(everything(), as.numeric))
-
-  # now we combine the room counts with the djoins table
-  djoins <- djoins |> 
-    left_join(
-      x[["cell_table"]] |> select(-center), 
-      by = join_by(farm_or_center == cell)
-    ) |> 
-    rename("population" = n_room)
-
-  # calculate the total population of the catchment around each center
-  # note that this involves a lot of double-counting, farms and centers
-  # counting towards the population size of multiple centers, but this is OK
-  # as we are interested in using the shared population as a measure of 
-  # interaction between centers, not as an estimate of a community's population
-  gen_pop <- djoins |> 
-    group_by(center) |> 
-    summarize(population = sum(population))
-
-  # for each center, what other centers have larger catchment populations?
-  # using outer() is extremely fast
-  # in this case, it creates an asymmetric logical matrix
-  R <- outer(gen_pop[["population"]], gen_pop[["population"]], FUN = '<=')
-  diag(R) <- FALSE
-
-  colnames(R) <- gen_pop[["center"]]
-  rownames(R) <- gen_pop[["center"]]
-
-  # now, we turn that matrix into a big list 
-  # each element is a vector with the cell IDs of the larger centers
-  R <- apply(R, 1, which)
-  R <- map(R, \(x){ as.numeric(names(x)) })
-
-  # for each center, we look at all the centers with bigger catchment
-  # populations than it, and ask: is the shared population size some fraction
-  # of the total population of the catchment?
-  neighbors_matrix <- purrr::imap(R, \(x, idx){
-
-    N <- gen_pop |> 
-      filter(center == as.numeric(idx)) |> 
-      pull(population)
-
-    focal_neighbors <- djoins |> 
-      filter(center == as.numeric(idx)) |> 
-      pull(farm_or_center)
-
-    cntrs <- djoins |> 
-      filter( # this filter gives the intersection of farms/cells 
-        center %in% x,
-        farm_or_center %in% c(idx, focal_neighbors)
-      ) |> 
-      group_by(center) |> 
-      summarize(proportion = sum(population)/N) |> 
-      filter(proportion >= p) |> 
-      pull(center)
-
-    # return a logical vector the length of R
-    names(R) %in% cntrs
-
-  })
-
-  neighbors_matrix <- do.call("rbind", neighbors_matrix)
-  diag(neighbors_matrix) <- TRUE
-
-  colnames(neighbors_matrix) <- names(R)
-  rownames(neighbors_matrix) <- names(R)
-
-  # there's a recursive dimension to this:
+  # this creates a binary/logical matrix of 1s and 0s,
+  # which is what you get from the indicator function:
+  # 1 if row is in distance djoin of column, 0 otherwise
+  B <- matrix(as.numeric(dm <= djoin), nrow = nrow(dm))
+
+  dimnames(B) <- dimnames(dm)
+
+  # square community centers matrix
+  G <- matrix(NA, nrow = ncol(dm), ncol = ncol(dm))
+
+  colnames(G) <- colnames(dm)
+  rownames(G) <- colnames(dm)
+
+  # get total number of rooms within djoin of each center
+  # this multiplies each column of B (which is just a bunch of 0s and 1s)
+  # by the room count vector Rk, then takes the sum of the columns. 
+  # we put this into the diagonal because it answers the question:
+  # how many rooms does a center share with itself?
+  # this is Nc. or Ncc since d = c.
+  diag(G) <- colSums(B * Rk)
+
+  # get shared room count (Ncd) within djoin of all pairs of centers
+  # note we dump this into the lower triangle! this is a real R gotcha
+  # as the vector is passed into the matrix by column, not row, so you 
+  # can't do this with the upper triangle!
+  lt <- lower.tri(G)
+
+  G[lt] <- combn(
+    ncol(B), 
+    m = 2, 
+    FUN = \(.x){ sum(B[, .x[1]] * B[, .x[2]] * Rk) }
+  )
+
+  # this matrix should be symmetric: 
+  # the diagonal is Ncc, the room count a center shares with itself
+  # the off-diagonals are Ncd = Ndc, the room counts shared between different centers
+  ut <- upper.tri(G)
+
+  G[ut] <- t(G)[ut]
+
+  # create adjacency matrix
+  # by dividing each column of G by the diagonal of G, we get the proportion
+  # of the shared room count to the total room count, so p or pcd
+  # we then compare this to the critical threshold, to determine adjacency
+  neighbors_matrix <- (G / diag(G)) >= p
+
+  # now to separate them into communities
+  # unfortunately, there's a recursive dimension to this:
   # if a center is joined with a center that's joined with a center, etc.
-  # i don't know how to do that in R
-  # so, igraph it is...
+  # i don't know how to do that in R, so igraph it is...
   g <- igraph::graph_from_adjacency_matrix(neighbors_matrix)
@@
 cib <- cib |> join_neighboring_centers(p = 0.8, djoin = half_hour)
+
+```
+
+```{r}
+#| code-fold: true
+
+list(cmv, nrg, cib) |> 
+  map(\(x){
+
+    tibble(
+      region = attr(x, "name"),
+      n_communities = attr(x, "n_communities")
+    )
+
+  }) |> 
+  bind_rows() |> 
+  gt() |> 
+  tab_header("No. of Communities") |> 
+  opt_align_table_header("left") |> 
+  tab_options(column_labels.hidden = TRUE)
 
@@
 
+```{r}
+#| code-fold: true
+
+list(cmv, nrg, cib) |> 
+  map(\(x){
+
+    tibble(
+      region = attr(x, "name"),
+      n_dropped = attr(x, "n_dropped")
+    )
+
+  }) |> 
+  bind_rows() |> 
+  gt() |> 
+  tab_header("No. of Dropped Communities") |> 
+  opt_align_table_header("left") |> 
+  tab_options(column_labels.hidden = TRUE)
+
+```
+
 ### Define boundaries
@@
 community. This prevents the search algorithm from searching out away from the
-other sites in the community.
+other sites in the community. The distance used for pruning is only meaningful
+for pedestrian commute times.
 
@@
 Here we show the community polygons themselves as well as the log-density of
-rooms in each.
+rooms in each. Note the very pronounced edge effect along Mesa Verde National
+Park's administrative boundary.
 
@@
   filename = fn,
-  width = 6.7,
-  height = 7.,
+  width = 7.1,
+  height = 7.3,
   dpi = 600

### 2024-03-02 R/labeled-overview-map.R
rev req: labeled overview map

@@ -1,4 +1,6 @@
 
+library(ggfx)
+library(ggspatial)
 library(here)
 library(sf)
 library(terra)
@@ -110,18 +112,23 @@ suppressWarnings({
   
 })
 
-x <- bb8[["xmax"]]
+y_ut <- four_corners |> 
+  filter(name == "Utah") |> 
+  st_bbox()
 
-y <- four_corners |> 
+y_ut <- y_ut[["ymin"]]
+
+y_co <- four_corners |> 
   filter(name == "Colorado") |> 
   st_bbox()
 
-y <- y[["ymin"]]
+y_co <- y_co[["ymin"]]
 
 state_labels <- tibble(
-  x = x - 0.05,
-  y = y + c(0.07, -0.07),
-  label = c("Colorado", "New Mexico")
+  x = c(bb8[["xmin"]] + c(0.05, 0.05), bb8[["xmax"]] - c(0.05, 0.05)), 
+  y = c(y_ut + c(0.07, -0.07), y_co + c(0.07, -0.07)),
+  label = c("UT", "AZ", "CO", "NM"),
+  hjust = c(0, 0, 1, 1)
 )
 
 landform_labels <- bind_rows(
@@ -209,7 +216,7 @@ landform_labels <- bind_rows(
     xs = -105.94354,
     ys = 35.68533,
     x = -105.94354,
-    y = 35.50000,
+    y = 35.48000,
     label = "Santa Fe",
     hjust = 1
   ),
@@ -218,7 +225,7 @@ landform_labels <- bind_rows(
     ys = 35.03788,
     x = -108.16500,
     y = 35.00788,
-    label = "El Morro",
+    label = "El Morro NM",
     hjust = 0
   ),
   tibble(
@@ -263,7 +270,7 @@ landform_labels <- bind_rows(
   )
 )
 
-remove(x, y)
+remove(y_ut, y_co)
 
 # map ---------------------------------------------------------------------
 
@@ -276,60 +283,77 @@ ggplot() +
   ) +
   geom_sf(
     data = cover,
-    fill = alpha("white", 0.75)
+    fill = alpha("white", 0.5)
   ) +
   geom_sf(
     data = four_corners,
     fill = "transparent",
+    color = "gray20",
     linewidth = 0.2
   ) +
   geom_sf(
     data = regions,
     color = "black",
-    fill = "transparent",
+    fill = alpha("white", 0.05),
     linewidth = 0.5
   ) +
[…]

### 2024-03-02 manuscript/ras-reviews-response.md
response to reviews

Reviewers provided important and insightful comments that helped us improve this
paper considerably. Perhaps the biggest revision is that we now include some
limited sensitivity analysis to evaluate what happens when parameter values are
changed. As two of the reviewers raised concerns about the resolution of the
grid, tests were also conducted to see if higher resolutions would have a
noticeable impact on the results - they did not. Finally, two of the reviewers
asked that we add some broader implications to the discussion. This was done by
highlighting differences between the types of communities we are trying to
understand and other types of communities (religious, linguistic, etc.) that
archaeologists might find relevant to their research.

Note: minor formatting and spelling errors identified by the reviewers were
corrected in the text but are not discussed here.

### Response to Reviewer 1

1.  Distance traveled should be measured with hypotenuse.

    Response: The reviewer is correct about this, so the analysis was updated to
    account for it. It required only minor changes to the R code. We also note
    that the effect was minimal as a maximum slope was defined to constrain
    travel paths, so the difference in travel time ended up being less than a
    minute in most cases.

2.  Use 5th percentile in Lorentz curve as recommended by Campbell.

    Response: This was a major oversight on our part that we have now corrected,
    and we thank the reviewer for bringing it to our attention. As requested, we
    also clarify that we are using the 5th percentile of the Lorentz curve in
    text.

3.  Aggregation of elevation model grid cells biases estimates.

    Response: The concern is valid, so we conduct an additional sensitivity
    analysis at four different levels of aggregation (from no aggregation to the
    level of aggregation used in the paper). Descriptions of this were added to
    the methods and results. To summarize, we found that increasing resolution
    beyond that used in our paper did not have a meaningful impact on the
    results.

4.  Bi-directional Dijkstra forces isotropy.

    Response: In bi-directional search on a directed graph, search from the
    destination node uses edge weights in the direction of the destination node,
    not edge weights in the direction of the origin node as the reviewer seems
    to be implying, so anisotropy is preserved. At any rate, it appears to be a
    moot point since `{cppRouting}` uses regular old Dijkstra's algorithm when
    `get_distance_matrix()` is applied to an uncontracted graph (the method used
    here). So, what the reviewer wants is what we actually did. We just
    described it wrong. This has been corrected in the text.

5.  Reviewer proposes alternative clustering strategy.

    Response: The proposed alternative strategy for defining communities is an
    intriguing one, but not fundamentally different than what we propose. There
    are three main differences.

    \(i\) Rather than searching the entire grid, search in our algorithm ends
    when all sites are reached, which reduces the time complexity.

    \(ii\) The proposed alternative doesn't specify a rule for joining
    communities, which we consider to be one of the two key contributions of our
    algorithm (the other is weighting distance by population size).

    \(iii\) If i understand it correctly, the proposed alternative for defining
    boundaries is to use the smallest isochrone around all the centers in a
    community that encompasses all the nearby farm sites. This is an intriguing
    suggestion that deserves further consideration, and we would love to see a
    follow-up paper that explores it more. We note, however, that it would not
    guarantee concavity in the community polygon with respect to time - meaning
    the shortest path between two sites in the community would not be guaranteed
    to remain in the community. Of course, it could be argued that that's a
    difference that doesn't actually make a difference, that these are just
    crude estimates anyway. Our argument would be that interactions did not
    occur only at community centers, so the shortest path between all sites is
    probably relevant to defining the spatial extent of these farming
    communities. If the people living on these farms were routinely visiting
    their neighbors, the routes they took would have formed an important element
    of their spatial or geographic community.

### Response to Reviewer 2

1.  Include additional citations in methods.

    Response: Done! See especially the section on defining community centers.

2.  Discuss broader connections outside the regional focus.

    Response: We add two paragraphs to the discussion elaborating on the type of
    community we are delineating and how that differs from other types of
    communities that may be of interest to archaeologists and anthropologists,
    focusing in particular on different dimensions of community organization -
    biological relatedness, shared metaphors, religion, language, etc.

3.  Add map with place names mentioned in text.

    Response: Done! This actually provides a lot of context that will hopefully
    help readers better interpret the results. In fact, any reference to
    specific locations in the text were updated to ensure that they used the
    labeled locations in the map. It was also an enjoyable exercise! Thanks for
    the request!

4.  Include some discussion of how visual landmarks might shape these
    communities.

    Response: This is an intriguing idea. Bernardini and Peeples (2017, "Sight
    communities," *Am Ant V*80 N2) provide a good example of this as they use
    visual landmarks and viewsheds to define "sight communities." That said, our
    goal here is to define the spatial extent of dispersed *farming*
    communities, not sight communities, so we focus on farmsteads that serve as
    indicators of the locations of farms and bind them together into communities
    using the centers at which farmers would have interacted on a fairly regular
    basis.

5.  Aggregation of elevation model grid cells biases estimates.

    Response: See response to reviewer 1.

6.  Clarify whether slope is degree slope or percent slope.

    Response: Campbell's hiking function is specified to work with degree slope
    by default. This was clarified in the text.

### Response to Reviewer 3

1.  Discuss broader connections outside the regional focus.

    Response: See response to Reviewer 2.

2.  Sensitivity analysis to evaluate model parameters.

    Response: Done! In general, we found that the algorithm behaved as expected,
    though we note some volatility in the Cibola data. This is probably owing to
    the geographic sparsity of that sample, particularly with respect to the
    ratio of farms to centers.

3.  Region-specific definitions of community centers.

    Response: To clarify our reasons for this, we added the following to the
    section on community centers:

    "For given differences in population density and settlement patterning, the
    frequency and intensity of interaction between individuals at centers is
    likely to vary by region, so it is reasonable to expect that community
    centers themselves (at least in so far as they play the role of centers)
    should vary in their size and composition across regions, too."

### 2024-03-02 manuscript/manuscript.qmd
multiple changes based on rev reqs

@@
   The transition from dispersed to aggregated forms of settlement reflects a critical shift in the relative value of social and primary (food) modes of production.
-Investigating trade-offs between them, however, requires estimates of the extent of community territories, including their nearby arable land.
+However, investigating trade-offs between them requires estimates of the extent of community territories, including their nearby arable land.
 Here we demonstrate a simple algorithm to do that.
 Our algorithm is analogous to that used to define core-based statistical areas for the US census, though instead of modern metropolitan areas, we rely on community centers (or areas of known and persistent interaction between unrelated individuals).
-We provide examples of our algorithm by applying it to archaeological sites in the central Mesa Verde, northern Rio Grande regions, and Cibola regions.
-Together, these data provide a comprehensive, high resolution record of prehistoric maize farming in the North American Southwest, representing varying levels of aggregation and dispersal through time.
+We provide examples of our algorithm by applying it to archaeological sites in the central Mesa Verde, northern Rio Grande, and Cibola regions.
+A sensitivity analysis is also conducted to demonstrate how each tuning parameter contributes to the algorithm.
 keywords:
@@
   - Community Detection
-  - Intensification
+  - Complex Systems
+  - Least-Cost Path Analysis
+  - Population Density
 target-journal: Journal of Archaeological Science
 bibliography: bibliography.bib
+corresponding_author: Kenneth.Vernon@colorado.edu
 linenumbers: false
-watermark: false
 format: 
   simple-pdf: default
+  docx: default
 execute: 
@@
 
-Urban geographers recognize density as a crucial property of human communities which influences many other properties, including productivity, resource needs, and infectious disease rates [@angel_etal2021; @bettencourt2021; @duranton_puga2020].
-The archaeological record of subsistence farming societies contains physical traces of community formations with a wide range of densities, often glossed as "dispersed" or "aggregated" [@birch2013; @drennan_etal2015; @gyucha2019].
-The primary difference between the two has to do with the spatial distribution of residents relative to primary (food) production.
-In dispersed formations, households are scattered across an area, interspersed with the land that they farmed, thus requiring them to commute to central places for various forms of social production.
-In aggregated formations, in contrast, households are clustered in a village or town, such that they reside in locations of social production, but must commute to fields for primary production.
+Urban geographers recognize density as a crucial property of human communities
+which influences many other properties, including productivity, resource needs,
+and infectious disease rates [@angel_etal2021; @bettencourt2021;
+@duranton_puga2020]. The archaeological record of subsistence farming societies
+contains physical traces of community formations with a wide range of densities,
+often glossed as "dispersed" or "aggregated" [@birch2013; @drennan_etal2015;
+@gyucha2019]. The primary difference between the two has to do with the spatial
+distribution of residents relative to primary (food) production. In dispersed
+formations, households are scattered across an area, interspersed with the land
+that they farmed, thus requiring them to commute to central places for various
+forms of social production. In aggregated formations, in contrast, households
+are clustered in a village or town, such that they reside in locations of social
+production, but must commute to fields for primary production.
 
-From a complex systems perspective, the main way these two settlement morphologies vary is with respect to which costs are being minimized.
-In the dispersed pattern, costs associated with primary (food) production by the household are being minimized, whereas in the aggregated pattern, it is costs associated with various forms of social production, from government and ritual to economic exchange and warfare.
-Presumably, choices regarding which sorts of costs to minimize are driven by the relative productivities of primary and social production, which is to say, their relative importance, broadly construed, for human well-being in a given context.
+From a complex systems perspective, the main way these two settlement
+morphologies vary is with respect to which costs are being minimized. In the
+dispersed pattern, costs associated with primary (food) production by the
+household are being minimized, whereas in the aggregated pattern, it is costs
+associated with various forms of social production, from government and ritual
+to economic exchange and warfare. Presumably, choices regarding which sorts of
+costs to minimize are driven by the relative productivity of primary and social
+production, which is to say, their relative importance, broadly construed, for
+human well-being in a given context.
 
-There are many episodes in history where human settlements transitioned from dispersed to aggregated, and presumably this is a signal of a change in the relative value of social vs.
-primary production for the residents.
-This can occur for a variety of reasons, including an increase in the social cost of not being aggregated (e.g., due to warfare); a decrease in transport costs for staples (due to animal traction, wheeled vehicles, roads, etc.); an increase in land productivity which changes the balance of costs of transport for primary vs.
-social production; or an increase in the contribution of exchange to household incomes, which can derive from increases in community size and density.
-In addition, while one can conceptualize transitions between dispersed and aggregated forms of settlement in general systems terms, in real human communities a variety of political or ideological factors can keep people from adopting an energetically balanced form of settlement given prevailing conditions.
-So, there are opportunities for archaeologists to study deviations from general equilibrium conditions in addition to factors that shift the equilibrium in a given context.
+There are many episodes in history where human settlements transitioned from
+dispersed to aggregated, and presumably this is a signal of a change in the
+relative value of social vs. primary production for the residents. This can
+occur for a variety of reasons, including an increase in the social cost of not
+being aggregated (e.g., due to warfare); a decrease in transport costs for
+staples (due to animal traction, wheeled vehicles, roads, etc.); an increase in
+land productivity which changes the balance of costs of transport for primary
+vs. social production; or an increase in the contribution of exchange to
+household incomes, which can derive from increases in community size and
+density. In addition, while one can conceptualize transitions between dispersed
+and aggregated forms of settlement in general systems terms, in real human
+communities a variety of political or ideological factors can keep people from
+adopting an energetically balanced form of settlement given prevailing
+conditions. So, there are opportunities for archaeologists to study deviations
+from general equilibrium conditions in addition to factors that shift the
+equilibrium in a given context.
 
-For all this work, however, there is a fundamental issue: the identification of community territories.
-It is relatively straightforward to define the boundaries of aggregated settlements by finding the extent of built space and/or artifactual remains.
-However, to define the community boundary one needs to know the extent of agricultural land used by the residents, and this is often difficult to determine from the village remains themselves [@varien1999].
-In contrast, it is often quite difficult to determine the boundaries of dispersed communities due to gaps in survey coverage or relatively consistent distributions of farmsteads, but once one has determined the community territory it is straightforward to define the associated agricultural land.
+For all this work, however, there is a fundamental issue: the identification of
+community territories. It is relatively straightforward to define the boundaries
+of aggregated settlements by finding the extent of built space and/or
+artifactual remains. However, to define the community boundary one needs to know
+the extent of agricultural land used by the residents, and this is often
+difficult to determine from the village remains themselves [@varien1999]. In
+contrast, it is often quite difficult to determine the boundaries of dispersed
+communities due to gaps in survey coverage or relatively consistent
+distributions of farmsteads, but once one has determined the community territory
+it is straightforward to define the associated agricultural land.
 
-In this paper, we present a method we developed to solve such problems, which enables us to compare the properties and resilience of dispersed vs.
-aggregated communities on the same landscapes.
-We draw on three extensive compilations of archaeological data from different portions of the ancestral Pueblo region of the US Southwest, using the locations of what we call community centers as tethers for defining territories that we infer were used by social communities over extended periods of time [@glowacki_ortman2012; @varien_etal2007].
-In some regions these are sites containing more households than could plausibly be related by reckoned kinship or some form of civic architecture.
-The key feature of these centers is that they represent locations against which travel time distances to other sites can be compared.
+In this paper, we present a method we developed to solve such problems, which
+enables us to compare the properties and resilience of dispersed vs. aggregated
+communities on the same landscapes. We draw on three extensive compilations of
+archaeological data from different portions of the ancestral Pueblo region of
+the US Southwest, using the locations of what we call community centers as
+tethers for defining territories that we infer were used by social communities
+over extended periods of time [@glowacki_ortman2012; @varien_etal2007]. In some
+regions these are sites containing more households than could plausibly be
+related by reckoned kinship or some form of civic architecture. The key feature
+of these centers is that they represent locations against which travel time
+distances to other sites can be compared.
 
-While not strictly required by our algorithm, we make the simplifying assumption that once established, community territories were relatively fixed in space for the duration of their associated sites.
-This allows us to use the distribution of all recorded residences, regardless of their periods of occupation, in defining community territories.
-There are several reasons why we believe this is a reasonable simplification.
-First, there are strong cross-cultural regularities in the time individuals spend in daily travel from and to their residences, so dispersed farming communities tend to be dispersed across the distance associated with this typical travel time, regardless of the number of households involved [@marchetti1994].
-Second, the archaeological record of our study areas supports the idea that community territories did not shrink when their populations aggregated into villages.
-In some cases, farmsteads were converted to field houses, with some building materials being reused elsewhere and evidence for continued limited activity use [@varien1999; @varien1999a; @varien2002].
-In other cases, grid gardens were constructed directly within older rubble mounds by residents of newer, aggregated villages nearby [@gauthier_herhahn2005].
-Third, an analysis that allowed community territories to vary by time step would reduce variation in the density of the resulting units, thus washing out variation in one of the most important social properties one would ideally want to examine.
-Fourth, in areas where there is strong survey coverage, dispersed residences form clusters, often but not always centered on a few larger settlements, and this clustering is apparent even when the dispersed residences are plotted by time step [@schachner2012; @varien2002].
-Fifth, the effects of population growth for dispersed farming communities are not parallel to the effects of population expansion in aggregated settlements.
-In the latter, the area of the settlement must grow somewhat because the residential density is already relatively high.
-But in dispersed settlements, there is often unused land between existing farmsteads that can be settled.
-This, combined with the strong cross-cultural regularity in typical commute times, leads to much greater consistency in the spatial extent of residents in a dispersed community than in the case of an aggregated community.
-Finally, given the inherent incompleteness of the archaeological record, it seems most appropriate to use all remains to define what is effectively the maximum extent of community territories.
-While it was undoubtedly true that individual community territories expanded or contracted over time, the assumption of constant area seems more reasonable than the assumption of constant density.
+While not strictly required by our algorithm, we make the simplifying assumption
+that once established, community territories were relatively fixed in space for
+the duration of their associated sites. This allows us to use the distribution
+of all recorded residences, regardless of their periods of occupation, in
+defining community territories. There are several reasons why we believe this is
+a reasonable simplification. First, there are strong cross-cultural regularities
+in the time individuals spend in daily travel from and to their residences, so
+dispersed farming communities tend to be dispersed across the distance
+associated with this typical travel time, regardless of the number of households
+involved [@marchetti1994]. Second, the archaeological record of our study areas
+supports the idea that community territories did not shrink when their
+populations aggregated into villages. In some cases, farmsteads were converted
+to field houses, with some building materials being reused elsewhere and
+evidence for continued limited activity use [@varien1999; @varien1999a;
+@varien2002]. In other cases, grid gardens were constructed directly within
+older rubble mounds by residents of newer, aggregated villages nearby
+[@gauthier_herhahn2005]. Third, an analysis that allowed community territories
+to vary by time step would reduce variation in the density of the resulting
+units, thus washing out variation in one of the most important social properties
+one would ideally want to examine. Fourth, in areas where there is strong survey
+coverage, dispersed residences form clusters, often but not always centered on a
+few larger settlements, and this clustering is apparent even when the dispersed
+residences are plotted by time step [@schachner2012; @varien2002]. Fifth, the
+effects of population growth for dispersed farming communities are not parallel
+to the effects of population expansion in aggregated settlements. In the latter,
+the area of the settlement must grow somewhat because the residential density is
+already relatively high. But in dispersed settlements, there is often unused
+land between existing farmsteads that can be settled. This, combined with the
+strong cross-cultural regularity in typical commute times, leads to much greater
+consistency in the spatial extent of residents in a dispersed community than in
+the case of an aggregated community. Finally, given the inherent incompleteness
+of the archaeological record, it seems most appropriate to use all remains to
+define what is effectively the maximum extent of community territories. While it
+was undoubtedly true that individual community territories expanded or
+contracted over time, the assumption of constant area seems more reasonable than
+the assumption of constant density.
 
@@
 
-The datasets we apply our algorithm to come from three regions in the US Southwest: Central Mesa Verde (CMV), the Northern Rio Grande (NRG), and the Cibola or Zuni region (CIB) (see @fig-overview).
-The CMV and NRG datasets derive from Phase II of the Village Ecodynamics Project [@ortman2016; @schwindt_etal2016].
-The CIB data derive from data syntheses produced by Schachner and Peeples [@peeples2018; @peeples_schachner2012; @schachner2012].
-All of these data are now included in cyberSW [@mills_etal2020].
-For summaries of the archaeological and environmental context in each region, see @tbl-regions.
+The datasets we apply our algorithm to come from three regions in the US
+Southwest: Central Mesa Verde (CMV), the Northern Rio Grande (NRG), and the
+Cibola or Zuni region (CIB) (see @fig-labeled-overview). The CMV and NRG
+datasets derive from Phase II of the Village Ecodynamics Project [@ortman2016;
+@schwindt_etal2016]. The CIB data derive from data syntheses produced by
+Schachner and Peeples [@peeples2018; @peeples_schachner2012; @schachner2012].
+All of these data are now included in cyberSW [@mills_etal2020]. The
+distribution of farms and centers is shown in @fig-overview. For summaries of
+the archaeological and environmental context in each region, see @tbl-regions.
 
-![
-Overview map showing the locations of farmsteads (black dots) and community centers (red triangles) in Central Mesa Verde, Northern Rio Grande, and Cibola.
-](../figures/overview-maps.png){#fig-overview layout-align="center" width="100%"}
+::: {#fig-labeled-overview fig-align="center"}
+![](../figures/labeled-overview-map.png){width="5.2in"}
 
-The CMV area encompasses the southwestern corner of Colorado, from the Mesa Verde cuesta in the south and east to the Dolores River in the north, and the Utah state line in the west.
-The elevation ranges from roughly 1,399.8 m to 3,032.8 m, with an average elevation of 1,999.0 m ($\sigma$ = 276.5 m).
-SKOPE [@bocinsky_etal2022] estimates suggest annual precipitation in the CMV study area has averaged around 39.5 cm ($\sigma$ = 9.1 cm) over the last 2,000 years, and it puts summer growing degree days for maize around 2,390.1 °F ($\sigma$ = 97.9 °F).
-Across most areas in the region, it is believed that direct precipitation was sufficiently high to support dry farming in most years [@bocinsky_kohler2014; @bocinsky_etal2016].
+Overview map showing the locations of the three study areas: Central Mesa Verde,
+Northern Rio Grande, and Cibola. Prominent towns, administrative units, and
+landforms are also labeled to help orient the reader.
+:::
 
-The NRG study area in north central New Mexico includes basically all of the Rio Grande valley between the Jemez Mountains to the west and the southernmost mountains of the Sangre De Cristo range to the east, with the Rio Grande itself running along the western edge of the valley at the foot of the Pajarito Plateau.
-The elevation ranges from roughly 1,587.9 m to 3,836.2 m, with an average elevation of 2,220.0 m ($\sigma$ = 371.3 m).
-SKOPE estimates the average annual precipitation in the NRG over the last 2,000 years to be around 40.6 cm ($\sigma$ = 8.2 cm), and it puts summer growing degree days for maize around 2,154.4 °F ($\sigma$ = 102.1 °F).
-It is argued that direct rainfall on the Pajarito Plateau made dry farming possible there, but in the valley farming required floodwater irrigation along watercourses [@duwe_anschuetz2013; @bocinsky_kohler2014; @bocinsky_etal2016; @ortman2016a].
+The CMV area encompasses the southwestern corner of Colorado, from Mesa Verde
+National Park in the southeast to the Dolores River in the north, and the Utah
+state line in the west. The elevation ranges from roughly 1,399.8 m to 3,032.8
+m, with an average elevation of 1,999.0 m ($\sigma$ = 276.5 m). SKOPE
+[@bocinsky_etal2022] estimates suggest annual precipitation in the CMV study
+area has averaged around 39.5 cm ($\sigma$ = 9.1 cm) over the last 2,000 years,
+and it puts summer growing degree days for maize around 2,390.1 °F ($\sigma$ =
+97.9 °F). Across most areas in the region, it is believed that direct
+precipitation was sufficiently high to support dry farming in most years
+[@bocinsky_kohler2014; @bocinsky_etal2016].
 
-The CIB region straddles the southern slope of the Colorado Plateau along the border between Arizona and New Mexico.
-Although this is a rugged terrain by any reasonable standard, it is notably less so than the other study areas.
-The elevation ranges from roughly 1,771.3 m to 2,777.1 m, with an average elevation of 2,145.9 m ($\sigma$ = 160.0 m).
-SKOPE estimates suggest annual precipitation in the CIB area over the last 2,000 years has averaged around 34.3 cm ($\sigma$ = 7.6 cm), and it puts summer growing degree days for maize around 2,317.3 °F ($\sigma$ = 93.8 °F).
-However, no area in the region receives sufficient direct rainfall to support dry farming, so some form of water management was required [@kintigh1985; @muenchrath_etal2002].
+The NRG study area in north central New Mexico includes basically all of the Rio
+Grande valley between the Jemez Mountains to the west and the southernmost
+mountains of the Sangre De Cristos to the east, with the Rio Grande itself
+running from the northeast down to the western edge of the valley at the foot of
+the Pajarito Plateau around Bandelier National Monument. The elevation ranges
+from roughly 1,587.9 m to 3,836.2 m, with an average elevation of 2,220.0 m
+($\sigma$ = 371.3 m). SKOPE estimates the average annual precipitation in the
+NRG over the last 2,000 years to be around 40.6 cm ($\sigma$ = 8.2 cm), and it
+puts summer growing degree days for maize around 2,154.4 °F ($\sigma$ = 102.1
+°F). It is argued that direct rainfall at Bandelier NM made dry farming possible
+there, but in the valley farming required floodwater irrigation along
+watercourses [@duwe_anschuetz2013; @bocinsky_kohler2014; @bocinsky_etal2016;
+@ortman2016a].
+
+The CIB region straddles the southern slope of the Colorado Plateau along the
+border between Arizona and New Mexico, from El Morro National Monument in the
+east to the centrally located Zuni Pueblo to the confluence of Jaralosa Draw and
+the Zuni River in the southwest. Although this is a rugged terrain by any
+reasonable standard, it is notably less so than the other study areas. The
+elevation ranges from roughly 1,771.3 m to 2,777.1 m, with an average elevation
+of 2,145.9 m ($\sigma$ = 160.0 m). SKOPE estimates suggest annual precipitation
+in the CIB area over the last 2,000 years has averaged around 34.3 cm ($\sigma$
+= 7.6 cm), and it puts summer growing degree days for maize around 2,317.3 °F
+($\sigma$ = 93.8 °F). However, no area in the region receives sufficient direct
+rainfall to support dry farming, so some form of water management was required
+[@kintigh1985; @muenchrath_etal2002].
 
 ::: {#tbl-regions}
-
 ```{=latex}
@@
 \begin{tabular}{lllll}
-    & \textbf{Area [km2]} & \textbf{Dates [CE]} & \textbf{Archaeology} & \textbf{Environment} \\ 
+    & 
+\textbf{Area [km2]} & 
+\textbf{Dates [CE]} & 
+\textbf{Archaeology} & 
+\textbf{Environment} 
+  \\ 
   \hline  
   \noalign{\vskip2mm}  
-  CMV & 4,556.4             & 725-1280            & \begin{tabular}[c]{@{}r@{}}Rooms: \hfill 67,746\\ Farms: \hfill 5,415\\ Centers: \hfill 512\end{tabular} & \begin{tabular}[r]{@{}r@{}}Precipitation: \hfill 39.5 cm ($\sigma$: 9.1)\\ Maize GDD: \hfill 2,390.1 $^{\circ}$F ($\sigma$: 97.9)\\ Elevation: \hfill 1,999.0 m ($\sigma$: 276.5)\end{tabular} \\[5mm]
+CMV & 
+4,566.4 & 
+725-1280 & 
+\begin{tabular}[c]{@{}r@{}}
+  Rooms: \hfill 68,940\\ 
+  Farms: \hfill 5,272\\ 
+  Centers: \hfill 170
+\end{tabular} & 
+\begin{tabular}[r]{@{}r@{}}
+  Precipitation: \hfill 39.5 cm ($\sigma$: 9.1)\\ 
+  Maize GDD: \hfill 2,390.1 $^{\circ}$F ($\sigma$: 97.9)\\ 
+  Elevation: \hfill 1,999.0 m ($\sigma$: 276.5)
+\end{tabular} 
+  \\[5mm]  
   \hline  
   \noalign{\vskip2mm}  
-  NRG & 6,958.0             & 900-1550            & \begin{tabular}[c]{@{}r@{}}Rooms: \hfill 46,226\\ Farms: \hfill 2,316\\ Centers: \hfill 181\end{tabular} & \begin{tabular}[r]{@{}r@{}}Precipitation: \hfill 40.6 cm ($\sigma$: 8.2)\\ Maize GDD: \hfill 2,154.4 $^{\circ}$F ($\sigma$: 102.1)\\ Elevation: \hfill 2,222.0 m ($\sigma$: 371.3)\end{tabular} \\[5mm]
+NRG & 
+6,958.0 & 
+900-1550 & 
+\begin{tabular}[c]{@{}r@{}}
+  Rooms: \hfill 46,226\\ 
+  Farms: \hfill 2,316\\ 
+  Centers: \hfill 181
+\end{tabular} & 
+\begin{tabular}[r]{@{}r@{}}
+  Precipitation: \hfill 40.6 cm ($\sigma$: 8.2)\\ 
+  Maize GDD: \hfill 2,154.4 $^{\circ}$F ($\sigma$: 102.1)\\ 
+  Elevation: \hfill 2,222.0 m ($\sigma$: 371.3)
+\end{tabular} 
+  \\[5mm]  
   \hline  
   \noalign{\vskip2mm}  
-  CIB & 7,424.1             & 700-1540            & \begin{tabular}[c]{@{}r@{}}Rooms: \hfill 27,234\\ Farms: \hfill 713\\ Centers: \hfill 78\end{tabular}    & \begin{tabular}[r]{@{}r@{}}Precipitation: \hfill 34.3 cm ($\sigma$: 7.6)\\ Maize GDD: \hfill 2,317.3 $^{\circ}$F ($\sigma$: 93.8)\\ Elevation: \hfill 2,145.9 m ($\sigma$: 160.0)\end{tabular} \\[5mm]
+CIB & 
+7,424.1 & 
+700-1540 & 
+\begin{tabular}[c]{@{}r@{}}
+  Rooms: \hfill 26,818\\ 
+  Farms: \hfill 713\\ 
+  Centers: \hfill 77
+\end{tabular} & 
+\begin{tabular}[r]{@{}r@{}}
+  Precipitation: \hfill 34.3 cm ($\sigma$: 7.6)\\ 
+  Maize GDD: \hfill 2,317.3 $^{\circ}$F ($\sigma$: 93.8)\\ 
+  Elevation: \hfill 2,145.9 m ($\sigma$: 160.0)
+\end{tabular} 
+  \\[5mm] 
   \hline
@@
 \end{singlespace}
-
 ```
-
 Regional Summaries
-
 :::
@@
 
-Our clustering algorithm draws inspiration from the procedure used by the United States Office of Management and Budget to define Core Based Statistical Areas (CBSA) [@usomb2021].
-We start by defining a core area, in this case a community center location.
-We then associate neighboring residential sites or farmsteads with this center based on their commute times, and we merge core areas into larger agglomerations based on the proportion of their populations that they share.
-Our method may be described as a guided density-based clustering algorithm in that it does not rely on a random set of points when initialized.
-Instead, the selection of core areas is guided by archaeological data and regional expertise.
-We think this is a key argument in its favor.
+Our clustering algorithm draws inspiration from the procedure used by the United
+States Office of Management and Budget to define Core Based Statistical Areas
+(CBSA) [@usomb2021]. We start by defining a core area, in this case a community
+center location. We then merge core areas into larger agglomerations based on
+the proportion of their populations that they share. Our method may be described
+as a guided density-based clustering algorithm in that it does not rely on a
+random set of points when initialized. Instead, the selection of core areas is
+guided by archaeological data and regional expertise. We think this is a key
+argument in its favor.
 
-From an archaeological perspective, one can think of our algorithm as combining insights from @varien1999 and @reese_etal2019, specifically Varien’s idea of a community catchment, which is an isochrone defined by a uniform commute time in all directions from a community center, and Reese’s suggestion that a spatial or geographic community can be identified by grouping individuals in terms of the commute distances between them.
-Our way of combining these ideas is to associate farm sites with specific community centers in terms of their commute distances and then draw the community catchment around those associated farms.
-We also share with them and others [@lipe_hegmon1989; @mahoney_etal2000; @peterson_drennan2005] a focus on "geographic" as opposed to "purely social" communities; that is, communities constrained in space and comprised of individuals that can be expected to interact on a regular, even daily, basis.
+From an archaeological perspective, one can think of our algorithm as combining
+insights from @varien1999 and @reese_etal2019, specifically Varien's idea of a
+community catchment, which is an isochrone defined by a uniform commute time in
+all directions from a community center, and Reese's suggestion that a spatial or
+geographic community can be identified by grouping individuals in terms of the
+commute distances between them. Our way of combining these ideas is to associate
+farm sites with specific community centers in terms of their commute distances
+and then draw the community catchment around those associated farms. We also
+share with them and others [@lipe_hegmon1989; @peterson_drennan2005;
+@murdock1949] a focus on "geographic" as opposed to "purely social" communities;
+that is, communities constrained in space and comprised of individuals that can
+be expected to interact on a regular, even daily, basis.
 
-In broad outline, our proposed community detection algorithm has the following steps:
+An important difference for our algorithm is that it incorporates information
+about the total mass or population size around each center, so it is at least
+conceptually similar to the classic Xtent model proposed by [@renfrew_level1979;
+see also @ducke_kroefges2008], as well as the more widely known gravity model
+[@isard1954]. In broad outline, it has the following steps:
 
 1.  Identify community centers
-2. Identify farms within a commute time threshold ($D{\text -}max$)
-3. Join communities to their overlapping neighbors ($P$, $D{\text -}join$)
-4. Exclude farms closer to another community center (tie-breaking step)
+2.  Join farms to their nearest community center
+3.  Exclude farms beyond a commute time threshold from all centers
+    ($D{\text -}max$)
+4.  Join centers based on their overlapping populations ($P$, $D{\text -}join$)
 5.  Draw smallest concave hull encompassing all farms, centers, and paths
 
-\noindent After step 4, we also apply a filter to remove communities that have less than a minimal number of dispersed farmsteads, specifically four.
-Although this is a somewhat arbitrary threshold, we note that three vertices are the bare minimum required to define a polygon and measure its area.
-By removing small communities, we also minimize potential underestimates of total arable land area.
-In addition, our larger ambition is to understand processes occurring within dispersed farming communities, not necessarily to define all the dispersed communities that once existed in an area.
-For community centers with few or no farms, there is no meaningful dispersed farming community to examine, regardless of whether this is due to limited survey coverage or actual past behavior, hence our desire to exclude them.
+\noindent After step 4, we also apply a filter to remove communities that have
+less than a minimal number of dispersed farmsteads, specifically four. Although
+this is a somewhat arbitrary threshold, we note that three vertices are the bare
+minimum required to define a polygon and measure its area. By removing small
+communities, we also minimize potential underestimates of total arable land
+area. In addition, our larger ambition is to understand processes occurring
+within dispersed farming communities, not necessarily to define all the
+dispersed communities that once existed in an area. For community centers with
+few or no farms, there is no meaningful dispersed farming community to examine,
+regardless of whether this is due to limited survey coverage or actual past
+behavior, hence our desire to exclude them.
 
-The primary tuning parameters in our algorithm are $P$, the proportion of a community population used to join communities; $D{\text -}join$, the commute time required to calculate P; and $D{\text -}max$, the commute time used to exclude distant farms.
-These are described in more detail below.
+The primary tuning parameters in our algorithm are $P$, the proportion of a
+community population used to join communities; $D{\text -}join$, the commute
+time required to calculate P; and $D{\text -}max$, the commute time used to
+exclude distant farms. While values for these variables were carefully chosen
+based on theory and empirical research, we recognize that they are, at the end
+of the day, arbitrary selections, so we include some limited sensitivity
+analysis to show how changes to these parameters affect the algorithm (see
+@fig-sensitivity). This involves changing the value of a focal parameter while
+holding the other parameters at their default values, which we describe below.
+
+::: {#fig-overview fig-align="center"}
+![](../figures/overview-maps.png){width="6.5in"}
+
+Overview map showing the locations of farmsteads (black dots) and community
+centers (red triangles) in Central Mesa Verde, Northern Rio Grande, and Cibola.
+:::
 
@@
 
-**Rule:** Any site, whether residential or otherwise, with known and persistent interaction between unrelated individuals should count as a community center.
+**Rule:** Any site, whether residential or otherwise, with known and persistent
+interaction between unrelated individuals should count as a community center.
 
-Why persistent interactions?
-Intuitively, a community center is a place where individuals from different households cross paths with the intention to interact.
-A simple way to demonstrate that intention is to show that individuals would still be willing to pay some non-negligible cost to visit those centers under various counterfactual changes to local conditions.
-This serves to distinguish them from merely accidental exchanges or chance encounters.
-Of course, there can be some path-dependence to this process, with locations of ephemeral or one-off encounters coming to be locations of persistent interaction.
-The point is that the location will come to be a community center so long as individuals continue to visit that location despite the underlying variance in their circumstances.
+Why persistent interactions? Intuitively, a community center is a place where
+individuals from different households cross paths with the intention to interact
+[@glowacki_ortman2012; @varien1999; @berrey_etal2021]. A simple way to
+demonstrate that intention is to show that individuals would still be willing to
+pay some non-negligible cost to visit those centers under various counterfactual
+changes to local conditions. This serves to distinguish them from merely
+accidental exchanges or chance encounters. Of course, there can be some
+path-dependence to this process, with locations of ephemeral or one-off
+encounters coming to be locations of persistent interaction. The point is that
+the location will come to be a community center so long as individuals continue
+to visit that location despite the underlying variability in their
+circumstances.
 
-Why unrelated individuals?
-The simple answer is that a community is not a family.
-As a unit of social organization, it does share something in common with a family, namely, a set of shared interests; but unlike in a family, those shared interests form only a small part of a much larger set of potentially conflicting goals and desires.
-As a consequence, communities are more susceptible to problems with coordination or collective action.
-On the other hand, the mere existence of a community center implies some level of coordination - even if that coordination is just about the location of the community center itself!
-So, community members share less in common than families, but more in common than a purely random sample of the larger population.
+Why unrelated individuals? The simple answer is that a community is not a
+family. As a unit of social organization, it does share something in common with
+a family, namely, a set of shared interests; but unlike in a family, those
+shared interests form only a small part of a much larger set of potentially
+conflicting goals and desires. As a consequence, communities are more
+susceptible to problems of coordination or collective action [@smith2003;
+@smith2010]. On the other hand, the mere existence of a community center implies
+some level of coordination - even if that coordination is just about the
+location of the community center itself! So, community members share less in
+common than families, but more in common than a purely random sample of the
+larger population.
 
-We can summarize these ideas by noting that persistent interactions at a location make that location a *center*, and persistent interactions between unrelated individuals at that location make it a *community* center.
-To make these considerations more useful for defining community boundaries, however, we need to operationalize them, to make them applicable to specific archaeological contexts, and that requires a deeper understanding of the archaeology in those contexts.
-For the two regions represented by the VEP II dataset, we rely on two different but related sets of criteria to identify community centers.
-In the Central Mesa Verde area, a site is considered a community center if it includes civic architecture, specifically a great kiva or great house, or eight or more pit structures (small kivas) representing eight or more households.
-In the Northern Rio Grande, a site counts as a community center if it is classified as a village (a site with 50 or more rooms) or town (a site with 500 or more rooms).
-In the Cibola region, a community center includes at least 50 rooms or civic architecture.
-The large number of households in these regions are thought to be too large to be plausibly related by reckoned kinship.
+We can summarize these ideas by noting that persistent interactions at a
+location make that location a *center*, and persistent interactions between
+unrelated individuals at that location make it a *community* center
+[@varien_potter2008; @gilpin2003; @stone2016; @peterson_drennan2005]. To make
+these considerations more useful for defining community boundaries, however, we
+need to operationalize them, to make them applicable to specific archaeological
+contexts, and that requires a deeper understanding of the archaeology in those
+contexts. For given differences in population density and settlement patterning,
+the frequency and intensity of interaction between individuals at centers is
+likely to vary by region, so it is reasonable to expect that community centers
+themselves (at least in so far as they play the role of centers) should vary in
+their size and composition across regions, too.
 
-## Join farms within a commute time
+For the two regions represented by the VEP II dataset, we rely on two different
+but related sets of criteria to identify community centers. In the Central Mesa
+Verde area, a site is considered a community center if it includes civic
+architecture, specifically a great kiva or great house, or eight or more pit
+structures (small kivas) representing eight or more households. In the Northern
+Rio Grande, a site counts as a community center if it is classified as a village
+(a site with 50 or more rooms) or town (a site with 500 or more rooms). In the
+Cibola region, a community center includes at least 50 rooms or civic
+architecture. The large number of households in these regions are thought to be
+too large to be plausibly related by reckoned kinship.
 
-**Rule:** For each center, join all farms within commute time, $D{\text -}max$.
+## Join farms to their nearest community center
 
-Following @reese_etal2019, we define nearness or proximity in terms of commute time rather than geographic distance.
-This is done for the obvious reason that the primary opportunity cost associated with daily pedestrian movement is the time spent walking, during which other productive activities cannot take place.
-There is also a much stronger cross-cultural regularity in the length of time individuals devote to commuting than the distance covered, as the latter is a function of the speed of movement, which is influenced by transport technology [@marchetti1994].
-As a result, both the frequency and intensity of interaction at community centers should decay with commute time.
-The longer it takes to get to a community center, the less frequent and less intense the interactions at that center should be [@varien1999; @peterson_drennan2005].
+**Rule:** For each farm, assign it to the community of its nearest community
+center.
 
-While different constraints will naturally arise for different modes of transportation, leading to differences in both the routes taken and the speed and distance covered, for the populations considered here the only available mode of transportation was pedestrian.
-The biggest obstacle to interaction was thus the landscape itself, in particular its topography - intuitively, the steeper the uphill slope, the slower the hiking speed, the longer the commute time.
-So, all else being equal, individuals at a farm that is equidistant (in terms of geographic distance) between two community centers should prefer traveling to the one separated by less rugged terrain.
+Following @reese_etal2019, we define nearness or proximity in terms of commute
+time rather than linear geographic distance. This is done for the obvious reason
+that the primary opportunity cost associated with daily pedestrian movement is
+the time spent walking, during which other productive activities cannot take
+place. There is also a much stronger cross-cultural regularity in the length of
+time individuals devote to commuting than the distance covered, as the latter is
+a function of the speed of movement, which is influenced by transport technology
+[@marchetti1994]. As a result, both the frequency and intensity of interaction
+at community centers should decay with commute time. The longer it takes to get
+to a community center, the less frequent and less intense the interactions at
+that center should be [@varien1999; @peterson_drennan2005].
 
-The concept of commute time is a critical variable in many places in our algorithm, but as a first approximation, we incorporate the idea by setting a commute time threshold, $D{\text -}max$.
-This variable, in effect, defines an isochrone around each community center, what @varien1999 calls a "catchment." All farm sites that fall within that isochrone get associated with its community center.
-This threshold also serves as a filter, as we drop all farm sites that are farther than that commute time to any center.
-Doing so helps to minimize potential overestimates of total arable land area for each community.
-Here, we have defined $D{\text -}max$ as one hour.
-This is about twice the median daily commute time observed in cross-cultural studies and it includes roughly 95% of the farmsteads in each region (see @fig-commutes).
+While different constraints will naturally arise for different modes of
+transportation, leading to differences in both the routes taken and the speed
+and distance covered, for the populations considered here the only available
+mode of transportation was pedestrian. The biggest obstacle to interaction was
+thus the landscape itself, in particular its topography - intuitively, the
+steeper the uphill slope, the slower the hiking speed, the longer the commute
+time. So, all else being equal, individuals at a farm that is equidistant (in
+terms of geographic distance) between two community centers should prefer
+traveling to the one separated by less rugged terrain.
 
-For details of how we estimate commute times over costly terrains, see “Least-cost path analysis,” below.
+Commute time is a critical variable in many places in our algorithm, but as a
+first approximation, we incorporate the idea by simply joining farms to the
+community of their nearest center, meaning the center with the shortest commute
+time from the farm. For details of how we estimate commute times over costly
+terrains, see "Least-cost path analysis," below.
+
+## Exclude distant farms
+
+**Rule:** Exclude all farms farther than commute time, $D{\text -}max$, from
+their nearest center.
+
+The commute time threshold, $D{\text -}max$, defines an isochrone around each
+community center. All farm sites that fall outside that isochrone are considered
+outliers and dropped from the analysis. Doing this helps to minimize potential
+overestimates of total arable land area for each community. This is owing to the
+fact that distant farms on average add more area to the community polygon than
+nearer farms. Here, we have defined $D{\text -}max$ as one hour. This is about
+twice the median daily commute time observed in cross-cultural studies and it
+includes roughly 95% of the farmsteads in each region (see @fig-commutes).
 
@@
 
-**Rule:** For centers $C_1$ and $C_2$ with nearby farm populations $N_1 ≤ N_2$, if $P \cdot N_1$ are within distance $D{\text -}join$ of $C_1$ and $C_2$, then join $C_1$ to $C_2$.
+**Rule:** For any two centers $c$ and $d$ with populations $N_{c} < N_{d}$, if
+$P \cdot N_{c}$ is within distance $D{\text -}join$ of $d$, then $c$ is part of
+the same community as $d$.
 
-The larger and more dense farming populations become, the harder it gets to tell them apart and, more importantly, the harder it gets to justify keeping them apart, so we need to articulate a rule that specifies when two communities should be merged into one.
-For CBSAs, that decision hinges on a distinction between central and outlying counties, or counties where some fraction of the population live in an urban core and counties where some fraction commute to and from a central county on a regular basis.
-Basically, if the central counties of one CBSA qualify as the outlying counties of another CBSA, then the two CBSAs are merged.
-The intuition here is that when rates of interaction and exchange between CBSAs are roughly equal to rates within a CBSA, the CBSAs should be merged (USOMB 2021).
+The larger and more dense farming populations become, the harder it gets to tell
+them apart and, more importantly, the harder it gets to justify keeping them
+apart, so we need to articulate a rule that specifies when two communities
+should be merged into one. For CBSAs, that decision hinges on a distinction
+between central and outlying counties, or counties where some fraction of the
+population live in an urban core and counties where some fraction commute to and
+from a central county on a regular basis. Basically, if the central counties of
+one CBSA qualify as the outlying counties of another CBSA, then the two CBSAs
+are merged. The intuition here is that when rates of interaction and exchange
+between CBSAs are roughly equal to rates within a CBSA, the CBSAs should be
+merged [@usomb2021].
 
-We carry that intuition over to our merger rule for dispersed maize farming communities, albeit with some modification to account for the fact that we are working with point locations rather than administrative boundaries.
-We assume that if some proportion, $P$, of the population related to one community center lives within a certain commute time, $D{\text -}join$, of another community center, that those two community centers are effectively parts of the same dispersed community.
-We do this in the direction of letting larger communities absorb smaller ones.
+We carry that intuition over to our merger rule for dispersed maize farming
+communities, albeit with some modification to account for the fact that we are
+working with point locations rather than administrative boundaries. We assume
+that if some proportion, $P$, of the population related to one community center
+lives within a certain commute time, $D{\text -}join$, of another community
+center, that those two community centers are effectively parts of the same
+dispersed community. We do this in the direction of letting larger communities
+absorb smaller ones.
 
-More precisely, the rule requires that $P \cdot N_1$ be within $D{\text -}join$ of *both* $C_1$ and $C_2$.
-This promotes joins when $C_1$ and $C_2$ are themselves in close proximity to each other, as the farther apart they are, the smaller the number of sites that overlap within $D{\text -}join$.
-It also tends to encourage joining two communities when a substantial proportion of the population lives between the two centers.
-In either case, the result is what @varien1999 sometimes refers to as a "macro-community" or "multi-community cluster," meaning a community organized around multiple community centers.
+Here we operationalize the population size $N$ of center $c$ using the
+well-established rule of thumb in US Southwest archaeology that each surface
+room in a settlement represents a single person, leading to a one to one
+conversion from rooms to residents [@duwe_etal2016; @kintigh1985; @lekson1989;
+@lipe1989; @ortman2016]. In this case, we use the catchment room count, meaning
+the number of rooms within $D{\text -}join$ of $c$, defined as:
 
-The rule was partly inspired by the density-based clustering algorithm known as DBSCAN [@ester_etal1996], as we rely on a moving window (an isochrone, in this case) set around each center, and look for some measure of density to guide join-decisions.
-As noted above, however, our method does not rely on a random initialization of points, but rather targets community centers.
-It is also hierarchical in the sense that it exploits additional structure in our data, namely the distinction between community centers and farm sites.
+$$N_{c} = \sum_{i=1}^{S} R_{i} \cdot I(t_{ic} \leq D{\text -}join)$$
 
-As we cannot measure population directly, however, we use information we have about farm sites, specifically estimates of their room counts, as a proxy.
-For many sites in the CMV, pit structures or small kivas are more reliably identified than surface rooms.
-Fortunately, each pit structure is typically the central roofed space of a unit pueblo residence that includes an average of six surface rooms [@adler1990], so we simply multiply the number of pit structures by six to obtain room estimates for each site.
-CIB and NRG sites all have room estimates, so no additional assumptions are made about them.
-Finally, we follow the well-established rule of thumb in US Southwest archaeology that each surface room in a settlement represents a single person, leading to a one to one conversion from rooms to residents [@duwe_etal2016; @kintigh1985; @lekson1989; @lipe1989; @ortman2016].
+\noindent for all sites $S$ (including both farms and centers), with $R_{i}$
+being the room count at site $i$, $t_{ic}$ the travel time from $i$ to $c$,
+$D{\text -}join$ the threshold travel time to a center, and $I$ the indicator
+function that is 1 if $t \leq D{\text -}join$ and 0 otherwise. Note that $S$
+includes $c$ and that $t_{cc} = 0$ (the travel time from a center to the same
+center is precisely 0), so $R_{c}$ is always included in $N_{c}$.
 
-For this analysis, we specify $D{\text -}join$ to be one half hour and $P$ to be 0.8.
-We acknowledge that these values are somewhat arbitrary.
-In their defense, they set an extremely high bar, so joining centers that satisfy the rule seems reasonable.
+The catchment room count within $D{\text -}join$ of two centers $c$ and $d$ is
+then
 
-## Exclude farms closer to another community center
+$$N_{cd} = \sum_{i=1}^{S} R_{i} \cdot I(t_{ic} \leq D{\text -}join) \cdot I(t_{id} \leq D{\text -}join)$$
 
-**Rule:** For each farm that falls within $D{\text -}max$ of multiple community centers, associate it with the community center to which it has a shorter commute time.
+\noindent and
 
-On occasion a farm may fall within $D{\text -}max$ of multiple community centers that are not joined into a single community, which makes them members of multiple communities.
-For the dispersed farming communities we aim to identify, this would imply that multiple communities could simultaneously farm the same plot of land, which is unrealistic, so we need a rule for breaking ties.
-Because we are already relying so heavily on commute time, the simplest solution is to associate a farm that falls within $D{\text -}max$ of multiple community centers with the center and, thus, the community that it is closer to, and this is what we have elected to do.
-We can justify this step on similar grounds as well, with the safe assumption that individuals should be more likely to interact the nearer they are in time and space.
-We also note that this tie-breaking rule ensures that community polygons are continuous, that they do not contain any holes, which would be quite odd considering these community boundaries are supposed to represent the spatial extent of their arable land.
-It also encourages some separation between communities, preventing them from overlapping to any significant degree.
+$$P_{cd} = N_{cd}/N_{c}$$
+
+\noindent so we could also state our join rule as: if $P_{cd} \geq \alpha$ for
+some critical threshold $\alpha$, then $c$ is part of the same community as $d$.
+This promotes joins when $c$ and $d$ are themselves in close proximity to each
+other, as the farther apart they are, the smaller the number of sites that
+overlap within $D{\text -}join$. It also tends to encourage joining two
+communities when a substantial proportion of the population lives between the
+two centers. In either case, the result is what @varien1999 sometimes refers to
+as a "macro-community" or "multi-community cluster," meaning a community
+organized around multiple community centers.
+
+The rule was partly inspired by the density-based clustering algorithm known as
+DBSCAN [@ester_etal1996], as we rely on a moving window (an isochrone, in this
+case) set around each center, and look for some measure of density to guide join
+decisions. As noted above, however, our method does not rely on a random
+initialization of points, but rather targets community centers. It is also
+hierarchical in the sense that it exploits additional structure in our data,
+namely the distinction between community centers and farm sites.
+
+An important obstacle to estimating population is that room counts cannot always
+be directly measured. For many sites in the CMV, pit structures or small kivas
+are more reliably identified than surface rooms. Fortunately, each pit structure
+is typically the central roofed space of a unit pueblo residence that includes
+an average of six surface rooms [@adler1990], so we simply multiply the number
+of pit structures by six to obtain room estimates for each site. CIB and NRG
+sites all have room estimates, so no additional assumptions are made about them.
+
+For this analysis, we specify $D{\text -}join$ to be one half hour and $P$ to be
+0.8. We acknowledge that these values are somewhat arbitrary. In their defense,
+they set an extremely high bar, so joining centers that satisfy the rule seems
+reasonable.
 
@@
 
-**Rule:** For each community, draw a polygon that encompasses all farms, community centers, and commute paths using a concave hull.
+**Rule:** For each community, draw a polygon that encompasses all farms,
+community centers, and commute paths using a concave hull.
 
-So far, we have only grouped site points into clusters or communities.
-The goal, however, is to define their spatial extent.
-That means, for each set of site points, that we need some way of drawing a polygon that encompasses all of them.
-This can be done in a number of ways.
-The simplest strategy would be to find the centroid of the set of points and draw a circle with the smallest radius that still includes them all.
-This would be utterly arbitrary, however, and frankly unrealistic, not representing any sort of meaningful boundary given the landscape.
-An alternative would be to use the convex hull of the set of points, but this tends to exaggerate the total area of the community, especially when the distribution of points is suggestive of a concave shape, like the letter ‘C’.
-A concave hull would handle peculiar shapes, but it would also introduce the opposite problem, restricting the area of the community to an unreasonable degree.
-Our somewhat brute force way of threading this needle is to identify paths with shortest commute times between every pair of farms in a community, extract the vertices from those paths, and then incorporate those into the set of points used to define the concave hull.
-The result is a polygon that is concave as well in the dimension of time, meaning the paths with shortest commute times between any two farms never leaves the community.
+So far, we have only grouped site points into clusters or communities. The goal,
+however, is to define their spatial extent. That means, for each set of site
+points, we need some way of drawing a polygon that encompasses all of them. This
+can be done in a number of ways. The simplest strategy would be to find the
+centroid of the set of points and draw a circle with the smallest radius that
+still includes them all. This would be utterly arbitrary, however, and frankly
+unrealistic, not representing any sort of meaningful boundary given the
+landscape. An alternative would be to use the convex hull of the set of points,
+but this tends to exaggerate the total area of the community, especially when
+the distribution of points is suggestive of a concave shape, like the letter
+'C'. A concave hull would handle peculiar shapes, but it would also introduce
+the opposite problem, restricting the area of the community to an unreasonable
+degree. Our somewhat brute force way of threading this needle is to identify
+paths with shortest commute times between the outer most farms in a community,
+extract the vertices from those paths, and then incorporate those into the set
+of points used to define the concave hull. The result is a polygon that is
+concave as well in the dimension of time, meaning the paths with shortest
+commute times between any two farms never leave the community.
 
-In some cases, this may lead to overlapping community boundaries, though the amount of overlap should be minimal given the way that we cluster points, particularly when there are many densely distributed points.
-To some extent, of course, this sort of overlap is both unavoidable and expected as even in the modern world there are disagreements about the exact locations of legal and political boundaries.
+In some cases, this may lead to overlapping community boundaries, though the
+amount of overlap should be minimal given the way that we cluster points,
+particularly when there are many densely distributed points. To some extent, of
+course, this sort of overlap is both unavoidable and expected as even in the
+modern world there are disagreements about the exact locations of legal and
+political boundaries.
 
-It should be noted, too, that we also dilate each concave hull, briefly expanding it with a positive buffer, then shrinking it with a negative buffer having a slightly smaller size.
-This should have only a minimal effect on the total area of each polygon and also serves to smooth out noise along the edges.
+As a final gloss on community boundaries, we also dilate each concave hull,
+briefly expanding it with a positive buffer, then shrinking it with a negative
+buffer having a slightly smaller size. This should have only a minimal effect on
+the total area of each polygon and also serves to smooth out noise along the
+edges.
 
@@
 
-The key to our path analysis is to take information about the terrain in our study regions and convert it first into meaningful estimates of hiking speed and then into meaningful estimates of travel time.
-Fortunately, attempts to model hiking speed across a range of slopes have advanced considerably in recent years, most notably in @campbell_etal2019.
-Drawing on a large, crowd-sourced fitness tracker data set consisting of approximately 420,000 individual hikes, jogs, and trail runs from nearly 30,000 individuals in Salt Lake City, UT, @campbell_etal2019 offer what is easily the most empirically robust estimate of hiking speed to date.
-An interesting feature of their analysis - though one less important for the current study - is its sensitivity to variance in the physical and athletic prowess of individuals, which allows for estimates of different speeds at a given slope for different deciles of the sample.
-For simplicity, our application of Campbell’s hiking function relies on coefficients from the first decile of individuals in their sample (the slowest ten percent of hikers) and assumes that everyone in the areas of our analysis were largely equal in their walking speed and endurance.
-This makes their walking speeds comparable to, but not the same as, what would be estimated by Tobler’s hiking function [@tobler1993].
+The key to our path analysis is to take information about the terrain in our
+study regions and convert it first into meaningful estimates of hiking speed and
+then into meaningful estimates of travel time. Fortunately, attempts to model
+hiking speed across a range of slopes have advanced considerably in recent
+years, most notably in @campbell_etal2019. Drawing on a large, crowd-sourced
+fitness tracker data set consisting of approximately 420,000 individual hikes,
+jogs, and trail runs from nearly 30,000 individuals in Salt Lake City, UT,
+@campbell_etal2019 offer what is easily the most empirically robust estimate of
+hiking speed to date. Our application of Campbell's hiking function relies on
+coefficients from the fifth percentile of individuals in their sample, using the
+Lorentz curve as recommended by Campbell, and assumes that everyone in the areas
+of our analysis were largely equal in their walking speed and endurance. This
+makes their walking speeds comparable to, but not the same as, what would be
+estimated by Tobler's hiking function [@tobler1993].
 
-We take the following steps to calculate least-cost travel times in our study regions.
-We first download a digital elevation model for each study area from the USGS National Elevation Dataset using the FedData package in R [@bocinsky2020].
-The original resolution of these gridded data is 30 m, but we aggregate grid cells by a factor of 3, making them closer to 100 m in resolution.
-Aggregation is a familiar speed-up technique for least-cost analysis on large grids containing tens of millions of cells.
-We believe the resulting loss in precision owing to such aggregation is reasonable given that our goal is not so much to find specific paths as it is to determine relations between the locations that these paths connect.
+To calculate travel or commute times, we first download a 1 Arc-second digital
+elevation model (DEM) for each study area from the @usgs2023 3D Elevation
+Program and convert it into a graph with each node representing a grid cell and
+each edge representing a straight line between adjacent grid cells in the Moore
+neighborhood (the eight adjacent grid cells). The degree slope of each edge is
+calculated, and edges with slope estimates greater than or equal to 45 degrees
+are removed. This prevents the path analysis from assuming that a hypothetical
+hiker would, for instance, scale steep slopes or walk off a cliff. The remaining
+slope estimates are then fed to Campbell's hiking function to derive estimates
+of hiking speed along each edge. The inverse of the hiking speed (the pace) is
+then multiplied by the distance along that slope to estimate travel time.
 
-The next step is to convert the aggregated elevation data into a graph in which each node represents a grid cell and each edge represents a straight line between adjacent grid cells in the Moore neighborhood (the eight adjacent grid cells).
-Edge weights are first defined by the slope between adjacent grid cells.
-Those slope estimates are then fed to Campbell’s hiking function to derive estimates of hiking speed along each edge.
-The inverse of the hiking speed (the pace) is then multiplied by the distance between adjacent grid cell centers to estimate travel time.
+Next, we associate each farm and community center with the grid cell that
+contains its centroid. In effect, we treat all the sites falling into a grid
+cell as a single site. We then apply Dijsktra's algorithm to the entire set of
+nodes with associated sites. This gives us two datasets: (i) a dense distance
+matrix with least cost travel-times from all farms and community centers to all
+farms and community centers and (ii) a set of linestring geometries whose
+vertices are the grid cells traversed on shortest paths. For simplicity, we
+average travel times to and from each origin and destination point, thus making
+the distance matrix symmetric. All of the clustering steps in our algorithm are
+implemented as operations on this matrix.
 
-Next, we associate each farm and community center with the grid cell that contains it, and hence to a specific node in the graph.
-In effect, we treat all the sites falling into a grid cell as a single site.
-We then apply a bi-directional form of Dijsktra’s algorithm to the entire set of nodes with associated sites, as this algorithm significantly reduces processing time when the desired outcome involves every combination of start and end points [@sturtevant_felner2018].
-This gives us two datasets: (i) a dense distance matrix with least cost travel-times from all farms and community centers to all farms and community centers and (ii) a set of linestring geometries whose vertices are the grid cells traversed on shortest paths.
+Importantly, the original resolution of the gridded data is \~30 m, but we
+aggregate grid cells by a factor of 3, taking the mean in all cases, and making
+them closer to \~90 m in resolution. Aggregation is a familiar speed-up
+technique for least-cost analysis on large grids containing tens of millions of
+cells, though it is not without its risks [@doyle_etal2012;
+@cushman_landguth2010; @etherington2016], so we also conduct some limited
+sensitivity analysis with varying grid resolutions, which are reported below
+(see also @fig-sensitivity).
 
-For simplicity, we average travel times to and from each origin and destination point, thus making the distance matrix symmetric, which effectively treats the round-trip commute as the relevant cost-distance.
-All of the clustering steps in our algorithm are implemented as operations on this matrix.
-All analyses, including the cost-distance modeling and our implementation of the algorithm, are conducted in the R programming language and environment [@rcoreteam2023].
-For more details, please see Supplementary Materials, which include all code necessary to reproduce these analyses and the figures in the text.
+All analyses, including the cost-distance modeling and our implementation of the
+algorithm, are conducted in the R programming language and environment
+[@rcoreteam2023]. For more details, please see Supplementary Materials, which
+include all code necessary to reproduce these analyses and the figures in the
+text.
 
@@
 
-The goals of this analysis are (i) to identify dispersed farming communities based on the distribution of farmsteads and their proximity to community centers and then (ii) to regionalize those communities by giving them meaningful boundaries.
-This was done by developing a simple algorithm inspired by the concept of a CBSA combined with previous work by @varien1999 and @reese_etal2019.
-Using parameter values D-join = 30 minutes, D-max = 1 hour, and P = 0.8, our algorithm identified 85 communities in CMV, 55 communities in NRG, and 21 communities in CIB (see @fig-communities).
-The mean areas of these communities are 11.56 km$^2$ ($\sigma$ = 7.54, $\eta$ = 10.5) in CMV, 7.20 km$^2$ ($\sigma$ = 5.56, $\eta$ = 5.86) in NRG, and 8.80 km$^2$ ($\sigma$ = 7.66, $\eta$ = 5.93) in CIB; the mean room counts (for all community centers and farms) are 781.62 ($\sigma$ = 991.89, $\eta$ = 420.0) in CMV, 627.73 ($\sigma$ = 594.33, $\eta$ = 443.0) in NRG, and 747.71 ($\sigma$ = 573.92, $\eta$ = 555.0) in CIB; and the mean room densities are 74.92 ($\sigma$ = 67.1, $\eta$ = 52.66) in CMV, 138.58 ($\sigma$ = 150.07, $\eta$ = 92.5) in NRG, and 158.04 ($\sigma$ = 179.00, $\eta$ = 92.32) in CIB.
-We emphasize that the means in several cases deviate to a considerable degree from median values (denoted above with the Greek letter $\eta$), indicating skew in the data and the potential presence of large outliers.
-This is illustrated in the boxplots in @fig-densities, which show the distributions of these and other variables of interest that can be derived from our analysis and also used to evaluate its merits (see also @tbl-results).
+The goals of this analysis are (i) to identify dispersed farming communities
+based on the distribution of farmsteads and their proximity to community centers
+and then (ii) to regionalize those communities by giving them meaningful
+boundaries. This was done by developing a simple algorithm inspired by the
+concept of a CBSA combined with previous work by @varien1999 and
+@reese_etal2019. Using parameter values $D{\text -}join$ = 30 minutes,
+$D{\text -}max$ = 1 hour, and $P$ = 0.8, our algorithm identified 104
+communities in CMV, 90 communities in NRG, and 45 communities in CIB (see
+@fig-communities). The number of communities removed because they included an
+insufficient number of sites was 4 for CMV, 35 for NRG, and 23 for CIB, leaving
+100, 55, and 22 communities with definable spatial extents.
 
-![
+The mean areas of these communities are 8.68 km$^2$ in CMV, 6.06 km$^2$ in NRG,
+and 7.19 km$^2$ in CIB; the mean room counts (for all community centers and
+farms) are 647.76 in CMV, 593.49 in NRG, and 766.68 in CIB; and the mean room
+densities are 99.69 rooms per km$^2$ in CMV, 219.77 rooms per km$^2$ in NRG, and
+238.76 rooms per km$^2$ in CIB. We emphasize that the means in several cases
+deviate to a considerable degree from median values (denoted in @tbl-results
+with the Greek letter $\eta$), indicating skew in the data and the potential
+presence of large outliers. This is illustrated in the boxplots in
+@fig-densities, which show the distributions of these and other variables of
+interest that can be derived from our analysis and also used to evaluate its
+merits (see also @tbl-results).
+
+::: {#fig-communities fig-align="center"}
+![](../figures/communities-all.png){width="6.5in"}
+
 Map showing the size, shape, and spatial distribution of community polygons.
-Color represents the log density of rooms (the number of rooms per square kilometer), with lighter colors indicating lower values and darker colors indicating higher values.
-Note that densities in each region were re-scaled to have the same range, so this figure cannot be used to evaluate absolute differences between regions!
-](../figures/communities-all.png){#fig-communities layout-align="center" width="100%"}
+Color represents the log density of rooms (the number of rooms per square
+kilometer), with lighter colors indicating lower values and darker colors
+indicating higher values. Note that densities in each region were re-scaled to
+have the same range, so this figure cannot be used to evaluate absolute
+differences between regions!
+:::
 
-In addition to showing the community polygons, @fig-communities also offers a simple demographic profile for each region based on the spatial distribution of all rooms (their count per square kilometer) across communities irrespective of time period.
-This provides a simple demonstration of the utility of this method.
-For visualization purposes, the log of the density is used, as it helps to normalize the distribution and, in particular, to rein in large outliers that swamp the variance.
-It is important to note, too, that values in each region are re-scaled to be in the same range, so the figure cannot be used to estimate absolute differences between regions.
-Still, the relative differences within each region can be enlightening, but even they must be interpreted with caution.
+In addition to showing the community polygons, @fig-communities also offers a
+simple demographic profile for each region based on the spatial distribution of
+all rooms (their count per square kilometer) across communities irrespective of
+time period. For visualization purposes, the log of the density is used, as it
+helps to normalize the distribution and, in particular, to rein in large
+outliers that swamp the variance. It is important to note, too, that values in
+each region are re-scaled to be in the same range, so the figure cannot be used
+to estimate absolute differences between regions. Still, the relative
+differences within each region can be enlightening, but even they must be
+interpreted with caution.
 
-Not surprisingly, virtually all of the Mesa Verde cuesta communities show high densities.
-The community that includes Goodman Point and Shield Pueblo in the central area of the CMV region just north of Ute Mountain also has a relatively high density, as do the communities in the far northwest that include Turkey House, Lowry, and Cottonwood Ruin.
-The Dolores area community around Grass Mesa Village in the northeast is also high density.
-In the NRG, the highest density community is a narrow band that lies in a shallow canyon running south from Cundiyo on the east side of the valley.
-The thin east-west running community near San Ildefonso Pueblo is also high density.
-Note that it crosses the Rio Grande, following the contours of the Pojoaque River drainage.
-Just west of there on the Pajarito Plateau, there is a high density community that includes the Sandia center.
-South of Sandia there are two additional high density communities, one near the center of the Pajarito Plateau and another further south along the Chiquito Rio northwest of Cochiti Lake.
-In the CIB area, the three densest communities can be found in the far eastern edge of the study area near El Morro and Tinaja.
-One dense community just to the west of there includes the Ramah School center.
-Another lies in the far northwest near Manuelito.
+Not surprisingly, virtually all of the Mesa Verde NP communities show high
+densities. The community that includes Goodman Point and Sand Canyon Pueblo just
+north of Ute Mountain also has a relatively high density, as do several of the
+communities west of Yellow Jacket and along the Dolores River. In the NRG, the
+highest density community is near the center of the region. It juts out from the
+Rio Grande at its confluence with the Pojoaque River and includes the San
+Ildefonso and Pojoaque puebloes. Another high density community lies in a
+shallow north-south canyon at the base of the Sangre de Cristos. Additional high
+density communities are observed on the Pajarito Plateau in Bandelier NM. In the
+CIB area, the three densest communities can be found in the far eastern edge of
+the study area near El Morro NM, with additional dense communities near Ramah
+just to the west of El Morro and Manuelito in the northwest.
 
 ::: {#tbl-results}
-
 ```{=latex}
@@
 \begin{tabular}{lrllllll}
-    &                                                                   & \textbf{\begin{tabular}[c]{@{}l@{}}Area\\ (km2)\end{tabular}}     & \textbf{\begin{tabular}[c]{@{}l@{}}Rooms\\ (N)\end{tabular}}        & \textbf{\begin{tabular}[c]{@{}l@{}}Farms\\ (N)\end{tabular}}     & \textbf{\begin{tabular}[c]{@{}l@{}}Centers\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Commute\\ (mins)\end{tabular}}  & \textbf{\begin{tabular}[c]{@{}l@{}}Room Density\\ (N/km2)\end{tabular}} \\
+    & 
+    & 
+\textbf{\begin{tabular}[c]{@{}l@{}}Area\\ (km2)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Rooms\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Farms\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Centers\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Commute\\ (mins)\end{tabular}}  & \textbf{\begin{tabular}[c]{@{}l@{}}Room Density\\ (N/km2)\end{tabular}}
+  \\[2mm]
   \hline
   \noalign{\vskip2mm}
-CMV & \begin{tabular}[c]{@{}l@{}}$\eta$\\ $\mu$\\ $\sigma$\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 10.5\\ : 11.56\\ : 7.54\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 420\\ : 781.62\\ : 991.89\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 28\\ : 59.29\\ : 85.78\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 1\\ : 6.01\\ : 13.33\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 21.61\\ : 22.35\\ : 8.46\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 52.66\\ : 74.92\\ : 67.1\end{tabular}      \\[5mm]
+CMV &
+\begin{tabular}[c]{@{}l@{}} $\eta$\\ $\mu$\\ $\sigma$ \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 7.1\\ : 8.68\\ : 6.73 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 402\\ : 647.76\\ : 701.36 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 24.5\\ : 44.38\\ : 55.86 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 1\\ : 1.65\\ : 1.37 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 22.35\\ : 23.28\\ : 8.83 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 57.49\\ : 99.69\\ : 97.36 \end{tabular}
+  \\[5mm]
   \hline
   \noalign{\vskip2mm}
-NRG & \begin{tabular}[c]{@{}l@{}}$\eta$\\ $\mu$\\ $\sigma$\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 5.86\\ : 7.2\\ : 5.56\end{tabular}   & \begin{tabular}[c]{@{}l@{}}: 443\\ : 627.73\\ : 594.33\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 20\\ : 39.53\\ : 37.63\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 2\\ : 2.62\\ : 2.75\end{tabular}  & \begin{tabular}[c]{@{}l@{}}: 16.59\\ : 17.41\\ : 7.21\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 92.5\\ : 138.58\\ : 150.07\end{tabular}    \\[5mm]
+NRG &
+\begin{tabular}[c]{@{}l@{}} $\eta$\\ $\mu$\\ $\sigma$ \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 5.53\\ : 6.06\\ : 4.56 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 479\\ : 593.49\\ : 444.7 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 20\\ : 36.73\\ : 34.96 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 2\\ : 2.49\\ : 2.06 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 19.12\\ : 19.04\\ : 8.77 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 99.43\\ : 219.77\\ : 450.86 \end{tabular}
+  \\[5mm]
   \hline
   \noalign{\vskip2mm}
-CIB & \begin{tabular}[c]{@{}l@{}}$\eta$\\ $\mu$\\ $\sigma$\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 5.93\\ : 8.8\\ : 7.66\end{tabular}   & \begin{tabular}[c]{@{}l@{}}: 555\\ : 747.71\\ : 573.92\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 19\\ : 28.24\\ : 21.96\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 2\\ : 2.29\\ : 1.55\end{tabular}  & \begin{tabular}[c]{@{}l@{}}: 18.52\\ : 19.64\\ : 9.11\end{tabular} & \begin{tabular}[c]{@{}l@{}}: 92.32\\ : 158.04\\ : 179\end{tabular} \\[5mm]
+CIB &
+\begin{tabular}[c]{@{}l@{}} $\eta$\\ $\mu$\\ $\sigma$ \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 4.37\\ : 7.19\\ : 7.39 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 492\\ : 766.68\\ : 626.09 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 13.5\\ : 24.14\\ : 25.34 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 2\\ : 2\\ : 1.11 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 21.16\\ : 24.74\\ : 12.27 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 170.25\\ : 238.76\\ : 270.64 \end{tabular}
+  \\[5mm]
   \hline
@@
 \end{singlespace}
-
 ```
+$\eta$: median, $\mu$: mean, $\sigma$: standard deviation
 
 Community Summaries
+:::
 
-:::
+Results of the sensitivity analysis are as expected given the role that the main
+tuning parameters play in the algorithm (see @fig-sensitivity). Increasing
+$D{\text -}join$ makes it easier to join two communities, so you see growth in
+the total area of communities and declines in the total number of resulting
+communities as it increases. Increasing $P$ makes it harder to join communities,
+so you see fewer and smaller communities as it increases. Increasing
+$D{\text -}max$ increases the total area of communities and the average commute
+time to centers as it serves to filter out distant farmsteads. Importantly,
+finer-grained grid resolutions had no meaningful impact on the results of this
+analysis, so the level of aggregation we chose is acceptable. One puzzling
+result is that the CIB region shows some volatility in its room densities with
+respect to the main tuning parameters. The number of CIB communities is also
+largely invariant. We comment on this in the discussion.
 
@@
 
-Our approach offers several advantages over previous efforts to define dispersed community territories.
-First, it handles variation in survey coverage by seeking to define communities we have evidence for, rather than all communities that may once have existed.
-Second, it handles variation in the degree to which surveyors have lumped or split architectural remains in defining sites [@kintigh2003].
-This is a special problem in the CMV.
-The tradition of site survey recording within the Mesa Verde cuesta has generally involved treating each room block as a separate site, even when there is a continuous artifact scatter between them.
-The tradition in the Montezuma Valley, in contrast, has been to include all room blocks within a single artifact scatter as part of a single site.
-Our method is robust to these differences in site definition, enabling properties of the resulting communities to be compared more directly across these areas.
-Finally, our method focuses on travel time, a strong regularity in human affairs, and can be applied to landscapes with dramatically different topographies.
-The resulting community territories take obvious natural topographic barriers to interaction into account, but also allow territories to cross such boundaries when this is suggested by the data themselves.
+Our approach offers several advantages over previous efforts to define dispersed
+community territories. First, it handles variation in survey coverage by seeking
+to define communities we have evidence for, rather than all communities that may
+once have existed. Second, it handles variation in the degree to which surveyors
+have lumped or split architectural remains in defining sites [@kintigh2003].
+This is a special problem in the CMV. The tradition of site survey recording
+within Mesa Verde National Park has generally involved treating each room block
+as a separate site, even when there is a continuous artifact scatter between
+them. Outside of the park, however, the tradition has been to include all room
+blocks within a single artifact scatter as part of a single site. Because our
+method incorporates room count estimates, it is robust to these differences in
+site definition, enabling properties of the resulting communities to be compared
+more directly across these areas. Finally, our method focuses on travel time, a
+strong regularity in human affairs, and can be applied to landscapes with
+dramatically different topographies. The resulting community territories take
+obvious natural topographic barriers to interaction into account, but also allow
+territories to cross such boundaries when this is suggested by the data
+themselves.
 
-![
-Figure shows the difference between the log density of commute times (in  minutes) from observed farms and random locations to the nearest community center for each region, along with the location of their 95% quantile (this is just the 95% quantile for observed farms, not random points).
-](../figures/commute-kl.png){#fig-commutes layout-align="center" width="60%"}
+::: {#fig-commutes fig-align="center"}
+![](../figures/commute-kl.png){width="4.2in"}
 
-Because our algorithm requires that we calculate the commute time between each farm and community center, we can use it to evaluate potential spatial interactions between them.
-@fig-commutes tries to get at this idea by showing the difference between (i) the log probability density of commute times from observed farms to their nearest centers and (ii) the log probability density of commute times from random locations to their nearest centers, along with the 95% quantiles of the observed distributions.
-This is analogous to what @reese_etal2019 refer to as the "null difference cost distance." One can interpret it as showing where commute times differ from what one might expect were farm sites distributed at random across the landscape, making it similar to the Kullback-Leibler Divergence - basically, what a point process model like MaxEnt tries to estimate [@elith_etal2011; @merow_etal2013].
-Values much larger than zero indicate that there are more farms located at those commute times from centers than you would expect by chance.
-Conversely, values less than zero indicate that there are less farms located at those locations than one might expect by chance.
-In either case, these differences tell us that there is positive spatial covariance between farmsteads and nearby centers at commute times of less than approximately one hour (the value we asisgned to $D-max$).
+Figure shows the difference between the log density of commute times (in
+minutes) from observed farms and random locations to the nearest community
+center for each region, along with the location of their 95% quantile (for
+observed farms, not random points).
+:::
 
-![
-Figure shows for each region the distribution of relevant variables across communities.  
-](../figures/results-boxplot.png){#fig-densities layout-align="center" width="75%"}
+Because our algorithm requires that we calculate the commute time between each
+farm and community center, it also allows us to evaluate potential spatial -
+strictly, commute time - interactions between them. @fig-commutes tries to get
+at this idea by showing the difference between (i) the log probability density
+of commute times from observed farms to their nearest centers and (ii) the log
+probability density of commute times from random locations to their nearest
+centers, along with the 95% quantiles of the observed distributions. This is
+analogous to what @reese_etal2019 refer to as the "null difference cost
+distance." One can interpret it as showing where commute times differ from what
+one might expect were farm sites distributed at random across the landscape.
+Values much larger than zero indicate that there are more farms located at those
+commute times from centers than you would expect by chance. Conversely, values
+less than zero indicate that there are less farms located at those locations
+than one might expect by chance. In either case, these differences tell us that
+there is positive covariance between farmsteads and nearby centers at commute
+times of less than approximately one hour (the value we assigned to
+$D{\text -}max$).
 
-The shape of the responses are not, however, identical.
-We attribute this to three factors: sampling intensity, local topography, and regional settlement patterning.
-For instance, the large spike in the CIB region at or around the 30 minute point along with the longer tail (and the larger 95% quantile) is at least partially owing to the fact that block survey in that area is limited to a few disconnected locations.
-The area is also much flatter than in the CMV and NRG regions, making the background commute distribution much more uniform, at least within the surveyed areas.
-We probably could have accounted for the limited sampling by sampling random background points closer to the observed site locations, but that is not really the point of this exercise.
+The shapes of the density curves are not, however, identical. We attribute this
+to three factors: sampling intensity, local topography, and regional settlement
+patterning. For instance, the spike in the CIB region at or around the 30 minute
+point along with the longer tail (and the larger 95% quantile) is at least
+partially owing to the fact that block survey in that area is limited to a few
+disconnected locations. The area is also much flatter than in the CMV and NRG
+regions, making the background commute distribution much more uniform, at least
+within the surveyed areas.
 
-Perhaps more interesting though is the fact that these regions represent different settlement patterns over different periods of time.
-In the CMV area, the pattern of settlement was largely one of growth and dispersal, with local populations reaching an asymptote after which they spread out into neighboring areas [@schwindt_etal2016].
-Nearly all centers in the CMV area arose within farming clusters, but they were relatively small compared to those in the other study areas, and occurred only to a limited degree in the period leading up to the Great Drought in 1280 CE (and to a lesser degree the period leading up to and including much of the 10th century), with no large aggregations occurring afterwards [@varien1999; @glowacki_ortman2012; @ortman2016a].
-In the NRG region, communities had a few dispersed farms early on, more in the period immediately following 1280 (mostly concentrated in the northern Pajarito Plateau), and then very large aggregations after that [@ortman2016; @ortman2016a].
-Communities in the CIB region had almost no dispersed farming for most of the sequence, mostly skipping over that pattern and moving straight into large aggregations as populations surged in the period after 1280 [@kintigh1985; @schachner2012; @peeples2018], as suggested by the relatively small ratio of farms to centers (see @fig-densities).
-As in the CMV area, aggregations in the NRG and CIB regions sometimes developed within dispersed farming communities, but unlike in the CMV, many also seem to have been established in areas away from where dispersed farms had been previously located [@schachner2015; @ortman2016a].
-This would explain why our algorithm dropped 29 of the 77 centers in the CIB region and 37 of the 181 centers in the NRG but only 1 of the 512 centers in the CMV.
+::: {#fig-densities fig-align="center"}
+![](../figures/results-boxplot.png){width="5.2in"}
 
-It should be emphasized that this is perhaps the biggest limitation of our algorithm, that it simply cannot define community boundaries when the relationship between community centers and dispersed farms is not obvious, either because the center is geographically isolated from the farms or because there is no center at all.
-We note, however, that our concept of a community center does not technically require the presence of an archaeological site.
-That is just our way of identifying the locations of centers in the archaeological record.
+Figure shows for each region the distribution of relevant variables across
+communities.
+:::
 
-In [@reese_etal2019], the area of the community is allowed to expand and contract, with the average area per farm remaining relatively constant on the Mesa Verde cuesta, in the range 0.15 to 0.27 km$^2$ per farm.
-In our algorithm, the total area of the community remains fixed, but the density is allowed to change.
-With the tuning parameter values we chose, the algorithm put about 50% of CMV communities in the range 0.2 to 0.55 km$^2$ per farm, with similar spreads for CIB and NRG.
-That, however, represents a minimum possible level in our analysis and would likely be greater as only a proportion of the farms were actually occupied at a given time.
-Still, this means that the communities we have defined are more likely to meet the 0.14 km$^2$ acre threshold required for a household to subsist on maize [@benson2011; @bocinsky_varien2017].
-But, when we look at the area per room (the inverse of the room density), the picture is a little different.
-For the Mesa Verde region, the median area per room is roughly 0.04 km$^2$, which puts it at the upper end of the range reported in [@reese_etal2019].
-In the NRG and CIB regions, however, the median area per room is closer to 0.01 km$^2$, with a first quartile closer to 0.005 km$^2$.
-Again, these are minimum thresholds assuming every room is occupied at the exact same time, so there was likely more elbow room at any given time.
-Stll, it does suggest that the area for farming in those regions may have been more circumscribed than in the CMV, which is also indicated by the boxplots showing total area in @fig-densities.
-That would imply that levels of intensification [*sensu* @boserup1965] in those regions were likely greater, too, with aggregation and social integration being one way that intensification probably played out.
+Perhaps more interesting though is the fact that these regions represent
+different settlement patterns over different periods of time. In the CMV area,
+the pattern of settlement was largely one of growth and dispersal, with local
+populations reaching an asymptote after which they spread out into neighboring
+areas [@schwindt_etal2016]. Nearly all centers in the CMV area arose within
+farming clusters, but they were relatively small compared to those in the other
+study areas, and occurred only to a limited degree in the period leading up to
+the Great Drought in 1280 CE (and to a lesser degree the period leading up to
+and including much of the 10th century), with no large aggregations occurring
+afterwards [@varien1999; @glowacki_ortman2012; @ortman2016a]. In the NRG region,
+communities had a few dispersed farms early on, more in the period immediately
+following 1280 (mostly concentrated in the northern Pajarito Plateau, the area
+of Bandelier NM), and then very large aggregations after that [@ortman2016;
+@ortman2016a]. As suggested by the relatively small ratio of farms to centers
+(see @fig-densities), communities in the CIB region had almost no dispersed
+farming for most of the sequence, with only an abbreviated phase of dispersed
+settlement that quickly morphed into large aggregations as populations surged in
+the period after 1280 [@kintigh1985; @schachner2012; @peeples2018]. As in the
+CMV area, aggregations in the NRG and CIB regions sometimes developed within
+dispersed farming communities, but unlike in the CMV, many also seem to have
+been established after 1280 CE in areas away from where dispersed farms had been
+previously located [@schachner2015; @ortman2016a].
+
+These settlement patterns explain why our algorithm dropped 23 communities in
+the CIB region, 35 in the NRG, but only 4 in the CMV. Many of the CIB community
+centers are simply too isolated from farmsteads for our algorithm to define a
+meaningful extent. Settlement patterning probably also explains the volatile
+response of the CIB communities to changes in the tuning parameters, as the
+small number of farms means small changes are just enough for communities to add
+or drop community centers that are excluded by our main analysis.
+
+This is perhaps the biggest limitation of our algorithm. It cannot define
+community boundaries when the relationship between community centers and
+dispersed farms is not obvious, either because centers are geographically
+isolated from farms or because there is an insufficient number of farms to serve
+as a bridge between centers. We note, however, that our concept of a community
+center does not technically require the presence of an archaeological site.
+Community center sites are just a clear signal of the locations of persistent
+interaction in the archaeological record.
+
+::: {#fig-sensitivity fig-align="center"}
+![](../figures/sensitivity-analysis.png){width="5.5in"}
+
+Figure shows results of sensitivity analysis, focusing on three main parameters
+($D{\text -}join$, $D{\text -}max$, and $P$) and approximate grid resolution.
+Measures of some key outcomes are also shown, including the number of
+communities, their total area, room density, and average commute time from farms
+to the nearest community center.
+:::
+
+Archaeologists and anthropologists have investigated various dimensions of
+community organization and the types of communities they foster. Examples
+include biological communities organized around genetic relatedness or physical
+similarities [@becker_juengst2017; @kakaliouras2017; @blom2005; @blom2017],
+imagined or ideational communities organized around perceived membership and
+metaphors [sometimes called *styles of imaginings;* @ortman2011; @anderson1983;
+@isbell2000], reproductive communities organized around mating opportunities
+[@mahoney_etal2000; @kolb_snead1997], religious communities organized around a
+shared sense of the sacred [@bernardini2004; @malville_malville2001], linguistic
+communities organized around a shared grammar and vocabulary [@gumperz1968;
+@silverstein1998], even sight communities organized around a shared set of
+visually striking landmarks [@bernardini_peeples2015]. In fact, there are
+probably as many types of communities as there are types of things for people to
+care about.
+
+Because the type of community we aim to delineate is a dispersed *farming*
+community, we structure our algorithm around the dimensions of food production,
+travel time, and social interaction, as these are major constraints on
+settlement for maize farmers [@bocinsky_kohler2014; @bocinsky_etal2016;
+@vernon_etal2022a; @vernon_etalinpress; @yaworsky_etal2023]. To give just one
+example, our algorithm suggests that 50% of CMV communities offer anywhere from
+15-50 hectares per farm. For the CIB and NRG regions, that's 25-45 and 15-35
+hectares per farm, respectively. This means that they meet the 14 hectare
+threshold required for a household to subsist on maize in these regions
+[@reese_etal2019; @benson2011; @bocinsky_varien2017]. And that is before taking
+into account that the area of the community is held fixed [pace @reese_etal2019]
+and all sites are used regardless of time period, so the derived ratios of land
+to farms represent minimum possible levels, not the levels actually experienced
+by farmers in these communities at any given time.
 
@@
 
-The primary goal of this analysis was to derive the boundaries of dispersed maize farming communities from the point locations of archaeological sites.
-To that end, we developed a method that leverages the difference between farmsteads and community centers to cluster sites into communities and define the boundaries of those communities.
-This was inspired by the concept of a CBSA, along with important contributions from @varien1999 and @reese_etal2019.
-An important difference between our approach and that of [@reese_etal2019] is that our community boundaries are static, though crucially this allows us to look at changes in density through time.
-So, the next step in our analysis will be to determine the number of rooms inhabited during each time step.
-A variety of methods are possible for this, but in future studies we intend to utilize uniform probability density analysis of pottery assemblages from each site [@ortman2016] to apportion rooms to time slices.
-This will allow us to capture the changing properties of both dispersed and aggregated communities using consistent measurements.
-The properties that can be compared include:
-
-- Community area (agricultural land in the case of aggregated settlements);
-- Settlement area (village area for aggregated communities, community area for dispersed communities);
-- Community population by time step;
-- Average distance between individuals (square root of community area / 2 for dispersed communities, and square root of village area / 2 for aggregated communities);
-- Environmental conditions (captured by importing community territories into the paleoCAR niche modeling tool in <https://www.openskope.org/>);
-- Topography (the intersection of community territories with a DEM);
-- Properties of civic architecture found within the community, by time step; and
-- Nearest neighbor distances to adjacent community centers.
-
-\noindent The ability to compare these properties for communities of varying scales and densities is crucial for evaluating the relative costs and benefits of these properties for human welfare.
+In this paper, we develop an algorithm that leverages the difference between
+farmsteads and community centers to cluster sites into dispersed farming
+communities and define their boundaries. This was inspired by the concept of a
+CBSA, along with important contributions from @varien1999 and @reese_etal2019.
+Using this algorithm will allow us to capture and compare the changing
+properties of both dispersed and aggregated communities using consistent
+measurements. For instance, we can use uniform probability density analysis of
+pottery assemblages from each site [@ortman2016] to apportion rooms to time
+slices and thus estimate changes in population density through time. The ability
+to compare these properties for communities of varying scales and densities is
+crucial for evaluating their relative costs and benefits for human welfare.
 
@@
 
-Portions of this research were supported by a grant from the National Science Foundation (#2213921, to Ortman).
+Portions of this research were supported by a grant from the National Science
+Foundation (#2213921, to Ortman).
 

### 2024-03-04 manuscript/manuscript.qmd
minor edits

@@
 (CBSA) [@usomb2021]. We start by defining a core area, in this case a community
-center location. We then merge core areas into larger agglomerations based on
-the proportion of their populations that they share. Our method may be described
-as a guided density-based clustering algorithm in that it does not rely on a
-random set of points when initialized. Instead, the selection of core areas is
-guided by archaeological data and regional expertise. We think this is a key
-argument in its favor.
+center location. We then associate outlying areas with their nearest core area
+and merge core areas into larger agglomerations based on the proportion of their
+populations that they share. Our method may be described as a guided
+density-based clustering algorithm in that it does not rely on a random set of
+points when initialized. Instead, the selection of core areas is guided by
+archaeological data and regional expertise. We think this is a key argument in
+its favor.
 
@@
 imagined or ideational communities organized around perceived membership and
-metaphors [sometimes called *styles of imaginings;* @ortman2011; @anderson1983;
+metaphors [sometimes called *styles of imaginings;* @anderson1983; @ortman2011;
 @isbell2000], reproductive communities organized around mating opportunities

### 2024-03-06 R/community-clustering.qmd
minor edits

@@
     x = "Commute to Nearest Center [mins]",
-    y = "Density\nlog(Observed) - log(Random)"
+    y = "Probability Density\nlog(Observed) - log(Random)"
   ) +

### 2024-03-06 manuscript/manuscript.qmd
minor edits

@@
   The transition from dispersed to aggregated forms of settlement reflects a critical shift in the relative value of social and primary (food) modes of production.
-However, investigating trade-offs between them requires estimates of the extent of community territories, including their nearby arable land.
+However, investigating trade-offs between these different forms of settlement requires estimates of the extent of community territories, including their nearby arable land.
 Here we demonstrate a simple algorithm to do that.
-Our algorithm is analogous to that used to define core-based statistical areas for the US census, though instead of modern metropolitan areas, we rely on community centers (or areas of known and persistent interaction between unrelated individuals).
-We provide examples of our algorithm by applying it to archaeological sites in the central Mesa Verde, northern Rio Grande, and Cibola regions.
+Our algorithm is analogous to that used to define core-based statistical areas for the US census, though instead of central business districts, we rely on community centers (or areas of known and persistent interaction between unrelated individuals).
+We provide examples of our algorithm by applying it to archaeological sites in the central Mesa Verde, northern Rio Grande, and Cibola regions in the US Southwest.
 A sensitivity analysis is also conducted to demonstrate how each tuning parameter contributes to the algorithm.
@@
 @gyucha2019]. The primary difference between the two has to do with the spatial
-distribution of residents relative to primary (food) production. In dispersed
-formations, households are scattered across an area, interspersed with the land
-that they farmed, thus requiring them to commute to central places for various
-forms of social production. In aggregated formations, in contrast, households
-are clustered in a village or town, such that they reside in locations of social
-production, but must commute to fields for primary production.
+distribution of residents relative to areas of primary (food) production. In
+dispersed formations, households are scattered across an area, interspersed with
+the land that they farmed, thus requiring them to commute to central places for
+various forms of social production. In aggregated formations, in contrast,
+households are clustered in a village or town, such that they reside in
+locations of social production, but must commute to fields for primary
+production.
 
@@
 over extended periods of time [@glowacki_ortman2012; @varien_etal2007]. In some
-regions these are sites containing more households than could plausibly be
-related by reckoned kinship or some form of civic architecture. The key feature
+regions, community centers are sites containing civic architecture, or more
+households than could plausibly be related by reckoned kinship. The key feature
 of these centers is that they represent locations against which travel time
@@
 that once established, community territories were relatively fixed in space for
-the duration of their associated sites. This allows us to use the distribution
-of all recorded residences, regardless of their periods of occupation, in
-defining community territories. There are several reasons why we believe this is
-a reasonable simplification. First, there are strong cross-cultural regularities
-in the time individuals spend in daily travel from and to their residences, so
-dispersed farming communities tend to be dispersed across the distance
-associated with this typical travel time, regardless of the number of households
-involved [@marchetti1994]. Second, the archaeological record of our study areas
-supports the idea that community territories did not shrink when their
-populations aggregated into villages. In some cases, farmsteads were converted
-to field houses, with some building materials being reused elsewhere and
-evidence for continued limited activity use [@varien1999; @varien1999a;
-@varien2002]. In other cases, grid gardens were constructed directly within
-older rubble mounds by residents of newer, aggregated villages nearby
-[@gauthier_herhahn2005]. Third, an analysis that allowed community territories
-to vary by time step would reduce variation in the density of the resulting
-units, thus washing out variation in one of the most important social properties
-one would ideally want to examine. Fourth, in areas where there is strong survey
-coverage, dispersed residences form clusters, often but not always centered on a
-few larger settlements, and this clustering is apparent even when the dispersed
-residences are plotted by time step [@schachner2012; @varien2002]. Fifth, the
-effects of population growth for dispersed farming communities are not parallel
-to the effects of population expansion in aggregated settlements. In the latter,
-the area of the settlement must grow somewhat because the residential density is
-already relatively high. But in dispersed settlements, there is often unused
-land between existing farmsteads that can be settled. This, combined with the
-strong cross-cultural regularity in typical commute times, leads to much greater
-consistency in the spatial extent of residents in a dispersed community than in
-the case of an aggregated community. Finally, given the inherent incompleteness
-of the archaeological record, it seems most appropriate to use all remains to
-define what is effectively the maximum extent of community territories. While it
-was undoubtedly true that individual community territories expanded or
-contracted over time, the assumption of constant area seems more reasonable than
-the assumption of constant density.
+the duration of the occupations of their associated sites. This allows us to use
+the distribution of all recorded residences, regardless of their periods of
+occupation, in defining community territories. There are several reasons why we
+believe this is a reasonable simplification. First, there are strong
+cross-cultural regularities in the time individuals spend in daily travel from
+and to their residences, so dispersed farming communities tend to be dispersed
+across the distance associated with this typical travel time, regardless of the
+number of households involved [@marchetti1994]. Second, the archaeological
+record of our study areas supports the idea that community territories did not
+shrink when their populations aggregated into villages. In some cases,
+farmsteads were converted to field houses, with some building materials being
+reused elsewhere and evidence for continued limited activity use [@varien1999;
+@varien1999a; @varien2002]. In other cases, grid gardens were constructed
+directly within older rubble mounds by residents of newer, aggregated villages
+nearby [@gauthier_herhahn2005]. Third, an analysis that allowed community
+territories to vary by time step would reduce variation in the density of the
+resulting units, thus washing out variation in one of the most important social
+properties one would ideally want to examine. Fourth, in areas where there is
+strong survey coverage, dispersed residences form clusters, often but not always
+centered on a few larger settlements, and this clustering is apparent even when
+the dispersed residences are plotted by time step [@schachner2012; @varien2002].
+Fifth, the effects of population growth for dispersed farming communities are
+not parallel to the effects of population expansion in aggregated settlements.
+In the latter, the area of the settlement must grow somewhat because the
+residential density is already relatively high. But in dispersed settlements,
+there is often unused land between existing farmsteads that can be settled.
+This, combined with the strong cross-cultural regularity in typical commute
+times, leads to much greater consistency in the spatial extent of residents in a
+dispersed community than in the case of an aggregated community. Finally, given
+the inherent incompleteness of the archaeological record, it seems most
+appropriate to use all remains to define what is effectively the maximum extent
+of community territories. While it was undoubtedly true that individual
+community territories expanded or contracted over time, the assumption of
+constant area seems more reasonable than the assumption of constant density.
 
@@
 the opposite problem, restricting the area of the community to an unreasonable
-degree. Our somewhat brute force way of threading this needle is to identify
+degree. Our somewhat brute force way of balancing these factors is to identify
 paths with shortest commute times between the outer most farms in a community,
@@
 coefficients from the fifth percentile of individuals in their sample, using the
-Lorentz curve as recommended by Campbell, and assumes that everyone in the areas
-of our analysis were largely equal in their walking speed and endurance. This
-makes their walking speeds comparable to, but not the same as, what would be
-estimated by Tobler's hiking function [@tobler1993].
+Cauchy (or Lorentz) distribution as recommended by Campbell, and assumes that
+everyone in the areas of our analysis were largely equal in their walking speed
+and endurance. This makes their walking speeds comparable to, but not the same
+as, what would be estimated by Tobler's hiking function [@tobler1993].
 
@@
 interest that can be derived from our analysis and also used to evaluate its
-merits (see also @tbl-results).
+merits (see also @tbl-results). Note that these summaries are time-insensitive.
+They represent the total archaeological sample in each region.
 
@@
 to estimate absolute differences between regions. Still, the relative
-differences within each region can be enlightening, but even they must be
+differences within each region can be enlightening, though even they must be
 interpreted with caution.
@@
 Rio Grande at its confluence with the Pojoaque River and includes the San
-Ildefonso and Pojoaque puebloes. Another high density community lies in a
-shallow north-south canyon at the base of the Sangre de Cristos. Additional high
-density communities are observed on the Pajarito Plateau in Bandelier NM. In the
-CIB area, the three densest communities can be found in the far eastern edge of
-the study area near El Morro NM, with additional dense communities near Ramah
-just to the west of El Morro and Manuelito in the northwest.
+Ildefonso and Pojoaque pueblos. Another high density community lies in a shallow
+north-south canyon at the base of the Sangre de Cristos. Additional high density
+communities are observed on the Pajarito Plateau in Bandelier NM. In the CIB
+area, the three densest communities can be found in the far eastern edge of the
+study area near El Morro NM, with additional dense communities near Ramah just
+to the west of El Morro and Manuelito in the northwest.
 
@@
 tuning parameters play in the algorithm (see @fig-sensitivity). Increasing
-$D{\text -}join$ makes it easier to join two communities, so you see growth in
-the total area of communities and declines in the total number of resulting
-communities as it increases. Increasing $P$ makes it harder to join communities,
-so you see fewer and smaller communities as it increases. Increasing
-$D{\text -}max$ increases the total area of communities and the average commute
-time to centers as it serves to filter out distant farmsteads. Importantly,
-finer-grained grid resolutions had no meaningful impact on the results of this
-analysis, so the level of aggregation we chose is acceptable. One puzzling
-result is that the CIB region shows some volatility in its room densities with
-respect to the main tuning parameters. The number of CIB communities is also
-largely invariant. We comment on this in the discussion.
+$D{\text -}join$ makes it easier to join two communities, thus leading to growth
+in the total area of communities and declines in the total number of resulting
+communities. Increasing $P$ makes it harder to join communities, thus leading to
+fewer and smaller communities. Increasing $D{\text -}max$ increases the total
+area of communities and the average commute time to centers as it serves to
+filter out distant farmsteads. Importantly, finer-grained grid resolutions had
+no meaningful impact on the results of this analysis, so the level of
+aggregation we chose appears acceptable. One puzzling result is that the CIB
+region shows some volatility in its room densities with respect to the main
+tuning parameters. The number of CIB communities is also largely invariant. We
+comment on this in the discussion.
 
@@
 Because our algorithm requires that we calculate the commute time between each
-farm and community center, it also allows us to evaluate potential spatial -
-strictly, commute time - interactions between them. @fig-commutes tries to get
-at this idea by showing the difference between (i) the log probability density
-of commute times from observed farms to their nearest centers and (ii) the log
-probability density of commute times from random locations to their nearest
-centers, along with the 95% quantiles of the observed distributions. This is
-analogous to what @reese_etal2019 refer to as the "null difference cost
-distance." One can interpret it as showing where commute times differ from what
-one might expect were farm sites distributed at random across the landscape.
-Values much larger than zero indicate that there are more farms located at those
-commute times from centers than you would expect by chance. Conversely, values
-less than zero indicate that there are less farms located at those locations
-than one might expect by chance. In either case, these differences tell us that
-there is positive covariance between farmsteads and nearby centers at commute
-times of less than approximately one hour (the value we assigned to
-$D{\text -}max$).
+farm and community center, we can also evaluate potential spatial - strictly,
+commute time - interactions between them. @fig-commutes illustrates this idea by
+showing the difference between (i) the log probability density of commute times
+from observed farms to their nearest centers and (ii) the log probability
+density of commute times from random locations to their nearest centers, along
+with the 95% quantiles of the observed distributions. This is analogous to what
+@reese_etal2019 refer to as the "null difference cost distance." One can
+interpret it as showing where commute times differ from what one might expect if
+farm sites were distributed randomly across the landscape. Values larger than
+zero indicate that there are more farms located at those commute times from
+centers than you would expect by chance. Conversely, values less than zero
+indicate that there are less farms located at those locations than one might
+expect by chance. In either case, these differences tell us that there is
+positive covariance between farmsteads and nearby centers at commute times of
+less than approximately one hour (the value we assigned to $D{\text -}max$).
 
@@
 Figure shows for each region the distribution of relevant variables across
-communities.
+communities. Note that these are time-insensitive distributions of the samples
+in each region.
 :::
@@
 travel time, and social interaction, as these are major constraints on
-settlement for maize farmers [@bocinsky_kohler2014; @bocinsky_etal2016;
+settlement for subsistence farmers [@bocinsky_kohler2014; @bocinsky_etal2016;
 @vernon_etal2022a; @vernon_etalinpress; @yaworsky_etal2023]. To give just one
 example, our algorithm suggests that 50% of CMV communities offer anywhere from
-15-50 hectares per farm. For the CIB and NRG regions, that's 25-45 and 15-35
-hectares per farm, respectively. This means that they meet the 14 hectare
-threshold required for a household to subsist on maize in these regions
-[@reese_etal2019; @benson2011; @bocinsky_varien2017]. And that is before taking
-into account that the area of the community is held fixed [pace @reese_etal2019]
-and all sites are used regardless of time period, so the derived ratios of land
-to farms represent minimum possible levels, not the levels actually experienced
-by farmers in these communities at any given time.
+15-50 hectares per farm. For the CIB and NRG regions, the corresponding figures
+are 25-45 and 15-35 hectares per farm, respectively. This means that farms in
+all three areas meet the 14 hectare threshold required for a household to
+subsist on maize in these regions [@reese_etal2019; @benson2011;
+@bocinsky_varien2017]. And that is before taking into account that the area of
+the community is held fixed [pace @reese_etal2019] and all sites are used
+regardless of time period, so the derived ratios of land to farms represent
+minimum possible levels, not the levels actually experienced by farmers in these
+communities at any given time.
 
@@
 CBSA, along with important contributions from @varien1999 and @reese_etal2019.
-Using this algorithm will allow us to capture and compare the changing
-properties of both dispersed and aggregated communities using consistent
-measurements. For instance, we can use uniform probability density analysis of
-pottery assemblages from each site [@ortman2016] to apportion rooms to time
-slices and thus estimate changes in population density through time. The ability
-to compare these properties for communities of varying scales and densities is
-crucial for evaluating their relative costs and benefits for human welfare.
+This algorithm allows us to capture and compare the changing properties of both
+dispersed and aggregated communities using consistent measurements. For
+instance, we can use uniform probability density analysis of pottery assemblages
+from each site [@ortman2016] to apportion rooms to time slices and thus estimate
+changes in population density through time. The ability to compare these
+properties for communities of varying scales and densities is crucial for
+evaluating their relative costs and benefits for human welfare.
 

### 2024-03-06 R/community-clustering.qmd
final edits before resubmission

@@
 
+  h <- list(
+    shade = hill,
+    grays = hcl.colors(1000, "Grays")[values(hill)]
+  )
+
   ggplot() +
     geom_spatraster(
-      data = hill, 
-      fill = hcl.colors(1000, "Grays")[values(hill)],
+      data = h[["shade"]], 
+      fill = h[["grays"]],
       maxcell = Inf
@@
     geom_spatraster(data = x[["dem"]], alpha = 0.5) +
+    scale_fill_distiller(palette = "Greys", guide = "none") +
+    ggnewscale::new_scale_fill() +
     scale_fill_hypso_c(
@@
         alpha = 0.5,
-        size = 0.6
+        size = 0.8
       ),
       colour = "white",
-      sigma = 0.5,
+      sigma = 1,
       expand = 1
@@
         color = "#CD3C00",
-        alpha = 0.8,
-        size = 1.1
+        alpha = 0.9,
+        size = 1
       ),
       colour = "white",
-      sigma = 0.5,
+      sigma = 1,
       expand = 1
@@
   filename = fn,
-  width = 7.5, 
+  width = 7, 
   height = 4.5,
@@
 
-  missing_cells <- setdiff(cell_ids, cells_in_graph)
-
-  if (length(missing_cells) > 0){
-
-    cli::cli_abort(c(
-      "Some grid cells have sites but are not in the graph!",
-      "i" = "Cells: {missing_cells}"
-    ))
-
-  }
-
   dm <- get_distance_matrix(
@@
 1.  Identify community centers
-2.  Join centers to their overlapping neighbors
+2.  Join farms to their nearest center
 3.  Remove distant farms
-4.  Join farms to their nearest center
+4.  Join centers to their overlapping neighbors
 5.  Draw smallest concave hull encompassing all farms, centers, and paths
 
-Step 1 is already done and stored in the data. Step 3 is technically done before
-2 since it means those sites can be ignored afterwards.
+Step 1 is already done and stored in the data. Step 2 is technically done after
+step 4. It just makes more sense to think of it coming first, like we're
+building *up* to the community - simply put, small things connect to big things,
+and big things merge into even bigger things.
 
@@
 
-  hell <- function(x, n = 1000){
-
-    slope <- terrain(x, "slope", unit = "radians")
-    aspect <- terrain(x, "aspect", unit = "radians")
+  slope <- terrain(x[["dem"]], "slope", unit = "radians")
+  aspect <- terrain(x[["dem"]], "aspect", unit = "radians")
 
   hill <- shade(slope, aspect, 30, 45)
-    hill <- setValues(hill, scales::rescale(values(hill), to = c(1, n)))
+  hill <- setValues(hill, scales::rescale(values(hill), to = c(1, 1000)))
   hill <- round(hill)
 
-    list(
+  h <- list(
     shade = hill,
-      grays = hcl.colors(n, "Grays")[values(hill)]
-    )
-
-  }
-
-  h <- hell(x[["dem"]])
+    grays = hcl.colors(1000, "Grays")[values(hill)]
+  )
 
@@
   legend.box.margin = margin(t = -10),
-  legend.key.size = unit(0.5, "cm"),
+  legend.key.size = unit(0.43, "cm"),
   legend.margin = margin(),

### 2024-03-06 manuscript/manuscript.qmd
final edits before resubmission

@@
 study area near El Morro NM, with additional dense communities near Ramah just
-to the west of El Morro and Manuelito in the northwest.
+to the west of El Morro, southeast of Zuni near the center of Cibola, and south
+of Manuelito in the northwest.
 

### 2024-03-06 R/labeled-overview-map.R

@@ -297,16 +297,11 @@ ggplot() +
     fill = alpha("white", 0.05),
     linewidth = 0.5
   ) +
-  with_outer_glow(
-    geom_segment(
-      data = landform_labels,
-      aes(x, y, xend = xs, yend = ys),
-      linewidth = 0.27,
-      color = "gray20"
-    ),
-    colour = "white",
-    expand = 4,
-    sigma = 2
+  geom_segment(
+    data = landform_labels,
+    aes(x, y, xend = xs, yend = ys),
+    linewidth = 0.5,
+    color = "white"
   ) +
   with_outer_glow(
     geom_label(
@@ -320,17 +315,11 @@ ggplot() +
     expand = 3,
     sigma = 2
   ) +
-  geom_segment(
-    data = landform_labels,
-    aes(x, y, xend = xs, yend = ys),
-    linewidth = 0.27,
-    color = "gray20"
-  ) +
   geom_point(
     data = landform_labels,
     aes(xs, ys),
     color = "white",
-    size = 0.85
+    size = 1
   ) +
   with_outer_glow(
     geom_text(
@@ -345,7 +334,7 @@ ggplot() +
       nudge_y = -0.01
     ),
     colour = "white",
-    expand = 4,
+    expand = 5,
     sigma = 1
   ) +
   geom_text(

### 2024-06-14 manuscript/jas-reviews-round-two-response.md
final edits

### Response to Reviewer 1

1. Dismissiveness in our response.

    Response: We seem to have communicated poorly in our original response as 
    reviewer one had many important and insightful criticisms of our original 
    submission. In fact, we went to great lengths to incorporate their ideas 
    into the current draft. So, we apologize if there was any misunderstanding
    here.

2. Use the 2022 version of Campbell's hiking function.

    Response: It is indeed an unfortunate oversight on our part that we did not 
    use the most up-to-date version of Campbell's hiking function, especially 
    because one of us (Vernon) worked down the hall from him while he was 
    evidently working on the 2022 paper. At any rate, we thank the reviewer for 
    bringing this to our attention and have incorporated it into the updated 
    analysis.

3. Include a figure to clarify how boundaries are drawn.

    Response: Done! This is now included as an additional figure in the 
    manuscript.

### Response to Reviewer 2

We thank the reviewer for their comments and suggestions for this paper.

### Response to Reviewer 4: Reproducibility Review

1. The Github repository is not accessible.

    Response: The repo was initially set to private, but it has now been made
    public.

2. It would be better to include the uncompiled Quarto document alongside the
   compiled HTML.

    Response: This is now done in two ways: The compiled HTML document now
    has the raw Quarto document embedded in it. This can be accessed by hitting 
    the "</> Code" button at the top of the document and selecting "View 
    Source." The raw content can then be copied and pasted into another a
    text file with a .qmd extension and compiled using Quarto. The HTML document
    also now included a link to the Github repository, which also contains the 
    raw Quarto document.

3. Upload all code and data to another trustworthy repository (meaning, not 
   Github).

    Response: We have uploaded all materials to Zenodo. DOI:

4. Include a README that includes all relevant project metadata.

    Response: This is included in the Github repo and is also uploaded to 
    Zenodo.

5. Include the geopackage "community-centers.gpkg."

    Response: This includes sensitive site location data that we do not have the
    authority to make public.

### 2024-06-14 R/community-clustering.qmd
final edits

@@
     highlight-style: a11y
+    code-tools: true
+    code-links:
+      - text: kbvernon/community-centers
+        icon: github
+        href: https://github.com/kbvernon/community-centers
 execute: 
@@
 
-  # Campbell's hiking function, see Campbell et al 2019
+  # Campbell's hiking function, see Campbell et al 2022
   # note: campbell function wants degrees!
@@
 
-    # values provided in Supplement Table S3
+    # values provided in Appendix Table S3
     # using the 5th percentile of the Lorentz curve as recommended by Campbell
-    a <- -1.527
-    b <- 14.041
-    c <- 36.813
-    d <- 0.320
-    e <- -0.00273
+    a <- -1.4579
+    b <- 22.0787
+    c <- 76.3271
+    d <- 0.0525
+    e <- -0.00032002
 
@@
 
+::: {#fig-boundary-example}
+![](../figures/boundary-example.png){width="100%"
+fig-alt="An example of the method used to define community boundaries."}
+
+Example of the method used to define community boundaries. Left panel shows two
+hull geometries: (i) convex hull with dark gray fill and (ii) concave hull 
+(ratio = 0.1) with light gray fill. Middle panel shows every pairwise short path 
+between the points along the outer edge of the convex hull. Right panel shows 
+community boundary derived by applying the concave hull (ratio = 0.5) to the 
+combined set of site points and vertices of short paths. The most notable
+differences are along the left boundary.
+:::
+
 Our somewhat brute force strategy for navigating between the Scylla and
@@
 
-cmv <- cmv |> define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
-nrg <- nrg |> define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
-cib <- cib |> define_boundaries(concavity_ratio = 0.1, dilate = c(130, -100))
+cmv <- cmv |> define_boundaries(concavity_ratio = 0.5, dilate = c(130, -100))
+nrg <- nrg |> define_boundaries(concavity_ratio = 0.5, dilate = c(130, -100))
+cib <- cib |> define_boundaries(concavity_ratio = 0.5, dilate = c(130, -100))
 

### 2024-06-14 manuscript/manuscript.qmd
final edits

@@
 the distribution of points is suggestive of a concave shape, like the letter
-'C'. A concave hull would handle peculiar shapes, but it would also introduce
+'C'. A concave hull would handle peculiar shapes, but it could also introduce
 the opposite problem, restricting the area of the community to an unreasonable
-degree. Our somewhat brute force way of balancing these factors is to identify
-paths with shortest commute times between the outer most farms in a community,
-extract the vertices from those paths, and then incorporate those into the set
-of points used to define the concave hull. The result is a polygon that is
+degree, depending on how strict the convexity ratio is. Our somewhat brute force
+way of balancing these factors is to identify paths with shortest commute times
+between the outer most farms in a community, extract the vertices from those
+paths, and then incorporate those into the set of points used to define the
+concave hull (see @fig-boundary-example). The result is a polygon that is
 concave as well in the dimension of time, meaning the paths with shortest
 commute times between any two farms never leave the community.
+
+::: {#fig-boundary-example}
+![](../figures/boundary-example.png){width="6.5in"}
+
+Example of the method used to define community boundaries. Left panel shows two
+hull geometries: (i) convex hull with dark gray fill and (ii) concave hull
+(ratio = 0.1) with light gray fill. Middle panel shows every pairwise short path
+between the points along the outer edge of the convex hull. Right panel shows
+community boundary derived by applying the concave hull (ratio = 0.5) to the
+combined set of site points and vertices of short paths. The most notable
+differences are along the left boundary.
+:::
 
@@
 hiking speed across a range of slopes have advanced considerably in recent
-years, most notably in @campbell_etal2019. Drawing on a large, crowd-sourced
-fitness tracker data set consisting of approximately 420,000 individual hikes,
-jogs, and trail runs from nearly 30,000 individuals in Salt Lake City, UT,
-@campbell_etal2019 offer what is easily the most empirically robust estimate of
-hiking speed to date. Our application of Campbell's hiking function relies on
-coefficients from the fifth percentile of individuals in their sample, using the
-Cauchy (or Lorentz) distribution as recommended by Campbell, and assumes that
-everyone in the areas of our analysis were largely equal in their walking speed
-and endurance. This makes their walking speeds comparable to, but not the same
-as, what would be estimated by Tobler's hiking function [@tobler1993].
+years, most notably in [@campbell_etal2022; see also @campbell_etal2019]. Our
+application of Campbell's hiking function relies on coefficients from the median
+range of individuals in their sample and assumes that everyone in the areas of
+our analysis was largely equal in their walking speed and endurance. This makes
+their walking speeds comparable to, but not the same as, what would be estimated
+by Tobler's hiking function [@tobler1993].
 
@@
 @reese_etal2019. Using parameter values $D{\text -}join$ = 30 minutes,
-$D{\text -}max$ = 1 hour, and $P$ = 0.8, our algorithm identified 104
-communities in CMV, 90 communities in NRG, and 45 communities in CIB (see
+$D{\text -}max$ = 1 hour, and $P$ = 0.8, our algorithm identified 103
+communities in CMV, 88 communities in NRG, and 45 communities in CIB (see
 @fig-communities). The number of communities removed because they included an
-insufficient number of sites was 4 for CMV, 35 for NRG, and 23 for CIB, leaving
-100, 55, and 22 communities with definable spatial extents.
+insufficient number of sites was 5 for CMV, 33 for NRG, and 23 for CIB, leaving
+98, 55, and 22 communities with definable spatial extents.
 
-The mean areas of these communities are 8.68 km$^2$ in CMV, 6.06 km$^2$ in NRG,
-and 7.19 km$^2$ in CIB; the mean room counts (for all community centers and
-farms) are 647.76 in CMV, 593.49 in NRG, and 766.68 in CIB; and the mean room
-densities are 99.69 rooms per km$^2$ in CMV, 219.77 rooms per km$^2$ in NRG, and
-238.76 rooms per km$^2$ in CIB. We emphasize that the means in several cases
+The mean areas of these communities are 9.51 km$^2$ in CMV, 6.5 km$^2$ in NRG,
+and 7.74 km$^2$ in CIB; the mean room counts (for all community centers and
+farms) are 663 in CMV, 598.65 in NRG, and 768.05 in CIB; and the mean room
+densities are 83.83 rooms per km$^2$ in CMV, 183.93 rooms per km$^2$ in NRG, and
+207.18 rooms per km$^2$ in CIB. We emphasize that the means in several cases
 deviate to a considerable degree from median values (denoted in @tbl-results
@@
     & 
-\textbf{\begin{tabular}[c]{@{}l@{}}Area\\ (km2)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Rooms\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Farms\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Centers\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Commute\\ (mins)\end{tabular}}  & \textbf{\begin{tabular}[c]{@{}l@{}}Room Density\\ (N/km2)\end{tabular}}
+\textbf{\begin{tabular}[c]{@{}l@{}}Area\\ (sq.km)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Rooms\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Farms\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Centers\\ (N)\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Commute\\ (mins)\end{tabular}}  & \textbf{\begin{tabular}[c]{@{}l@{}}Room Density\\ (N/sq.km)\end{tabular}}
   \\[2mm]
@@
 \begin{tabular}[c]{@{}l@{}} $\eta$\\ $\mu$\\ $\sigma$ \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 7.1\\ : 8.68\\ : 6.73 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 402\\ : 647.76\\ : 701.36 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 24.5\\ : 44.38\\ : 55.86 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 1\\ : 1.65\\ : 1.37 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 22.35\\ : 23.28\\ : 8.83 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 57.49\\ : 99.69\\ : 97.36 \end{tabular}
+\begin{tabular}[c]{@{}l@{}} : 7.61\\ : 9.51\\ : 7.2 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 399\\ : 663\\ : 753.62 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 25.5\\ : 45.61\\ : 59.11 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 1\\ : 1.67\\ : 1.39 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 21.51\\ : 22.72\\ : 8.3 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 55.08\\ : 83.83\\ : 71.93 \end{tabular}
   \\[5mm]
@@
 \begin{tabular}[c]{@{}l@{}} $\eta$\\ $\mu$\\ $\sigma$ \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 5.53\\ : 6.06\\ : 4.56 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 479\\ : 593.49\\ : 444.7 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 20\\ : 36.73\\ : 34.96 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 2\\ : 2.49\\ : 2.06 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 19.12\\ : 19.04\\ : 8.77 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 99.43\\ : 219.77\\ : 450.86 \end{tabular}
+\begin{tabular}[c]{@{}l@{}} : 5.45\\ : 6.5\\ : 5.07 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 475\\ : 598.65\\ : 468.12 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 18\\ : 36.91\\ : 37.55 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 2\\ : 2.53\\ : 2.17 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 18.14\\ : 18.28\\ : 8.65 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 98.69\\ : 183.93\\ : 269.84 \end{tabular}
   \\[5mm]
@@
 \begin{tabular}[c]{@{}l@{}} $\eta$\\ $\mu$\\ $\sigma$ \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 4.37\\ : 7.19\\ : 7.39 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 492\\ : 766.68\\ : 626.09 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 13.5\\ : 24.14\\ : 25.34 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 4.59\\ : 7.74\\ : 7.96 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 492\\ : 768.05\\ : 626.91 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 13.5\\ : 24.23\\ : 25.39 \end{tabular} &
 \begin{tabular}[c]{@{}l@{}} : 2\\ : 2\\ : 1.11 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 21.16\\ : 24.74\\ : 12.27 \end{tabular} &
-\begin{tabular}[c]{@{}l@{}} : 170.25\\ : 238.76\\ : 270.64 \end{tabular}
+\begin{tabular}[c]{@{}l@{}} : 20.85\\ : 24.36\\ : 12.14 \end{tabular} &
+\begin{tabular}[c]{@{}l@{}} : 157.01\\ : 207.18\\ : 218.53 \end{tabular}
   \\[5mm]
@@
 communities. Increasing $P$ makes it harder to join communities, thus leading to
-fewer and smaller communities. Increasing $D{\text -}max$ increases the total
-area of communities and the average commute time to centers as it serves to
-filter out distant farmsteads. Importantly, finer-grained grid resolutions had
-no meaningful impact on the results of this analysis, so the level of
-aggregation we chose appears acceptable. One puzzling result is that the CIB
-region shows some volatility in its room densities with respect to the main
-tuning parameters. The number of CIB communities is also largely invariant. We
-comment on this in the discussion.
+more smaller communities. Increasing $D{\text -}max$ increases the total area of
+communities and the average commute time to centers as it serves to filter out
+distant farmsteads. Importantly, finer-grained grid resolutions had no
+meaningful impact on the results of this analysis, so the level of aggregation
+we chose appears acceptable. One puzzling result is that the CIB region shows
+some volatility in its room densities with respect to the main tuning
+parameters. The number of CIB communities is also largely invariant. We comment
+on this in the discussion.
 
@@
 minutes) from observed farms and random locations to the nearest community
-center for each region, along with the location of their 95% quantile (for
-observed farms, not random points).
+center for each region, along with the location of their 95% quantile for
+observed farms (the location of the point next to the respective region label,
+all being between 60 and 90 minutes).
 :::

### 2024-06-14 R/boundary-example.R

@@ -1,0 +1,132 @@
+
+library(here)
+library(patchwork)
+library(sf)
+library(sfdep)
+library(smoothr)
+library(tidyverse)
+
+# plot defaults -----------------------------------------------------------
+
+theme_set(theme_bw())
+
+theme_update(
+  axis.text = element_blank(),
+  axis.ticks = element_blank(),
+  axis.title = element_blank(),
+  panel.grid = element_blank(),
+  plot.margin = margin(1,1,1,1),
+  plot.title = element_text(size = rel(0.9), margin = margin(b = 1))
+)
+
+# data --------------------------------------------------------------------
+
+pts <- here("data", "example-boundary-data.geojson") |> 
+  read_sf() |> 
+  slice(1)
+
+short_paths <- here("data", "example-boundary-data.geojson") |> 
+  read_sf() |> 
+  slice(-1)
+
+# plots -------------------------------------------------------------------
+
+hulls <- ggplot() +
+  geom_sf(
+    data = pts |> st_convex_hull(),
+    fill = "#8b8b8b",
+    color = "#3e3e3e"
+  ) +
+  geom_sf(
+    data = pts |> st_concave_hull(0.1),
+    fill = "#f6f6f6",
+    color = "#3e3e3e"
+  ) +
+  geom_sf(
+    data = pts,
+    size = 1.6,
+    color = "white"
+  ) +
+  geom_sf(
+    data = pts,
+    shape = 21,
+    size = 1,
+    fill = "#6f6f6f",
+    color = "#222222"
+  ) +
+  ggtitle("Hull Geometries")
+
+paths <- ggplot() +
+  geom_sf(
+    data = short_paths,
+    color = "#6f6f6f",
+    linewidth = 0.3
+  ) +
+  geom_sf(
+    data = pts,
+    size = 1.6,
+    color = "white"
+  ) +
+  geom_sf(
+    data = pts,
+    shape = 21,
+    size = 1,
+    fill = "#6f6f6f",
+    color = "#222222"
+  ) +
+  ggtitle("Short Paths")
+
+comm <- ggplot() +
+  geom_sf(
+    data = short_paths |> 
+      st_cast("POINT") |> 
+      bind_rows(pts) |> 
+      st_combine() |> 
+      st_concave_hull(0.5),
+    fill = "#e2e2e2",
+    color = "#3e3e3e"
+  ) +
+  geom_sf(
+    data = pts,
+    size = 1.6,
+    color = "white"
+  ) +
+  geom_sf(
[…]

### 2024-06-14 manuscript/manuscript.qmd
include zenodo doi and citation

@@
 algorithm, are conducted in the R programming language and environment
-[@rcoreteam2023]. For more details, please see Supplementary Materials, which
-include all code necessary to reproduce these analyses and the figures in the
-text.
+[@rcoreteam2023]. For more details, please see the Zenodo repository
+[@kenneth_blake_vernon_2024_11661270], which includes all code necessary to
+reproduce all analyses and figures in the text.
 

### 2024-06-14 R/boundary-example.R
magick, by the cars

@@ -1,5 +1,6 @@
 
 library(here)
+library(magick)
 library(patchwork)
 library(sf)
 library(sfdep)
